---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---

# Does a Pandemic Lead to a Change in Behavior that Affects Unnatural Deaths?\--Team Rabbit

## Introduction

------------------------------------------------------------------------

As observed globally, pandemics alter the lives of people in and around
affected areas in many ways. Studying these changes in behavior of the
people affected by a pandemic and varying pandemic mitigation measures
is fundamental to understanding secondary effects of these changes. The
objective of this analysis is to assess the statistical significance of
unnatural causes of death during a pandemic time period verses during a
non-pandemic time period, and afterwards to raise the question as to
whether possible changes in behavior might have had a positive or
negative effect on total unnatural deaths during a pandemic. Unnatural
deaths are defined as all deaths relating to intentional and
unintentional injury. They are the fourth leading cause of death in the
United States over the time period analyzed. We have sampled 13 months
for our non-pandemic data (January 1, 2019 to January 31, 2020); and we
have sampled 10 months for our pandemic data (March 1, 2020 to December
31, 2020). We exclude February 2020 to avoid ambiguity of the definition
of "pandemic".

In the following sections we break down our statistical analysis into
three hypothesis tests. First, we test a difference in two means over
neighboring time periods to assess whether there is a statistical
significance in raw number differences in the average deaths in each
respective time period. Second, we test a difference in two proportions
of unnatural deaths to total deaths in each time period respectively.
The proportion test will show whether unnatural deaths are expected to
scale proportionally to total deaths during a pandemic as they would in
a non-pandemic time period. Third, we run a permutation test on the
difference in means in order to gather a third and final perspective.
The results will show fundamental generalities to an otherwise complex
topic, and we will discuss the results at the end of this analysis.

## Preliminary Work and General Data

------------------------------------------------------------------------

```{r}
# Original Data Extract Lives Online
# Data from CDC: https://wonder.cdc.gov/ucd-icd10.html

mortality <- read.delim("https://raw.githubusercontent.com/CU-Denver-MathStats-OER/Statistical-Theory/main/Data/UnderlyingCauseofDeath1999-2020.txt")

# There are 41 NA's in Deaths
# Removing observations with any missing data
mortality <- na.omit(mortality)

# There are rows of aggregated data (with no month) we remove
mortality <- mortality[mortality$Month.Code > 0, ] 

n <- nrow(mortality)  # number of rows in dataframe

# Create new variables natural and unnatural deaths
Natural  <- numeric(n)
Unnatural  <- numeric(n)

for (i in 1:n) {
  if (mortality$ICD.Chapter.Code[i] == "V01-Y89"){
    Unnatural[i] = mortality$Deaths[i]
  } else {
    Natural[i] = mortality$Deaths[i]
  }
}

mortality$Natural <- Natural
mortality$Unnatural <- Unnatural

mortality <- mortality[c("Month.Code", "Natural","Unnatural")]

# Convert strings to categorical variables using factors
mortality$Month.Code <- factor(mortality$Month.Code)

# Aggregate by month
mortality2 <- aggregate(cbind(Natural, Unnatural) ~ Month.Code, data = mortality, FUN = sum)

# Create a Total column
mortality2$Total <- mortality2$Natural + mortality2$Unnatural
tail(mortality2)

##########################
##########################
##########################

library(stringr)

mortality2[c("year", "month")] <- str_split_fixed(mortality2$Month.Code, '/', 2)

#subset unnatural deaths by non pandemic and pandemic times
mortality2$month <- strtoi(mortality2$month, 10L)

# Unnatural Deaths -- Unnatural deaths not specified, but natural deaths are

NonPandemic <- subset(mortality2, select="Unnatural", ((year == "2019" & month >= "1") & (year == "2019" & month <= 12)) | (year == "2020" & month == 1),  drop=T)

Pandemic <- subset(mortality2, select="Unnatural", ((year == "2020" & month >= 3) & (year == "2020" & month <= 12)), drop=T)

# Natural Deaths

NonPandemicNatural <- subset(mortality2, select="Natural", ((year == "2019" & month >= "1") & (year == "2019" & month <= 12)) | (year == "2020" & month == 1),  drop=T)

PandemicNatural <- subset(mortality2, select="Natural", ((year == "2020" & month >= 3) & (year == "2020" & month <= 12)), drop=T)
```

## Summary Statistics

------------------------------------------------------------------------

### Numerical Summaries

```{r}
summary(mortality2)
```

### Barplot Comparing Proportions of Unnatural Deaths 2019 to 2020

```{r}
# Find proportion of death 2019:
p.nonpandemic <- sum(NonPandemic) / (sum(NonPandemicNatural) + sum(NonPandemic))

# Find proportion of death 2020:
p.pandemic <- sum(Pandemic) / (sum(PandemicNatural) + sum(Pandemic))

# Bind together for bar chart
bind.proportion <- cbind(p.nonpandemic, p.pandemic)

# Plot Proportion Differences
prop.barplot <- barplot(bind.proportion,
        beside = TRUE,
        main = "Proportions of Unnatural Deaths Pandemic vs Non-pandemic",
        col = "Red",
        xlab = "Sample Time Period",
        ylab = "(Total Unnatural Deaths)/(Total Deaths)",
        ylim = c(0, .12))
text(prop.barplot, 0, round(bind.proportion, 5), cex = 1, pos = 3)
```

### Box Plot Comparing Unnatural Deaths 2019 to 2020

```{r}
#boxplot(mortality2$Natural)

#Bind together for Box plot
bind.unnatural <- cbind(NonPandemic, Pandemic)

boxplot(bind.unnatural, 
        beside = TRUE,
        main = "Unnatural Deaths Pandemic vs Non-pandemic",
        col = "pink",
        xlab = "Sample", 
        ylab = "Death Count", 
        ylim = c(20000, 28000))
```

### Histogram of Unnatural Death of All Time

```{r}
hist(mortality2$Unnatural, 
     xlab = "Unnatural Death Counts Per Month",
     main = "Histogram of Unnatural Death Counts of All Time")
```

## Hypothesis Tests

------------------------------------------------------------------------

### Difference in Two Means

We want to know if the mean total unnatural deaths during a pandemic is
different from the mean total unnatural deaths not during a pandemic per
month in the US.

We choose the $5\%$ significance level and our hypotheses are below:

$H_0 : \mu_{p} = \mu_{n}$ : The mean number of unnatural deaths per
month during a pandemic is the same as as the mean number of deaths per
month in a non-pandemic year

$H_a: \mu_{p} \neq \mu_{n}$ : The mean number of unnatural deaths per
month during a pandemic is different from the mean number of deaths per
month in a non-pandemic year.

Since we do not know the population standard deviation and our sample
size is small, we are using the Student's t-distribution to test our
hypothesis. Here we show the results for both "by hand" and using the R
command `t.test()`.

By "by hand", it means we select two sub-samples from the original
sample: `Pandemic` and `NonPandemic` and calculate their sample means
separately. Then, we use the formula of $t$-score for independent sample
means:

$${\large {t = \frac{\bar{X}-\bar{Y}}{\sqrt{\frac{S_1^2}{n_1}+ \frac{S_2^2}{n_2}}}}}.$$
We use $df = n_{\rm min} -1$ as the degrees of freedom when calculating
the $p$-value with R command `qt()`, where $n_{\rm min}$ is the smaller
of the two sample sizes.

```{r}
# Calculate t-score and p-value "by hand"
mean.pandemic <- mean(Pandemic)
se.pandemic <- sd(Pandemic)

mean.nonpandemic <- mean(NonPandemic)
se.nonpandemic <- sd(NonPandemic)

test.stat <- mean.pandemic - mean.nonpandemic
test.stat

t.score <- test.stat/sqrt(se.pandemic^2/length(Pandemic)+se.nonpandemic^2/length(NonPandemic))
t.score

pt(t.score, min(length(Pandemic), length(NonPandemic)) - 1, lower.tail = FALSE)*2
```

We get a test statistic of 3157.215 more unnatural deaths per month
during a pandemic than not in a pandemic. The $t$-score we get is 5.845
and the $p$-value we get is approximately $.00025 \le \alpha=0.05$.

We also use the command
`t.test(Variable ~ HowGroupsSplit, data = Dataset_name, alt = "two.sided")`
to test our hypotheses on the $t$-distribution with degrees of freedom
computed from Welch's approximation. Here we use the pooled data with a
binary (categorical) variable called `pandemic`. When the observation is
from before February 2020, we assign it with `pandemic = 1` and 0
otherwise.

However, this command calculates the test statistic as the difference
between non-pandemic and pandemic, which shows negative sign. This does
not affect the $p$-value. This $p$-value is different from the "by hand"
$p$-value because we were using the smaller number of observations
between the two samples minus one to approximate Welch's degrees of
freedom.

```{r}
# Using pooled data (excluding Feb 2020)
mortality2_nofeb20<-subset(mortality2, Month.Code != "2020/02")

# generate a dummy variable called pandemic
mortality2_nofeb20$pandemic <- ifelse(mortality2_nofeb20$year=="2019" 
                                      | mortality2_nofeb20$Month.Code=="2020/01", 0, 1)

t.test(Unnatural ~ pandemic, data = mortality2_nofeb20, alt = "two.sided")
```

Alternatively, we graph the $t$-distribution below with the $95\%$
confidence interval marked with blue solid lines. The orange curve
represents the probability density function (PDF) for the
$t$-distribution. Our $t$-score is marked with a red dash line, which is
well within the rejection region.

```{r}
# Store the Welch's degrees of freedom
DF <- t.test(Unnatural ~ pandemic, data = mortality2_nofeb20, alt = "two.sided")$parameter

# Take a random sample from the t-distribution of size 10000 and draw a histogram
rsample <- rt(10000, df=DF)
hist(rsample, prob=TRUE, 
     xlim=c(-8,8), breaks = seq(-8, 8, 0.3), 
     xlab = "t-score with df = 13.913",
     main = "Student's t-distribution (PDF)")
lines(density(rsample), col="orange", lwd=2)

# Add a red line at the t-score
abline(v = t.score, col = "red", lwd = 2, lty = 2)

# Add blue lines at the 95% confidence level for the t-distribution
lower.t <-  qt(0.025, df=DF)
upper.t <-  qt(0.975, df=DF)

abline(v = lower.t, col = "blue", lwd = 2, lty = 1)
abline(v = upper.t, col = "blue", lwd = 2, lty = 1)
```

#### Results

The $p$-value from the $t$-score we hand-calculated is approximately
$.00025 < \alpha=0.05$. When using Welch's approximated degrees of
freedom, our $p$-value is $4.36\times10^{-5} < \alpha = 0.05$.
Therefore, we reject the null and have evidence to support the
alternative hypothesis that the mean unnatural deaths per month during
the pandemic is different than not during a pandemic in the US at the
$5\%$ significance level.

------------------------------------------------------------------------

### Difference in Proportions

For this section, we want to see if there is a difference in proportion
between unnatural deaths that occur during a pandemic versus not during
a pandemic.

The null hypothesis will be that there is no difference between the
proportion of deaths during a pandemic versus non-pandemic times, while
the alternative hypothesis is that there is a difference. The
mathematical expressions of the hypotheses are below:

$H_0 : p_{p} = p_{n}$

$H_a : p_{p}\neq p_{n}$

In order to find the $p$-value, two different methods are used. The
first method is the "easy" method, found by using
`prop.test(c(x1, x2), c(n1, n2), alt="two.sided")`. In this case, `x1`
and `x2` represent the number of unnatural deaths in pandemic and
non-pandemic time, while `n1` and `n2` represent the total number of
deaths for pandemic and non-pandemic time.

```{r}
#Totaling the number of unnatural deaths
n.nonpandemic <- sum(mortality2$Unnatural[1:13])
n.pandemic <- sum(mortality2$Unnatural[15:24])

#Totaling the number of deaths in nonpandemic and pandemic years
total.nonpandemic <- sum(mortality2$Total[1:13])
total.pandemic <- sum(mortality2$Total[15:24])

#Finding the Necessary Proportions for the Foruma
prop.pandemic <- n.pandemic/total.pandemic
prop.nonpandemic <- n.nonpandemic/total.pandemic

#Pooled Proportion
pooled.prop <- (n.pandemic + n.nonpandemic)/(total.pandemic+total.nonpandemic)

#Finding the P Value using a "more simple approach"
prop.test(c(n.pandemic,n.nonpandemic),c(total.pandemic,total.nonpandemic), alt="two.sided")
```

The second method makes use of finding a test statistic, and using that
test statistic in a `pnorm()` function. The test statistic is found by
using the formula,
$${\large {z= \frac{\hat{p}_1-\hat{p}_2}{\sqrt{\hat{p}_p(1-\hat{p}_p) \left( \frac{1}{n_1}+\frac{1}{n_2} \right) }}}}.$$

This $z$-score is then used in a `pnorm()` function to get the
$p$-value, which is almost 0 (because of rounding).

```{r}
#finding the Test Statistic
z <- (prop.pandemic-prop.nonpandemic)/sqrt(pooled.prop*(1-pooled.prop)*(1/total.pandemic+1/total.nonpandemic))
z

#Finding the P value "by hand"
pnorm(z)*2
```

#### Results

In both tests, we end up with a p-value extremely small (almost 0).
Using the $5\%$ significance level, we can reject the null hypothesis,
and there is evidence that supports the alternative hypothesis. If we
were to reduce the significance interval to $1\%$, the results still
point to being able to reject the null hypothesis.

------------------------------------------------------------------------

### Permutation Test for Difference in Two Means

We sustain the hypotheses in the classical hypothesis test for
[difference in two means]. Here we do not repeat them in redundance.

```{r}
#permutation test for difference in two means
all_unnatural <- mortality2_nofeb20$Unnatural
total_unnatural <- length(all_unnatural)
length.pandemic <- length(Pandemic)

N <- 10^5 - 1
result_mean <- numeric(N)

for(i in 1:N){
  #get random indices to get random values of unnatural deaths from original samples to use for resample
  index <- sample(total_unnatural, size=length.pandemic, replace = FALSE)
  #calculate diff in means for this resample
  result_mean[i] <- mean(all_unnatural[index]) - mean(all_unnatural[-index])
}

hist(result_mean, xlab = "Difference in sample mean",
     main = "Permutation Distribution for Difference in Sample mean", breaks = 20)
abline(v = c(test.stat), col = c("blue"))
```

```{r}
#get p-value for diff in two means (we are doing two tailed test)
p_valmean <- (sum(abs(result_mean) >= abs(test.stat))+1)/(N+1)
p_valmean
```

#### Results

a)  The test statistic is taking the difference in the mean of unnatural
    deaths during a pandemic and mean of unnatural deaths not during a
    pandemic in the US. The observed test statistic is 3157.215.

b)  The underlying null distribution used to compute the $p$-value was a
    permutation distribution.The $p$-value calculated from this test was
    0.00002.

c)  When using a significance level of $\alpha = 0.05$, we can see that
    $2\times10^{-5} < 0.05$. We can safely reject the null hypothesis
    and have evidence to support the alternative hypothesis that there
    is a difference in the mean of unnatural deaths during a pandemic
    and not during a pandemic in the US.

## Conclusions

------------------------------------------------------------------------

In our [Hypothesis Tests] section, we performed both parametric tests
and a permutation test. The $p$-values we got were extremely small
$(10^{-5})$, indicating a very significant test statistic. We rejected
all of our null hypotheses at the $5\%$ significance level and concluded
that---

1.  The mean number of unnatural deaths per month during a pandemic
    during a pandemic is different from not during a pandemic in the US;

2.  The ratio of unnatural deaths to total deaths in the United States
    during a pandemic is different from not during a pandemic in the US.

According to the results of our hypothesis tests, we think the pandemic
is indeed associated with a change of behavior among people that may
have caused significant difference in the number and ratio of unnatural
deaths per month. We suggest future studies to look into more specific
mechanisms potentially causing the behavioral change and its other
health and economic consequences. Alternatively, policy scholars might
inquire into this phenomenon and recommend policy changes to mitigate
part of the losses from the pandemic.

## Updated List for Division of Work

------------------------------------------------------------------------

Data gathering and cleaning: Garry

Hypothesis Test for difference in two means: Yuxin

Permutation Test for difference in two means: Ivy

Hypothesis Test for difference in two proportions: Kaysi

Make relevant predictions and interpret results: Yuxin

Design Presentation Slides: Everyone

Report: Everyone
