---
title: 'MATH 3382: Homework 2 - Due Sep. 8 at 2PM'
author: "Ivy Truong"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Instructions:

------------------------------------------------------------------------

**Submit your work electronically as an R Markdown file extension `.Rmd`.**

Upload your file into the Homework 2 assignment in Canvas before 2 PM on Thurs, September 8.

**You must provide supporting work for your all of your answers.**

-   [That work might be inserting some R code cells.]{style="color: blue;"}

    -   **Feel free to insert as much R code as you like.**
    -   You must interpret or explain how the output of a code cell is being utilized to solve the problem.
    -   **Code with incomplete or missing explanations of why the output is useful may not receive full credit.**

-   [It might involve doing some algebra and/or calculus (integrating, differentiating, etc).]{style="color: blue;"}

    -   **Be sure to type/write out the steps of your work so I can see how you arrived at your answer.**
    -   **If typing takes too long no problem.** You can write out the support work on paper and hand that in before class.

-   [Probably it a little bit of both on most problems, but you decide how you want to solve the problems!]{style="color: blue;"}

## Important Academic Policies:

------------------------------------------------------------------------

-   **Be your own work.** Though you may collaborate with others, everyone is responsible for writing up the work in their own way using their own methods. Plagiarism of any form is not tolerated.
-   **Be complete.** You must provide all work and/or explanations needed to find the solution. Answers with insufficient or incomplete supporting work may lose credit.
-   **Adhere to the Code of Academic Honesty.**
-   **Be clearly written (and legible if written).** Your solution to a problem must be clear, written in complete sentences. You may lose credit for work that is unclear or hard to follow.

### [Let me know if you need help, and **GOOD LUCK!!!**]{style="color: blue;"}

------------------------------------------------------------------------

# Question 1:

------------------------------------------------------------------------

Home mortgages can be classified as fixed rate $F$ or variable rate $V$. A sample of four home mortgages are selected and the type of mortgage of each is recorded, for example $FFVV$.

## Question 1a:

------------------------------------------------------------------------

What are the 16 outcomes in the sample space $\Omega$ of this experiment?

### Solution to Question 1a:

------------------------------------------------------------------------

Enter your answers in the table below using a string of 4 letters such as FFFF for one example.

| Number | Outcome | Number | Outcome |
|--------|---------|--------|---------|
| 1\.    | $FFFF$  | 9\.    | $VVFF$  |
| 2\.    | $FFFV$  | 10\.   | $FVVV$  |
| 3\.    | $FFVF$  | 11\.   | $VFVV$  |
| 4\.    | $FVFF$  | 12\.   | $VVFV$  |
| 5\.    | $VFFF$  | 13\.   | $VVVF$  |
| 6\.    | $FFVV$  | 14\.   | $VVVV$  |
| 7\.    | $FVFV$  | 15\.   | $VFVF$  |
| 8\.    | $FVVF$  | 16\.   | $VFFV$  |

\
\

## Question 1b:

------------------------------------------------------------------------

List all the outcomes that are in the event "exactly three of the selected mortgages are fixed rate"?

### Solution to Question 1b:

------------------------------------------------------------------------

The outcomes that are in the event "exactly three of the selected mortgages are fixed rate" are: FFFV, FFVF, FVFF, VFFF.\
\
\

## Question 1c:

------------------------------------------------------------------------

List all the outcomes that are in the event "all four mortgages are of the same type"?

### Solution to Question 1c:

------------------------------------------------------------------------

The outcomes that are in the event "all four mortgages are of the same type" are: FFFF, VVVV.\
\
\

## Question 1d:

------------------------------------------------------------------------

List all the outcomes that are in the event "at most one of the four is a variable-rate mortgage"?

### Solution to Question 1d:

------------------------------------------------------------------------

The outcomes that are in the event "at most one of the four is a variable-rate mortgage" are: FFFV, FFVF, FVFF, VFFF, FFFF.\
\
\

## Question 1e:

------------------------------------------------------------------------

What is the union of the events in parts (c) and (d), and what is the intersection of these two events?

### Solution to Question 1e:

------------------------------------------------------------------------

The union consists of the outcomes: FFFF, VVVV, FFFV, FFVF, FVFF, VFFF.\
The intersection consists of the outcomes: FFFF.\
\
\

## Question 1f:

------------------------------------------------------------------------

What is the union of the events in parts (b) and (c), and what is the intersection of these two events?

### Solution to Question 1f:

------------------------------------------------------------------------

The union consists of the outcomes: FFFV, FFVF, FVFF, VFFF, FFFF, VVVV.\
The intersection consists of the outcomes: none of the outcomes are part of the intersection of these two events.\
\
\

## Question 1g:

------------------------------------------------------------------------

Assuming 55% of all mortgages are fixed rate and 45% are variable rate, and that each mortgage selected is independent of the others that are selected, what is the probability of selecting four mortgages which has exactly one variable rate mortgage? **Show or explain work to support your answer.**

### Solution to Question 1g:

------------------------------------------------------------------------

Outcomes of selecting exactly one variable rate mortgage out of 4 mortgages: VFFF, FVFF, FFVF, FFFV.

Probability of selecting 1 outcome with exactly one variable rate mortgage out of 4 mortgages:

= 0.45 \* 0.55 \* 0.55 \* 0.55

```{r}
(0.45)*(0.55)**3
```

= 0.07486875

Probability of selecting any outcome where there is exactly 1 variable rate mortgage out of 4 mortgages:

= number of possible outcomes in the sample that meet the criteria for the event above \* probability of selecting 1 outcome that meets the event above

= 4 \* 0.07486875

```{r}
4*0.07486875
```

= 0.299475\

The probability of selecting four mortgages which has exactly one variable rate mortgage in the sample experiment is 0.299475.\
\

# Question 2:

------------------------------------------------------------------------

The population of a particular country consists of three ethnic groups. Each individual belongs to one of the four major blood groups. The accompanying joint probability table gives the proportions of individuals in the various ethnic group--blood group combinations. For example the entry $0.008$ in th e table means $P(B \cap \mbox{ Ethnic group } 1)=0.008$.

| Ethnic Group | O       | A       | B       | AB      | Total |
|--------------|---------|---------|---------|---------|-------|
| 1            | $0.082$ | $0.106$ | $0.008$ | $0.004$ | $0.2$ |
| 2            | $0.135$ | $0.141$ | $0.018$ | $0.006$ | $0.3$ |
| 3            | $0.215$ | $0.200$ | $0.065$ | $0.020$ | $0.5$ |
| Total        | $0.432$ | $0.447$ | $0.091$ | $0.03$  | $1$   |

Suppose that an individual is randomly selected from the population, and define events:

-   Event $A$: A person with blood type $A$ is selected.
-   Event $B$: A person with blood type $B$ is selected.
-   Event $C$: A person from ethnic group 3 is selected.

## Question 2a:

------------------------------------------------------------------------

Calculate $P(A)$, $P(C)$, and $P(A \cup C)$. Show your work.

### Solution to Question 2a:

------------------------------------------------------------------------

Provide supporting work along with your answer.

P($A$) = 0.106 + 0.141 + 0.200 = 0.447

```{r}
0.106+0.141+0.2
```

The probability of a person with blood type A is selected is 0.447.\

P($C$) = 0.215 + 0.2 + 0.065 + 0.02 = 0.5

```{r}
0.215+0.2+0.065+0.02
```

The probability of selecting someone from ethnic group 3 is 0.5.

P($A\cup C$) = 0.106 + 0.141 + 0.2 + 0.215 + 0.065 + 0.02 = 0.747

```{r}
0.106+0.141+0.2+0.215+0.065+0.02
```

The probability of selecting someone with blood type A or from ethnic group 3 is 0.747.

## Question 2b:

------------------------------------------------------------------------

If we know the selected person is from from ethnic group 3, what is the probability they have type A blood?

### Solution to Question 2b:

------------------------------------------------------------------------

Provide supporting work along with your answer.\
\
P($A \ | \ C$) = 0.2 / 0.5 = 0.4

```{r}
0.2/0.5
```

0.5 is the probability that the person is from ethnic group 3 out of the whole sample, and 0.2 is the probability that the person is blood type A from ethnic group 3.

The probability that the selected person has blood type A given that they are from ethnic group 3 is 0.4.

## Question 2c:

------------------------------------------------------------------------

If we know the selected person has type A blood, what is the probability they are from ethnic group 3?

### Solution to Question 2c:

------------------------------------------------------------------------

Provide supporting work along with your answer.

P($C \ | \ A$) = 0.2 / 0.447 = 0.4474273

```{r}
0.2/0.447
```

0.447 is the probability that the person is blood type A from the whole sample, and 0.2 is the probability that the person is blood type A from ethnic group 3.

The probability that the selected person is from ethnic group 3 given that they are blood type A is 0.4474273.

\

## Question 2d:

------------------------------------------------------------------------

Calculate both $P(A \vert C)$ and $P(C \vert A)$ and explain in context what each of these probabilities represents.

### Solution to Question 2d:

------------------------------------------------------------------------

Provide supporting work along with each answer.\

The probabilities have been calculated previously from question 2c and 2d.

P($A \ | \ C$) = 0.2 / 0.5 = 0.4

Out of the people from ethnic group 3, the chance of choosing someone with blood type A is 0.4.

P($C \ | \ A$) = 0.2 / 0.447 = 0.4474

Out of the people with blood type A, the chance of choosing someone from ethnic group 3 is 0.4474.\
\

## Question 2e:

------------------------------------------------------------------------

If we know the selected individual does not have type $B$ blood, what is the probability that they are from ethnic group 1?

### Solution to Question 2e:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

P($ethnic \ group \ 1 \ | \ B^C$) = (0.2 - 0.008) / (1 - 0.091) = 0.2112

Alternate way: (0.082 + 0.106 + 0.004) / (0.432 + 0.447 + 0.03) = 0.2112

```{r}
(0.2 - 0.008) / (1 - 0.091)
(0.082+0.106+0.004) / (0.432+0.447+0.03)
```

0.2 is the total people that are from ethnic group 1, 0.008 is the people that are from ethnic group 1 and are blood type B, 1 is the total people in the sample, 0.091 is the people that are blood type B. We subtract the values from each other to get the probability of not having blood type B. An alternative way to calculate the same probability is to add up all the probabilities of the event that a person selected is of blood type O, A, and AB in ethnic group 1 and divided it by the sum of the probabilities of the event that the person selected is of blood type O, A, and AB.

The probability that the selected person is from ethnic group 1 given that the selected individual does not have blood type B is 0.2112.\

# Question 3:

------------------------------------------------------------------------

Suppose that 55% of all adults regularly consume coffee, 45% regularly consume soda, and 70% regularly consume at least one of these two products.

## Question 3a:

------------------------------------------------------------------------

What is the probability that a randomly selected adult regularly consumes both coffee and soda?

### Solution to Question 3a:

------------------------------------------------------------------------

Provide supporting work along with your answer.\
\
C = coffee

S = soda

P($C \cup S$) = P($C$) + P($S$) - P($C \cap S$) =\> P($C \cap S$) = P($C$) + P($S$) - P($C \cup S$)

P($C \cap S$) = 0.55 + 0.45 - 0.7 = 0.3

```{r}
0.55 + 0.45 - 0.7
```

The probability that a randomly selected adult consumes both coffee and soda is 0.3.

\

## Question 3b:

------------------------------------------------------------------------

What is the probability that a randomly selected adult doesn't regularly consume either of these two products?

### Solution to Question 3b:

------------------------------------------------------------------------

Provide supporting work along with your answer.

P($C^C \cap S^C$) = 1 - (P($C$) + P($S$) - P($C \cap S$)) = 1 - (0.55 + 0.45 - 0.3)

Alternative way: 1 - P($C \cup S$) = 1 - 0.7

```{r}
1-(0.55+0.45-0.3)
1-0.7
```

The probability that a randomly selected adult doesn't consume coffee or soda would be 0.3.\

# Question 4:

------------------------------------------------------------------------

Most medical screening tests are not perfectly accurate. The accuracy is typically measured using two conditional probabilities:

-   The [sensitivity]{style="color: red;"} of the test is the probability that a person with the disease correctly receives a positive test. This is a [Correct Positive]{style="color: red;"}.

$$\color{red}{\mbox{Sensitivity} = P\left( + \mbox{ Test } \big| \mbox{ Has Disease} \right)}$$

-   The [specificity]{style="color: blue;"} of the test is the probability that a person without the disease correctly receives a negative test. This is a [Correct Negative]{style="color: blue;"}.

$$\color{blue}{\mbox{Specificity} = P( - \mbox{ Test } \big| \mbox{ No Disease})}$$

This problem will focus on the Mini Mental State Examination (MMSE), which is the most commonly used test for complaints of problems with memory or other mental abilities.[^1]

[^1]: <https://www.alzheimers.org.uk/about-dementia/symptoms-and-diagnosis/diagnosis/mmse-test>

-   [The sensitivity of the MMSE is 80%,]{style="color: red;"} $\color{red}{P\left( + \mbox{ Test } \big| \mbox{ Has Disease} \right) = 0.8}$.
-   [The specificity of the MMSE is 90%,]{style="color: blue;"} $\color{blue}{P\left( - \mbox{ Test } \big| \mbox{ No Disease} \right) = 0.9}$.

Approximately 10% of Americans over the age of 65 suffer from Alzheimer's dementia.[^2]

[^2]: <https://www.alz.org/alzheimers-dementia/facts-figures>

It might help to organize this information in a two-way such as the one below. For example, if 1000 Americans over the age of 65 are randomly screened by the MMSE for Alzheimer's dementia:

|        | Has Alzheimer's | Does Not Have Alzheimer's | Total   |
|--------|-----------------|---------------------------|---------|
| Test + | $80$            | $90$                      | $170$   |
| Test - | $20$            | $810$                     | $830$   |
| Total  | $100$           | $900$                     | $1,000$ |

The table above is suggested as a hint to help. You do not need to use to use it, but it might be helpful to fill in the blanks before answering the questions below.

## Question 4a:

------------------------------------------------------------------------

What is the probability that a randomly tested adult over the age of 65 will test positive?

### Solution to Question 4a:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

Probability of testing positive out of the whole sample = number of people who tested positive / total people in sample

```{r}
170 / 1000
```

The probability that a randomly tested adult over the age of 65 will test positive is 0.17.\
\

## Question 4b:

------------------------------------------------------------------------

Given that a patient tests positive for Alzheimer's, what is the probability they actually have Alzheimer's?

### Solution to Question 4b:

------------------------------------------------------------------------

Provide supporting work along with your answer.\
\
P($Alzheimer's \ | \ tested \ positive$) = number of people having Alzheimer's / number of people tested positive

```{r}
80 / 100 #wrong
80 / 170
```

The probability that a patient actually has Alzheimer's given that the patient tests positive is 0.8.

# Question 5:

------------------------------------------------------------------------

Seventy percent of all vehicles examined at a certain emissions inspection station pass the inspection. Assuming that successive vehicles pass or fail independently of one another, calculate the following probabilities:

## Question 5a:

------------------------------------------------------------------------

$P(\mbox{all of the next three vehicles inspected pass})$

### Solution to Question 5a:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

0.7 vehicles pass inspection

Probability that next 3 inspections pass = 0.7 \* 0.7 \* 0.7

```{r}
0.7**3
```

The probability of the next three vehicles passing the inspection (consecutively) is 0.343.\
\

## Question 5b:

------------------------------------------------------------------------

$P(\mbox{at least one of the next three inspected fails})$

### Solution to Question 5b:

------------------------------------------------------------------------

Provide supporting work along with your answer.\
\
Possible outcomes for the next three inspections: PPP, FFF, PFP, FPP, PPF, PFF, FPF, FFP

Possible outcomes for at least one fail in the next three inspections: FFF, PFP, FPP, PPF, PFF, FPF, FFP

Calculating the probability of at least 1 fail: (number of possible outcomes for 1 fail \* probability of 1 fail out of 3 inspections) + (number of possible outcomes for 2 fails \* probability of 2 fails out of 3) + (number of possible outcomes for 3 fails \* probability of 3 fails)

$P(\mbox{at least one of the next three inspected fails})$= (3 \* 0.3 \* 0.7 \* 0.7) + (3 \* 0.3 \* 0.3 \* 0.7) + (1 \* 0.3 \* 0.3 \* 0.3)

```{r}
(3 * 0.3 * 0.7**2) + (3 * 0.3**2 * 0.7) + (1 * 0.3**3)
```

The probability that at least one inspection out of the 3 next inspections fail is 0.657.

## Question 5c:

------------------------------------------------------------------------

$P(\mbox{exactly one of the next three inspected passes})$

### Solution to Question 5c:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

Possible outcomes of exactly 1 passed inspection: PFF, FPF, FFP

```{r}
3 * (0.3**2 * 0.7)
```

The probability of exactly one inspection passing out of the three next inspections is 0.189.

## Question 5d:

------------------------------------------------------------------------

$P(\mbox{at most one of the next three vehicles inspected passes})$

### Solution to Question 5d:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

Possible outcomes of at most 1 passed vehicle inspection: FFF, PFF, FPF, FFP

```{r}
#probability of no passes + probability of exactly 1 pass
0.3**3 + (3 * 0.7 * 0.3**2)
```

The probability that at most 1 of the next 3 vehicles pass the inspection is 0.216.

## Question 5e:

------------------------------------------------------------------------

Given that at least one of the next three vehicles passes inspection, what is the probability that all three pass?

### Solution to Question 5e:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

Possible outcomes for at least one of the next three vehicles pass inspection: PFF, FFP, FPF, PPF, PFP, FPP, PPP

```{r}
#probability of at least one of the next three vehicles pass inspection
#probability of 1 pass and 2 fails + 2 passes and 1 fail + 3 passes
at_least_one_pass <- (3 * 0.7 * 0.3**2) + (3 * 0.3 * 0.7**2) + (0.7**3)
at_least_one_pass
```

```{r}
#probability that all three vehicles pass inspection
all_pass <- 0.7**3
all_pass
```

```{r}
#probability that all three vehicles pass given that at least one of the next three vehicles pass
all_pass / at_least_one_pass
```

The probability that all three vehicles pass given that at least one of the next three vehicles pass is 0.352518.

# Question 6:

------------------------------------------------------------------------

A mail-order computer business has six telephone lines. Let $X$ denote the number of lines in use at a specified time. Suppose the probability distribution function (pdf) of $X$ is given in the accompanying table.

| $x$      | 0      | 1      | 2      | 3      | 4      | 5      | 6      |
|----------|--------|--------|--------|--------|--------|--------|--------|
| $P(X=x)$ | $0.05$ | $0.15$ | $0.20$ | $0.30$ | $0.15$ | $0.10$ | $0.05$ |

## Question 6a:

------------------------------------------------------------------------

Calculate the probability that:

i.  At most three lines are in use.

ii. Fewer than three lines are in use.

iii. At least three lines are in use.

iv. Between two and five lines, inclusive, are in use.

v.  At least four lines are not in use.

### Solution to Question 6a:

------------------------------------------------------------------------

Provide supporting work along with each answer.

i.  0.7

ii. 0.4

iii. 0.6

iv. 0.75

v.  0.4

```{r}
#i. at most three lines are in use (add all probabilities of X <= 3)
0.05 + 0.15 + 0.2 + 0.3
#ii. fewer than three lines are in use (add all probabilities of X < 3)
0.05 + 0.15 + 0.2
#iii. at least 3 lines in use (add all probabilities X >= 3)
0.3 + 0.15 + 0.1 + 0.05
#iv. between 2 and 5 lines, inclusive, in use (X >= 2 and X <=5)
0.2 + 0.3 + 0.15 + 0.1
#v. at least 4 lines not in use (when X = 0, this means 6 lines are not in use/0 lines are being used) (X <= 2)
0.05 + 0.15 + 0.2
```

## Question 6b:

------------------------------------------------------------------------

Calculate the expected value, $E(X)=\mu$, of $X$.

### Solution to Question 6b:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

E(X) is the sum of products of x and p(x)

```{r}
(0*0.05) + (1*0.15) + (2*0.2) + (3*0.3) + (4*0.15) + (5*0.1) + (6*0.05)
```

The expected value of the number of lines in use is 2.85.

## Question 6c:

------------------------------------------------------------------------

Calculate the variance, $\mbox{Var}(X)=\sigma^2$, of $X$.

### Solution to Question 6c:

------------------------------------------------------------------------

Provide supporting work along with your answer.\

$Var(X) = E(X^2) - (E(X))^2$

```{r}
EX1 <- ((0**2)*(0.05) + (1**2)*(0.15) + (2**2)*(0.2) + (3**2)*(0.3) + (4**2)*(0.15) + (5**2)*(0.1) + (6**2)*(0.05)) #E(X^2)
EX2 <- ((0*0.05) + (1*0.15) + (2*0.2) + (3*0.3) + (4*0.15) + (5*0.1) + (6*0.05))**2 #E(X)^2
EX1
EX2
cat("variance:", EX1 - EX2) #variance
```

The variance of X is 2.2275.

# Question 7

------------------------------------------------------------------------

Let $X$ be the damage incurred (in dollars) in a certain type of accident during a given year. Possible $X$ values are losses of \$0, \$1000, \$5000, and \$10,000, with probabilities $0.75$, $0.12$, $0.10$, and $0.03$, respectively. A particular company offers a \$500 deductible policy.This means:

-   When no claim is filed (a loss of \$0), no money is paid by either the driver or the insurance company.
-   If a loss of \$1000 is reported, then the drive pays a \$500 deductible and the insurance company pays the remaining \$500.
-   If a loss of \$5000 is reported, then the drive pays a \$500 deductible and the insurance company pays the remaining \$4500.
-   If a loss of \$10,000 is reported, then the drive pays a \$500 deductible and the insurance company pays the remaining \$9500.

If the company wishes its expected profit **(profit equals premium minus the amount the insurance company pays, excluding the deductible which is paid by the driver)** for each policy to be \$100, what premium amount should it charge?

## Solution to Question 7:

------------------------------------------------------------------------

Be sure to provide supporting work for your final answer.\

profits for each damage amount is \$500 except \$0 in charges

E(X) = 0(0.75) + 1000(0.12) + 5000(0.1) + 10000(0.03), this value is the expected value/average cost of loss damage for a year

Can also be interpreted as the expected value of the loss paid by both consumer and insurance company

```{r}
(0*0.75) + (1000*0.12) + (5000*0.1) + (10000*0.03) #expected cost of damage for a year
```

Below is the expected value of the deductible paid by the consumer

```{r}
expected_loss <- (0 * 0.75) + (500 * 0.12) + (4500 * 0.1) + (9500 * 0.03) #expected cost the company must pay in damages with a $500 deductible for the year (loss for the company)
expected_loss
```

To profit \$100 per policy, charge \$100 more per policy than the expected loss of the policy.

```{r}
#charge = 100 + expected_loss
profit = 100
charge <- profit + expected_loss
charge
```

To profit \$100 per policy, the premium charge should be \$895.

# Question 8

------------------------------------------------------------------------

Suppose that only 30% of all drivers come to a complete stop at an intersection having flashing red lights in all directions when no other cars are visible.

## Question 8a:

------------------------------------------------------------------------

Use R to answer the following questions. Be sure to include code cells with the command(s) you enter into R to obtain your final answer. What is the probability that, of 20 randomly chosen drivers coming to an intersection under these conditions:

i.  Exactly 7 will come to a complete stop?

ii. At most 7 will come to a complete stop?

iii. At least 7 will come to a complete stop?

### Solution to Question 8a:

------------------------------------------------------------------------

```{r}
# show your R code in one or more code cells
# Then type final answers below
dbinom(7, 20, 0.3) #exactly 7 cars will come to a stop
pbinom(7, 20, 0.3) #at most 7 cars will come to a stop
1-pbinom(6, 20, 0.3) #at least 7 will come to a stop (1 - at most 6 cars will come to a stop)
pbinom(20-7, 20, 0.7) #at most 13 will not come to a stop
```

Provide supporting work in the code cell above for each answer. Then type your answers below.

i.  0.164262

ii. 0.7722718

iii. 0.3919902

\
\

## Question 8b:

------------------------------------------------------------------------

What is the expected value for the number of drivers (out of 20) that will come to a complete stop?

### Solution to Question 8b:

------------------------------------------------------------------------

Be sure to provide supporting work for your final answer.\

```{r}
#expected value
ex <- 0
for (x in 0:20) {
  ex <- ex + (x * dbinom(x, 20, 0.3)) #calculate the expected value by adding each x * p(x)
  cat("exactly", x, "cars come to a complete stop:", dbinom(x, 20, 0.3), "\n") #x cars will come to a complete stop
}
cat("expected value:", ex)
#E(X) = np (quick formula with binomial distribution)
ex_quick <- 20 * 0.3
ex_quick
```

$E(X) = 6$

The expected number of drivers out of 20 total drivers that will make a complete stop is 6.\

## Question 8c:

------------------------------------------------------------------------

What is the variance of the number of drivers (out of 20) that will come to a complete stop?

### Solution to Question 8c:

------------------------------------------------------------------------

Be sure to provide supporting work for your final answer.\

```{r}
#variance = E(X^2) - E(X)^2
ex2 <- 0
for (x in 0:20) {
  ex2 = ex2 + (x**2 * dbinom(x, 20, 0.3))
  # cat("exactly", x, "cars come to a complete stop:", dbinom(x, 20, 0.3), "\n") #x cars will come to a complete stop
}
ex2 #E(X^2)
ex**2 #E(X)^2
var <- ex2 - ex**2
cat("variance:", var)
```

```{r}
#Var(X) = npq (quick formula to get variance in binomial distribution)
var_quick <- 20 * 0.3 * 0.7
var_quick
```

The variance of the number of drivers (out of 20) that will come to a complete stop is 4.2.\
\
