---
title: "MATH 3382: Homework 4 - Due Sep. 27 at 2PM"
author: "Ivy Truong"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

# Instructions:

------------------------------------------------------------------------

**Submit your work electronically as an R Markdown file extension `.Rmd`.**

Upload your file into the Homework 4 assignment in Canvas before **2 PM on Tues, September 27**.

**You must provide supporting work for your all of your answers.**

-   [That work might be inserting some R code cells.]{style="color: blue;"}

    -   **Feel free to insert as much R code as you like.**
    -   You must interpret or explain how the output of a code cell is being utilized to solve the problem.
    -   **Code with incomplete or missing explanations of why the output is useful may not receive full credit.**

-   [It might involve doing some algebra and/or calculus (integrating, differentiating, etc).]{style="color: blue;"}

    -   **Be sure to type/write out the steps of your work so I can see how you arrived at your answer.**
    -   **If typing takes too long no problem.** You can write out the support work on paper and hand that in before class.

-   [Probably it a little bit of both on most problems, but you decide how you want to solve the problems!]{style="color: blue;"}

## Important Academic Policies:

------------------------------------------------------------------------

-   **Be your own work.** Though you may collaborate with others, everyone is responsible for writing up the work in their own way using their own methods. Plagiarism of any form is not tolerated.
-   **Be complete.** You must provide all work and/or explanations needed to find the solution. Answers with insufficient or incomplete supporting work may lose credit.
-   **Adhere to the Code of Academic Honesty.**
-   **Be clearly written (and legible if written).** Your solution to a problem must be clear, written in complete sentences. You may lose credit for work that is unclear or hard to follow.

### [Let me know if you need help, and **GOOD LUCK!!!**]{style="color: blue;"}

------------------------------------------------------------------------

# Question 1:

------------------------------------------------------------------------

Suppose the reaction temperature $X$ (in  ${}^{\circ}$C) in a chemical process has a uniform distribution with minimum value $-5^{\circ}$C and maximum value $8^{\circ}$C.

## Question 1a:

------------------------------------------------------------------------

Replace the ??
in the LaTeX code below to complete the formula for the probability density function of $X$.

### Solution to Question 1a:

------------------------------------------------------------------------

**Replace the ?? in the LaTeX code below with an appropriate expression or value.**

\

$$f(x) = \left\{ \begin{array}{ll}
\frac{1}{13} & -5 \leq x \leq 8 \\
0 & \mbox{otherwise}
\end{array} \right. $$

Temperature is a continuous random variable.
The problem states this is a uniform distribution.
The formula pdf of a continuous uniform distribution is $\frac{1}{b-a}$.
For this problem, it is $\frac{1}{8-(-5)} = \frac{1}{13}$.\

## Question 1b:

------------------------------------------------------------------------

Compute $P(1 < X < 3)$.
Be sure to provide supporting work that shows how you found your answer.

### Solution to Question 1b:

------------------------------------------------------------------------

P(1 \< X \< 3) = 1 - (P(X \< 1) + P(X \> 3)) or P(X \< 3) - P(X \< 1)

P(X \< 3) = F(3) - F(-5)

P(X \< 1) = F(1) - F(-5)

P(1 \< X \< 3) = F(3) - F(-5) - (F(1) - F(-5)) =\> F(3) - F(-5) - F(1) + F(-5) = F(3) - F(1)

To get the probability, we need to find the area under the curve.
Take the integral of the pdf to get the cdf.

The cdf formula of a continuous uniform distribution is: $\frac{x-a}{b-a}$

To get P(1 \< X \< 3), F(3) - F(1)

```{r}
#P(1 < X < 3) = P(X < 3) - P(X < 1)
cdf3 <- (3 - (-5)) / (8-(-5))
cdf1 <- (1 - (-5)) / (8-(-5))
prob1_3 <- cdf3 - cdf1
prob1_3
```

The probability that X is greater than 1 and less than 3 is 0.1538462.\

## Question 1c:

------------------------------------------------------------------------

Compute $E(X)$ and $\mbox{Var}(X)$.
Be sure to provide supporting work that shows how you found your answer.

### Solution to Question 1c:

------------------------------------------------------------------------

E(X) = $\frac{a+b}{2}$

Var(X) = $\frac{(b-a)^2}{12}$

```{r}
ex <- ((-5) + 8) / 2
variance <- ((8-(-5))^2) / 12
ex
variance
```

The expected value of X is 1.5.
The variance of X is 14.08333.

## Question 1d:

------------------------------------------------------------------------

Give a formula for the cdf of $X$, $F(X)$.
Be sure to provide supporting work that shows how you found your answer.

### Solution to Question 1d:

------------------------------------------------------------------------

To get the cdf, we need to take the integral of the pdf.

$\int_{a}^x \frac{1}{b-a} dt$ =\> $\frac{1}{b-a} * t \ |_{a}^x$ =\>$\frac{x}{b-a} - \frac{a}{b-a}$ =\> $\frac{x-a}{b-a}$

F(X) = $\frac{x-a}{b-a}$

We also know the range of the formula given the pdf is $-5 \leq x \leq 8$

F(X) = $\frac{x-(-5)}{8-(-5)}$

The cdf formula of X, F(X) is $\frac{x+5}{13}$.\

# Question 2:

------------------------------------------------------------------------

The plasma cholesterol level (mg/dL) for patients with no prior evidence of heart disease who experience chest pain is normally distributed with mean 200 and standard deviation 35.
Consider randomly selecting an individual of this type.

## Question 2a:

------------------------------------------------------------------------

What is the $z$-score of a plasma cholesterol level that is 120 mg/dL?
Show your work, and then **interpret the practical meaning of your answer.**

### Solution to Question 2a:

------------------------------------------------------------------------

$z = \frac{x-\mu}{\sigma}$

$z = \frac{120-200}{35}$

```{r}
z <- (120 - 200) / 35
z
```

The z score is -2.285714.
This means that a person with a cholesterol level that is 120mg/dL are -2.285714 standard deviations away from the mean.
In general, this person's cholesterol level is below the mean cholesterol level by 80 mg/dL.

## Question 2b:

------------------------------------------------------------------------

If the $z$-score of a person's plasma cholesterol level that is $2.3$, then what is their plasma cholesterol level (in mg/dL)?
Provide your supporting work to justify your answer.

### Solution to Question 2b:

------------------------------------------------------------------------

$2.3 = \frac{x-200}{35}$\

```{r}
x <- (2.3 * 35) + 200
x
```

The person's plasma cholesterol level is 280.5 mg/dL if their z score is 2.3.

## Question 2c:

------------------------------------------------------------------------

What is the probability that the plasma cholesterol level is at most 120 mg/dL?
Provide supporting work below (which may involve using R code).

### Solution to Question 2c:

------------------------------------------------------------------------

P(X \< 120) = Cholesterol level of at most 120 mg/dL

```{r}
pnorm(120, 200, 35)
```

The probability that a person has a plasma cholesterol level of at most 120 mg/dL is 0.01113549.

## Question 2d:

------------------------------------------------------------------------

What is the probability that the plasma cholesterol level is between 250 and 350 mg/dL?
Provide supporting work below (which may involve using R code).

### Solution to Question 2d:

------------------------------------------------------------------------

P(250 \< X \< 350) = P(X \< 350) - P(X \< 250)

```{r}
lessthan250 <- pnorm(250, 200, 35)
lessthan350 <- pnorm(350, 200, 35)
between250_350 <- lessthan350 - lessthan250
between250_350
```

The probability that the plasma cholesterol level is between 250 and 350 mg/dL is 0.07655462.

## Question 2e:

------------------------------------------------------------------------

What is the probability that the plasma cholesterol level differs from the mean by at least $1.5$ standard deviations (cholesterol level is either above or below the mean by $1.5$ standard deviations)?
Provide supporting work below (which may involve using R code).

### Solution to Question 2e:

------------------------------------------------------------------------

P(-1.5 \< Z \< 1.5) = P(Z \< 1.5) - P(Z \< -1.5)

```{r}
zlessneg1.5 <- pnorm(-1.5, 0, 1)
# zlessneg1.5
zless1.5 <- pnorm(1.5, 0, 1)
# zless1.5
betweenneg1.5_1.5 <- zless1.5 - zlessneg1.5
betweenneg1.5_1.5
```

The probability that the plasma cholesterol level differs from the mean by at least 1.5 standard deviations is 0.8663856.

# Question 3:

------------------------------------------------------------------------

In studies of anticancer drugs it was found that if mice are injected with cancer cells, the survival time can be modeled with the exponential distribution.
Without treatment the expected survival time was 9 hours.

## Question 3a:

------------------------------------------------------------------------

What is the probability that a randomly selected mouse will survive at least 7 hours?
Provide supporting work.

### Solution to Question 3a:

------------------------------------------------------------------------

P(X \> 7)

We know that E(X) = 9 hours

$\lambda = \frac{1}{\mu} = \frac{1}{9}$

Pdf formula for exponential distribution in this problem: $f(x) = \frac{1}{9} e^{-\frac{1}{9} x}$

To get probability, take integral of pdf to get cdf.

$\int_7^{\infty} \frac{1}{9} e^{-\frac{1}{9}x} => -e^{-\frac{1}{9}x} \ |_7^\infty$\

$\lim_{N \to \infty} \left(-e^{-N/9} + e^{-7/9}\right) => 0 + e^\frac{-7}{9}$

```{r}
1 - pexp(7, 1/9) #1 - P(X < 7)
exp(-7/9)
```

The probability that the selected mouse will survive at least 7 hours is 0.4594258.

## Question 3b:

------------------------------------------------------------------------

What is the probability that a randomly selected mouse will survive at most 11 hours?
Provide supporting work.

### Solution to Question 3b:

------------------------------------------------------------------------

Probability of surviving at most 11 hours = $-e^{-\frac{1}{9}x} \ |_0^{11} = -e^{-\frac{11}{9}} - (-e^0)$\

```{r}
pexp(11, 1/9) #P(X < 11)
-exp(-11/9) + 1
```

The probability of a mouse surviving at most 11 hours is 0.7054252.

## Question 3c:

------------------------------------------------------------------------

What is the probability that a randomly selected mouse will survive between 8 and 12 hours?
Provide supporting work.

### Solution to Question 3c:

------------------------------------------------------------------------

Probability of surviving between 8 and 12 hours: $-e^{-\frac{1}{9}x} \ |_8^{12} = -e^{-\frac{12}{9}} - (-e^{-\frac{8}{9}})$\
\

```{r}
pexp(12, 1/9) - pexp(8, 1/9) #P(X < 12) - P(X < 8) = P(8 < X  < 12)
-exp(-12/9) + exp(-8/9)
```

The probability of surviving between 8 and 12 hours is 0.1475152.

## Question 3d:

------------------------------------------------------------------------

Find $E(X)$ and $\mbox{Var}(X)$.
Provide supporting work which simply be applying a "shortcut" formula.

### Solution to Question 3d:

------------------------------------------------------------------------

\
E(X) = $\frac{1}{\lambda}$

Var(X) = $\frac{1}{\lambda^2}$

```{r}
ex <- 9 #1/(1/9)
variance <- ex**2
ex
variance
# 1/((1/9)**2)
```

The expected value of X is 9.
The variance of X is 81.

## Question 3e:

------------------------------------------------------------------------

Using the value of $\sigma = \mbox{SD}(X)= \sqrt{\mbox{Var}(X)}$ from the previous part, what is the probability that the survival time of a mouse exceeds the mean survival time by more than 2 standard deviations?
Provide supporting work.

### Solution to Question 3e:

------------------------------------------------------------------------

\
SD = 9

P(Z \> 2): probability that the survival time exceeds the mean by more than 2 standard deviations

if Z = 2, and the mean is, 9, then two standard deviations away from the mean is: mean + 2SD = 9 + 2\*9 = 27

P(X \> 27): probability that the survival time exceeds 27 hours

```{r}
1 - pexp(27, 1/9) #1 - P(X < 27)
exp(-27/9)
```

The probability that the survival time of a mouse exceeds the mean survival time by more than 2 standard deviations is 0.04978707.

# Question 4:

------------------------------------------------------------------------

Let $X$ be the birth weight, in grams, of a randomly selected full-term baby.
An article[^1] suggests that $X$ is normally distributed with mean 3500 grams and standard deviation 600 grams.

[^1]: "Fetal Growth Parameters and Birth Weight: Their Relationship to Neonatal Body Composition'" (Ultrasound in Obstetrics and Gynecology, 2009: 441--446)

## Question 4a:

------------------------------------------------------------------------

What is $P(2500 < X < 4200)$?
Provide supporting work.

### Solution to Question 4a:

------------------------------------------------------------------------

Normal distribution: use pnorm

```{r}
#P(2500 < X < 4200) = P(X < 4200) - P(X < 2500)
pnorm(4200, 3500, 600) - pnorm(2500, 3500, 600)
```

The probability of a newborn being greater than 2500g and less than 4200g is 0.8305371.

## Question 4b:

------------------------------------------------------------------------

What is the probability that the weight of such a newborn is less than 4200 g?
Provide supporting work.

### Solution to Question 4b:

------------------------------------------------------------------------

```{r}
pnorm(4200, 3500, 600)
```

The probability that the weight of a newborn is less than 4200g is 0.8783275.

## Question 4c:

------------------------------------------------------------------------

What is the probability that the weight of such a newborn exceeds 5000 g (roughly 11 lb)?
Provide supporting work.

### Solution to Question 4c:

------------------------------------------------------------------------

```{r}
1 - pnorm(5000, 3500, 600) #1 - P(X < 5000)
```

The probability of a newborn exceeding 5000g is 0.006209665.

## Question 4d:

------------------------------------------------------------------------

Give value corresponding to the 95th percentile for birth weight.
Provide supporting work.

### Solution to Question 4d:

------------------------------------------------------------------------

```{r}
qnorm(0.95, 3500, 600) #calculates P(weight < X) = 0.95
```

The birth weight in the 95th percentile is 4486.912.

# Question 5:

------------------------------------------------------------------------

A large but sparsely populated county has two small hospitals, one at the south end of the county and the other at the north end.
The south hospital's emergency room has 4 beds, whereas the north hospital's emergency room has only 3 beds.
Let $X$ denote the number of south beds occupied at a particular time on a given day, and let $Y$ denote the number of north beds occupied at the same time on the same day.
**Suppose that these two random variables are independent**, and that the marginal pdf's of $X$ and $Y$ are

$$ f_X(x) = \left\{ \begin{array}{ll} 
0.1, & x=0 \\ 
0.2, & x=1 \\ 
0.3, & x=2 \\ 
0.2, & x=3 \\ 
0.2, & x=4\\
0 & \mbox{otherwise} \end{array} \right.
\quad \mbox{and} \quad
f_Y(y) = \left\{ \begin{array}{ll} 
0.1, & y=0 \\ 
0.3, & y=1 \\ 
0.4, & y=2 \\ 
0.2, & y=3\\  
0 & \mbox{otherwise} 
\end{array} \right. $$

## Question 5a:

------------------------------------------------------------------------

Display the joint pdf of $X$ and $Y$ in a joint probability table below.

### Solution to Question 5a:

------------------------------------------------------------------------

Replace each ??
with an appropriate value.

| $p(x,y)$ | $Y=0$ | $Y=1$ | $Y=2$ | $Y=3$ | Total |
|----------|-------|-------|-------|-------|-------|
| $X=0$    | 0.01  | 0.03  | 0.04  | 0.02  | 0.1   |
| $X=1$    | 0.02  | 0.06  | 0.08  | 0.04  | 0.2   |
| $X=2$    | 0.03  | 0.09  | 0.12  | 0.06  | 0.3   |
| $X=3$    | 0.02  | 0.06  | 0.08  | 0.04  | 0.2   |
| $X=4$    | 0.02  | 0.06  | 0.08  | 0.04  | 0.2   |
| Total    | 0.1   | 0.3   | 0.4   | 0.2   | 1.0   |

Since discrete random variables X and Y are independent, p(x,y) = p(x) \* p(y)

```{r}
# X <- list(0.1, 0.2, 0.3, 0.2, 0.2)
# Y <- list(0.1, 0.3, 0.4, 0.2)
# for (i in X) {
#   for (j in Y) {
#     cat(i,",", j, ":", i * j, "\n")
#   }
# }
```

## Question 5b:

------------------------------------------------------------------------

Compute $P(X \leq 1 \mbox{ and } Y \leq 1)$.
Show your supporting work below.

### Solution to Question 5b:

------------------------------------------------------------------------

$P(X \leq 1) = P(X = 0) \ and \ P(X = 1)$

$P(Y \leq 1) = P(Y = 0) \ and \ P(Y = 1)$\

P(X = 0 and Y = 0) + P(X = 0 and Y = 1) + P(X = 1 and Y = 0) + P(X = 1 and Y = 1)

```{r}
0.01 + 0.03 + 0.02 + 0.06
```

The probability that X is less than equal to 1 and Y is less than equal to 1 is 0.12.

## Question 5c:

------------------------------------------------------------------------

Calculate the probability that the total number of beds occupied at the two hospitals combined is **at most 1**.
Show your supporting work below.

### Solution to Question 5c:

------------------------------------------------------------------------

Total number of beds occupied at two hospitals is at most 1 means X + Y = 1.

Possibilities:

X = 0, Y = 1; X = 0, Y = 0; X = 1, Y = 0

P(X = 0 and Y = 0) + P(X = 0 and Y = 1) + P(X = 1 and Y = 0)

```{r}
0.01 + 0.03 + 0.02
```

The probability that the total number of beds occupied from both hospitals is at most 1 is 0.06.

# Question 6: Joint Discrete

------------------------------------------------------------------------

A market has both an express checkout line and a superexpress checkout line.
Let $X$ denote the number of customers in line at the express checkout at a particular time of day, and let $S$ denote the number of customers in line at the superexpress checkout at the same time.
Suppose the joint pdf of $X$ and $S$ is as given in the accompanying table.

| $p(x,s)$ | $S=0$  | $S=1$  | $S=2$  | $S=3$  |
|----------|--------|--------|--------|--------|
| $X=0$    | $0.08$ | $0.07$ | $0.04$ | $0.00$ |
| $X=1$    | $0.06$ | $0.15$ | $0.05$ | $0.04$ |
| $X=2$    | $0.05$ | $0.04$ | $0.10$ | $0.06$ |
| $X=3$    | $0.00$ | $0.03$ | $0.04$ | $0.07$ |
| $X=4$    | $0.00$ | $0.01$ | $0.05$ | $0.06$ |

## Question 6a:

------------------------------------------------------------------------

What is $P(X = S)$, that is, the probability that the numbers of customers in the two lines are identical?
Provide supporting work below.

### Solution to Question 6a:

------------------------------------------------------------------------

P(X = S) = P(X = 0 and S = 0) + P(X = 1 and S = 1) + P(X = 2 and S = 2) + P(X = 3 and S = 3)

```{r}
0.08 + 0.15 + 0.1 + 0.07
```

The probability that the number of customer in the two lines are identical is 0.4.

## Question 6b:

------------------------------------------------------------------------

Let $A$ denote the event that there are at least two more customers in one line than in the other line.
Calculate $P(A)$.
Provide supporting work below.

### Solution to Question 6b:

------------------------------------------------------------------------

Cases where there are at least 2 more customers in one line than in the other line

P(A) = P(X = 0, S \>= 2) + P(X = 1, S = 3) + P(X \>= 2, S = 0) + P(X \>= 3, S = 1) + P(X = 4, S = 2)

```{r}
px0 <- 0.04 + 0.0 #P(X = 0, S = 2 and S = 3)
px1 <- 0.04 #P(X = 1 and S = 3)
ps0 <- 0.05 + 0.0 + 0.0 #P(X = 2 and X = 3 and X = 4, S = 0)
ps1 <- 0.03 + 0.01 #P(X = 3 and X = 4, S = 1)
ps2 <- 0.05 #P(X = 4, S = 2)
total <- px0 + px1 + ps0 + ps1 + ps2
total
```

The probability that there are at least 2 more customers in one line than in the other line is 0.22.

## Question 6c:

------------------------------------------------------------------------

Determine the marginal pdf of $X$, **and then calculate the expected number of customers in line at the express checkout**.
Express your formula for $p_X(x)$ as a piecewise function by filling in the ??'
s marks in the LaTeX code below.

### Solution to Question 6c:

------------------------------------------------------------------------

Replace each ??
in the LaTeX code below with an appropriate value or expression.

\

$$ p_X(x) = \left\{ \begin{array}{ll} 
0.19 & x=0 \\ 
0.3 & x=1 \\ 
0.25 & x=2 \\ 
0.14 & x=3 \\ 
0.12 & x=4\\
0 & \mbox{otherwise} \end{array} \right.$$

\
Add the values in each row.

```{r}
x0 <- 0.08 + 0.07 + 0.04 + 0
x1 <- 0.06 + 0.15 + 0.05 + 0.04
x2 <- 0.05 + 0.04 + 0.1 + 0.06
x3 <- 0 + 0.03 + 0.04 + 0.07
x4 <- 0 + 0.01 + 0.05 + 0.06
x0
x1
x2
x3
x4
```

**Then calculate the expected number of customers in line at the express checkout below. Provide supporting work.**

\
$E(X) = \sum x * p(x)$

```{r}
ex <- (0 * 0.19) + (1 * 0.3) + (2 * 0.25) + (3 * 0.14) + (4 * 0.12)
ex
```

The expected number of customers in line at the express checkout is 1.7.

## Question 6d:

------------------------------------------------------------------------

Determine the marginal pdf of $S$.
Express your formula for $p_S(s)$ as a piecewise function by filling in the ??'
s marks in the LaTeX code below.

### Solution to Question 6d:

------------------------------------------------------------------------

Replace each ??
in the LaTeX code below with an appropriate value or expression.

\

$$ p_S(s) = \left\{ \begin{array}{ll} 
0.19 & s=0 \\ 
0.3 & s=1 \\ 
0.28 & s=2 \\ 
0.23 & s=3 \\ 
0 & \mbox{otherwise} \end{array} \right.$$

```{r}
s0 <- 0.08 + 0.06 + 0.05 + 0 + 0
s1 <- 0.07 + 0.15 + 0.04 + 0.03 + 0.01
s2 <- 0.04 + 0.05 + 0.1 + 0.04 + 0.05
s3 <- 0 + 0.04 + 0.06 + 0.07 + 0.06
s0
s1
s2
s3

# es <- (0 * 0.19) + (1 * 0.3) + (2 * 0.28) + (3 * 0.23)
# es
```

## Question 6e:

------------------------------------------------------------------------

By inspection of the probabilities $P(X =4)$, $P(S =0)$, and $P(X = 4 \mbox{ and } S =0)$, are $X$ and $S$ independent random variables?
Explain your answer using a complete sentence.

### Solution to Question 6e:

------------------------------------------------------------------------

P(X = 4) = 0.12

P(S = 0) = 0.19

P(X = 4 and S = 0) = 0 (from the table)

P(X = 4) \* P(S = 0) = 0.0228

```{r}
0.12 * 0.19
```

No, random variables X and S are not independent variables because P(X = 4) \* P(S = 0) is not equal to P(X = 4 and S = 0).

# Question 7:

------------------------------------------------------------------------

Each front tire of a vehicle is supposed to be filled to a pressure of $26$ psi.
Suppose the actual air pressure in each tire is a random variable, $X$ for the right tire and $Y$ for the left tire, with joint pdf $f(x,y)$ given below.

$$f(x,y) = \left\{ \begin{array}{ll} 
k(x^2+y^2)  & 20 \leq x \leq 30, \ \ \ 20 \leq y \leq 30\\
0 & \mbox{otherwise} 
\end{array} \right. .$$

## Question 7a:

------------------------------------------------------------------------

What is the value of $k$?
Provide supporting work.

### Solution to Question 7a:

------------------------------------------------------------------------

The area under the curve must equal 1.
Therefore,

$\int_{20}^{30} \ \int_{20}^{30} kx^2 + ky^2 \ dx \ dy = 1$

$\int_{20}^{30} \ \int_{20}^{30} kx^2 + ky^2 \ dx \ dy = \int_{20}^{30} \frac{1}{3}kx^3 + kxy^2 \ |_{20}^{30} \ dy = \int_{20}^{30} \ (\frac{30^3}{3}k + 30ky^2) - (\frac{20^3}{3}k + 20ky^2) \ dy = \int_{20}^{30} \ (\frac{30^3}{3}k - \frac{20^3}{3}k + 10ky^2) \ dy$\

$\frac{30^3}{3}ky - \frac{20^3}{3}ky + \frac{10}{3}ky^3 \ |_{20}^{30} = (\frac{30^4}{3}k - \frac{20^3*30}{3}k + \frac{10 * 30^3}{3}k) - (\frac{30^3*20}{3}k - \frac{20^4}{3}k + \frac{10*20^3}{3}k)$

$(\frac{30^4}{3}k - \frac{20^3*30}{3}k + \frac{10 * 30^3}{3}k) - (\frac{30^3*20}{3}k - \frac{20^4}{3}k + \frac{10*20^3}{3}k) = 1$

$(\frac{810000}{3}k-\frac{240000}{3}k + \frac{270000}{3}k) - (\frac{540000}{3}k - \frac{160000}{3}k + \frac{80000}{3}k) = 1$

```{r}
(810000-240000+270000)-(540000-160000+80000)
```

The value of k is $\frac{3}{380000}$.

## Question 7b:

------------------------------------------------------------------------

What is the probability that both tires are under-filled?
Provide supporting work.

### Solution to Question 7b:

------------------------------------------------------------------------

P(X \< 26 and Y \< 26)

$\int_{20}^{26} \ \int_{20}^{26} \frac{3}{380000}x^2 + \frac{3}{380000}y^2 \ dx \ dy = \int_{20}^{30} \ \frac{1}{380000}x^3 + \frac{3}{380000}y^2 \ |_{20}^{26} \ dy = \int_{20}^{30} \ (\frac{1}{380000}26^3 + \frac{3}{380000}y^2) - (\frac{1}{380000}20^3 + \frac{3}{380000}y^2) \ |_{20}^{26} \ dy = \int_{20}^{26} \ \frac{26^3}{100} - \frac{20^3}{100} \ dy$\
\
$= \frac{26^3}{190000}y - \frac{20^3}{190000}y \ |_{20}^{26} = \frac{26^4}{190000} - \frac{20^4}{190000}$\

```{r}
((26**4)/380000) - ((20**4)/380000)
```

The probability that both tires are under-filled is 0.7815158.

## Question 7c:

------------------------------------------------------------------------

Determine the (marginal) distribution of air pressure in the right tire alone.
Provide supporting work.

### Solution to Question 7c:

------------------------------------------------------------------------

$f_x(x) = \int_{-\infty}^{\infty} f(x, y) \ dy => \int_{20}^{30} \frac{3}{380000}x^2 + \frac{3}{380000}y^2 \ dy$\

$= \frac{3}{380000}x^2y + \frac{1}{380000}y^3 \ |_{20}^{30} = (\frac{3 * 30}{380000}x^2 + \frac{30^3}{380000}) - (\frac{3 * 20}{380000}x^2 + \frac{20^3}{380000}) = \frac{30}{380000}x^2 + \frac{19000}{380000}$\
\

## Question 7d:

------------------------------------------------------------------------

Are $X$ and $Y$ independent random variables?
Provide supporting work/explanation.

### Solution to Question 7d:

------------------------------------------------------------------------

Let's use this as an example: P(X \< 26 and Y \< 26), P(X \< 26), P(Y \< 26)

From 7b: P(X \< 26 and Y \< 26) = 0.7815158

If they were independent, then $f(x,y) = f_x(x) \cdot f_y(y)$\
\
$f_x(x) = \frac{30}{380000}x^2 + \frac{19000}{380000}$\

$f_y(y) = \frac{19000}{380000} + \frac{30}{380000}y^2$

$P(X < 26) = f_x(26) - f_x(20) = (\frac{30}{380000}26^2 + \frac{19000}{380000}) - (\frac{30}{380000}20^2 + \frac{19000}{380000}) = 0.0217894737$

$P(Y < 26) = f_y(26) - f_y(20) = (\frac{19000}{380000} + \frac{30}{380000}26^2) - (\frac{19000}{380000} + \frac{30}{380000}20^2) = 0.0217894737$

P(X \< 26) \* P(Y \< 26) = 0.0004747812

Since $f(x,y) is \ not \ equal \ to f_x(x) \cdot f_y(y)$, X and Y variables are dependent.

## Question 7e:

------------------------------------------------------------------------

Compute $P(Y \geq 25)$.
Provide supporting work, and interpret the meaning of your answer to this part in practical terms.

### Solution to Question 7e:

------------------------------------------------------------------------

$P(Y \geq 25) = 1 - P(Y < 25)$\

$P(Y < 25) = f_y(25) - f_y(20) = (\frac{19000}{380000} + \frac{30}{380000}25^2) - (\frac{19000}{380000} + \frac{30}{380000}20^2) = 0.0177631579$\

$P(Y \geq 25) = 1 - 0.0177631579 = 0.9822368421$\

The probability that the tire pressure of the left tire will be at least 25 psi is 0.9822368421.

# Question 8:

------------------------------------------------------------------------

You install an incandescent lightbulb and an LED lightbulb in a lamp.
Let $X$ be the lifetime of the fluorescent bulb and $Y$ be the lifetime of the LED lightbulb (both measured in ten thousands hours).
Suppose that $X$ and $Y$ are independent and that each has an exponential distribution with the mean lifetime of $X$ and $Y$ being $\mu_X =2$ and $\mu_Y = 5$ (meaning the average lifetime of each is $20,\!000$ and $50,\!000$ hours).

## Question 8a:

------------------------------------------------------------------------

What are the values of $\lambda_X$ and $\lambda_Y$?
Provide supporting work.

### Solution to Question 8a:

------------------------------------------------------------------------

Since it is an exponential distribution, we know that $\lambda = \frac{1}{\mu}$

$\lambda_x = \frac{1}{2}$

$\lambda_y = \frac{1}{5}$\
\

## Question 8b:

------------------------------------------------------------------------

What is the joint pdf $f(x,y)$ of $X$ and $Y$?
Provide supporting work.

### Solution to Question 8b:

------------------------------------------------------------------------

The problem states that random variables X and Y are independent.
Therefore,

$f(x,y) = f_x(x) * f_y(y)$

The pdf formula for exponential distribution is: $f(x) = \lambda e^{-\lambda x}$

$f_x(x) = \frac{1}{2}e^{-\frac{1}{2}x}$

$f_y(y) = \frac{1}{5}e^{-\frac{1}{5}y}$

$f_x(x) * f_y(y) = \frac{1}{2}e^{-\frac{1}{2}x} * \frac{1}{5}e^{-\frac{1}{5}y} = \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y}$\

$f(x,y) = \left\{ \begin{array}{ll} \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y} & \ x \geq 0, \ \ \ y \geq 0\\ 0 & \ \mbox{otherwise} \end{array} \right. .$\
\

## Question 8c:

------------------------------------------------------------------------

What is the $P(X \leq 3 \mbox{ and } Y \leq 3)$?
Provide supporting work, and interpret the meaning of your answer in practical terms.

### Solution to Question 8c:

------------------------------------------------------------------------

Formula for computing probabilities of continuous random variables:

$\int_c^d \ \int_a^b \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y} \ dx \ dy$

$P(X \leq 3, Y \leq 3) = \int_0^3 \ \int_0^3 \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y} \ dx \ dy$\
\
Solving innermost integral:

$\int_0^3 \ \frac{-1}{5}e^{-\frac{1}{2}x - \frac{1}{5}y} \ |_0^3 \ dy = \frac{-1}{5}e^{\frac{-3}{2}-\frac{1}{5}y} + \frac{1}{5}e^{0-\frac{1}{5}y}$\

Solving outermost integral:

$\int_0^3 \ \frac{-1}{5}e^{\frac{-3}{2}-\frac{1}{5}y} + \frac{1}{5}e^{-\frac{1}{5}y} \ dy = e^{\frac{-3}{2}-\frac{1}{5}y} - e^{-\frac{1}{5}y} \ | _ 0^3 = (e^{\frac{-3}{2}-(\frac{1}{5}*3)} - e^{-\frac{1}{5}*3}) - (e^{\frac{-3}{2}-(\frac{1}{5} * 0)} - e^{-\frac{1}{5} * 0}) = (e^\frac{-21}{10} - e^\frac{-3}{5}) - (e^\frac{-3}{2} - 1)$

```{r}
(exp(-21/10) - exp(-3/5)) - (exp(-3/2) - exp(0))
```

The probability that the lifetime of light bulb X is at most 30000 hours and the lifetime of light bulb Y is at most 30000 hours is 0.3505146.

## Question 8d:

------------------------------------------------------------------------

What is the probability that the total lifetime of the two bulbs, $X+Y$, is at most $70,\!000$ hours.
Provide supporting work.

*Hint: Draw a picture of the region* $A = \left\{ (x, y): x \geq 0, y \geq 0, x + y \leq 7\right\}$ before integrating.

### Solution to Question 8d:

------------------------------------------------------------------------

```{r}
# par(mfrow=c(1,2))  # plot in a 1 by 2 array
# 
# y <- seq(0, 15, length=100)
# fy <- dexp(y, (1/5) * exp(-1/5))
# 
# x<- seq(0, 15, length=100)
# fx <- dexp(x, (1/2) * exp(-1/2))
# 
# 
# plot(y, fy, 
#      type="l", 
#      lty=1, 
#      col = "blue",
#      xlab="Lifetime(10,000 hours)",
#      ylab="Density", 
#      main="Lightbulb Y Lifetime")
# 
# lines(x, fx, col="red")
# lines(x, 7 - x, col="darkgreen")
# 
# par(mfrow=c(1,1)) # return to single plot
```

$\int_0^{7} \ \int_0^{7-x} \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y} \ dy \ dx$

In this case, the values are dependent on each other, so y changes with respect to x since x+y must be less than or equal to 7.

$x + y \leq 7 => y \leq 7 - x \ (bound \ for \ y \ value)$

$\int_0^{7} \ \int_0^{7-x} \frac{1}{10}e^{-\frac{1}{2}x-\frac{1}{5}y} \ dy \ dx = \int_0^7 \frac{-1}{2}e^{\frac{-1}{2}x - \frac{1}{5}y} \ |_0^{7-x} \ dx = \int_0^7 (\frac{-1}{2}e^{\frac{-1}{2}x - \frac{1}{5}(7-x)}) - \frac{-1}{2}e^{\frac{-1}{2}x - \frac{1}{5}*0} \ dx =$

$\int_0^7 (\frac{-1}{2}e^{\frac{-1}{2}x - (\frac{7}{5}-\frac{1}{5}x)}) - \frac{-1}{2}e^{\frac{-1}{2}x} \ dx = \int_0^7 (\frac{-1}{2}e^{\frac{-3}{10}x - \frac{7}{5}}) - \frac{-1}{2}e^{\frac{-1}{2}x} \ dx = \frac{5}{3}e^{\frac{-3}{10}x - \frac{7}{5}} - e^{\frac{-1}{2}x} \ |_0^7 = (\frac{5}{3}e^{\frac{-3}{10}*7 - \frac{7}{5}} - e^{\frac{-1}{2}*7}) - (\frac{5}{3}e^{\frac{-3}{10}*0 - \frac{7}{5}} - e^{\frac{-1}{2}*0})$

$(\frac{5}{3}e^{\frac{-35}{10}} - e^{\frac{-7}{2}}) - (\frac{5}{3}e^{-\frac{7}{5}} - e^{0})$

```{r}
((5/3)*exp(-35/10) - exp(-7/2)) - ((5/3)*exp(-7/5) - 1)
```

The probability that the light bulb lifetime of both light bulbs at the same time is at most 70,000 hours is 0.6091366.

# Question 9: Discrete Continuous

------------------------------------------------------------------------

The two most common types of errors made by programmers are syntax errors and logic errors.
Let $X$ denote the number of syntax errors and $Y$ the number of logic errors on the first run of a program.
Suppose $X$ and $Y$ have the following joint pdf for a particular programming assignment:

| $p(x,y)$ | $X=0$  | $X=1$  | $X=2$  | $X=3$  |
|----------|--------|--------|--------|--------|
| $Y=0$    | $0.71$ | $0.03$ | $0.02$ | $0.01$ |
| $Y=1$    | $0.04$ | $0.06$ | $0.03$ | $0.01$ |
| $Y=2$    | $0.03$ | $0.03$ | $0.02$ | $0.01$ |

## Question 9a:

------------------------------------------------------------------------

Find the marginal pdf's of $X$ and $Y$.
Express your formulas for $p_X(x)$ and $p_Y(y)$ as piecewise functions by filling in the ??'
s marks in the LaTeX code below.

### Solution to Question 9a:

------------------------------------------------------------------------

Replace each ??
in the LaTeX code below with an appropriate value or expression.

\

$$ p_X(x) = \left\{ \begin{array}{ll} 
0.78 & x=0 \\ 
0.12 & x=1 \\ 
0.07 & x=2 \\ 
0.03 & x=3 \\ 
0 & \mbox{otherwise} \end{array} \right.$$

```{r}
x0 <- 0.71 + 0.04 + 0.03
x1 <- 0.03 + 0.06 + 0.03
x2 <- 0.02 + 0.03 + 0.02
x3 <- 0.01 + 0.01 + 0.01
x0
x1
x2
x3
```

Replace each ??
in the LaTeX code below with an appropriate value or expression.

\

$$ p_Y(y) = \left\{ \begin{array}{ll} 
0.77 & y=0 \\ 
0.14 & y=1 \\ 
0.09 & y=2 \\ 
0 & \mbox{otherwise} \end{array} \right.$$

```{r}
y0 <- 0.71 + 0.03 + 0.02 + 0.01
y1 <- 0.04 + 0.06 + 0.03 + 0.01
y2 <- 0.03 + 0.03 + 0.02 + 0.01
y0
y1
y2
```

Are $X$ and $Y$ independent?
How can you tell?

They are dependent.
It only takes one combination to prove that p(x,y) is not equal to p(x) \* p(y).
Let's take X = 0 and Y = 0 as an example.

P(X = 0) = 0.78

P(Y = 0) = 0.77

P(X = 0 and Y = 0) = 0.71

P(X = 0) \* P(Y = 0) = 0.6006

## Question 9b:

------------------------------------------------------------------------

What is the average number of syntax errors in the first run of a program?
What is the average number of logic errors?
Answer both questions, and be sure to provide supporting work.

### Solution to Question 9b:

------------------------------------------------------------------------

E(X) = expected number of syntax errors

E(Y) = expected number of logic errors

```{r}
ex <- (0 * 0.78) + (1 * 0.12) + (2 * 0.07) + (3 * 0.03)
ey <- (0 * 0.77) + (1 * 0.14) + (2 * 0.09)
ex
ey
```

The average number of syntax errors in the first run of a program is 0.35.
The average number of logic errors in the first run of a program is 0.32.

## Question 9c:

------------------------------------------------------------------------

Suppose an evaluator assigns points to each program with the formula $Z=100-4X-9Y$.
What is the expected point score, $Z$, for a randomly selected program?
Provide supporting work.

### Solution to Question 9c:

------------------------------------------------------------------------

\
Z = 100 - 4(0.35) - 9(0.32)

```{r}
z <- 100 - (4 * 0.35) - (9 * 0.32)
z
```

The expected point score for a randomly selected program is 95.72.
