---
title: "MATH 3382: Homework 9 - Not Collected (but is on Exam 2)"
author: "Enter your name"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

# Instructions: This Work is Not Being Collected

------------------------------------------------------------------------

-   Though your work is not being collected, most of the concepts on this homework assignment will be covered on Exam 2.

    -   The material that is not on exam 2 is indicated as optional.

-   It is very important to get practice and feedback on this material before the exam.

    -   Solutions will be posted.
    -   Compare your work to the solutions.
    -   If you have any questions, feel free to let me know!

# Question 1

------------------------------------------------------------------------

Errors in medical prescriptions occur and a study[^1] examined whether electronic prescribing may help reduce errors.
Two groups of doctors used written prescriptions and had similar error rates before the study.
One group switched to e-prescriptions while the other continued to use written prescriptions, and error rates were measured one year later.

[^1]: Kaushal, R., et al., "Electronic Prescribing Improves Medical Safety in Community-Based Office Practices," *Journal of General Internal Medicine*, June 2010

| Type       | Error | No Error | Total |
|------------|-------|----------|-------|
| Electronic | 254   | 3594     | 3848  |
| Written    | 1478  | 2370     | 3848  |

## Question 1a

------------------------------------------------------------------------

Find a 95% confidence interval for the difference in the proportion of errors between the two groups using the Wald confidence interval for the difference between two proportions.

### Solution to Question 1a

------------------------------------------------------------------------

**Complete the R code block below to compute the cutoffs for a 95% Wald Confidence Interval.**

```{r}
# Calculate the sample proportion for each separately
elec.p <- ??
write.p <- ??

# Approximate the SE using Wald Conf Int
wald.se <- ??

# Compute the z-alpha-over-2 value for 95% Conf Int
z.star <- ??
  
# Compute the Margin of Error  
moe <- z.star * ??

# Compute upper and lower cutoffs for Conf Int
lower.q1 <- ??
upper.q1 <- ??

# Print answers to the screen
lower.q1
upper.q1
```

## Question 1b

------------------------------------------------------------------------

Find a 95% confidence interval for the difference in the proportion of errors between the two groups using the `prop.test` function.

### Solution to Question 1b

------------------------------------------------------------------------

**Complete the R code block below to compute the cutoffs for a 95% Wald Confidence Interval.**

```{r}
prop.test(??)$conf.int
```

## Question 1c

------------------------------------------------------------------------

Are your solutions from [Question 1b] and [Question 1c] equal?

### Solution to Question 1c

------------------------------------------------------------------------

-   If your answers are equal, explain why.
-   If your answers are not equal, explain why not and state which of the two solutions is more accurate.

## Question 1d

------------------------------------------------------------------------

Interpret the meaning of your interval from [Question 1b] in practical terms.
Can we claim that there is a difference in the proportion of errors made?

### Solution to Question 1d

------------------------------------------------------------------------

# Question 2

------------------------------------------------------------------------

State whether each set of hypotheses is valid for a statistical test.
If not, explain why not.
If so, would you use a one-tail or two-tail test.

a.  $H_0: \mu = 50$ and $H_a: \mu > 50$.

b.  $H_0: p_a - p_b <0$ and $H_a: p_a - p_b=0$.

c.  $H_0: \mu\_1 - \mu\_2 \ne 0$ and $H_a: \mu_1 -\mu_2 =0$.

d.  $H_0: p_a = p_b$ and $H_a: p_a \ne p_b$.

e.  $H_0: \bar{x} = 5$ and $H_a: \bar{x} \ne 5$.

f.  $H_0: p = 0.25$ and $H_a: p>0.5$.

g.  $H_0: p = 0.25$ and $H_a: p>0.25$.

h.  $H_0: \mu = 10$ and $H_a: \mu \ne 10$.

## Solution to Question 2

------------------------------------------------------------------------

a.  

b.  

c.  

d.  

e.  

f.  

g.  

h.  

\
\

# Question 3

------------------------------------------------------------------------

A casino wants to see if a six-sided die is fair.
They perform an experiment in which they roll the die 600 times and count the number of times the result is divisible by 3 (in other words they count the number of times a 3 or a 6 is rolled).

a.  Is this a one or two-sided test?
    Explain your answer in a complete sentence.

b.  Set up the hypotheses for this test both in words and in terms the parameter $p$ which denotes the theoretical probability for getting a roll of the die giving a number that is divisible by 3.

c.  Out of 600 rolls, they find that 180 result in a number that is divisible by 3 being rolled.
    What is the observed test statistic?
    Use appropriate notation for your answer.

d.  [Optional not on exam 2: What is the nulll distribution for this example?]{style="color: blue;"}

e.  [Optional not on exam 2:Calculate the $P$-value of the observed test statistic using the null distribution you identified in the previous part.]{style="color: blue;"}

f.  [Optional not on exam 2: Interpret the practical meaning of the $P$-value. For example write out a complete sentence that explains what the $P$-value is measuring.]{style="color: blue;"}

## Solution to Question 3

------------------------------------------------------------------------

a.  

b.  

c.  

d.  [Optional not on exam 2]{style="color: blue;"}

e.  [Optional not on exam 2]{style="color: blue;"}

f.  [Optional not on exam 2]{style="color: blue;"}

# Question 4

------------------------------------------------------------------------

In a hypothesis testing $H_0$: $\mu_1-\mu_2=0$ versus $H_a$: $\mu_1 - \mu_2 > 0$, which of the following test statistics would have the smallest $P$-value?

a.  $\bar{x}_1-\bar{x}_2 = -0.25$

b.  $\bar{x}_1-\bar{x}_2 = -2.5$

c.  $\bar{x}_1-\bar{x}_2 = 1.5$

d.  $\bar{x}_1-\bar{x}_2 = 0.15$

## Solution to Question 4

------------------------------------------------------------------------

**Support your answer with a one or two sentence explanation.**

\
\
\

# Question 5

------------------------------------------------------------------------

In a hypothesis testing $H_0$: $p_1-p_2=0$ versus $H_a$: $p_1 - p_2 \ne 0$, which of the following test statistics is the most statistically significant?

a.  $\hat{p}_1-\hat{p}_2 = -0.25$

b.  $\hat{p}_1-\hat{p}_2 = -0.05$

c.  $\hat{p}_1-\hat{p}_2 = 0.005$

d.  $\hat{p}_1-\hat{p}_2 = 0.15$

## Solution to Question 5

------------------------------------------------------------------------

**Support your answer with a one or two sentence explanation.**

\
\
\

# Question 6

------------------------------------------------------------------------

Suppose you do an experiment as follows.
Five volunteers wish to quit smoking cigarettes.
They are randomly split into two groups.
Two people are placed into a reward group.
If after six months they have successfully quit, they receive $\$100$.
The other three people make a $\$50$ deposit.
If after six months they have successfully quit, they receive $\$150$.
If they have not quit smoking after six months, they lose their $\$50$ deposit.
The results of the experiment are given below.
$S$ denotes they successfully quit.
$N$ denotes they were not successful.

| Group       | Result                |
|-------------|-----------------------|
| Reward Only | $S_1$ , $F_1$         |
| Deposit     | $S_2$ , $F_2$ , $F_3$ |

a.  If researchers want to test to see if the proportion of people that quit is higher when they are rewarded as opposed to when they are penalized by losing a deposit, set up the hypotheses for this test.
    State your hypotheses in words, and using appropriate notation.

b.  Compute the observed test statistic.
    That is, find the difference in the proportion of people in each sample that successfully quit smoking after six months.

c.  [Optional not on exam 2: How many total permutation resamples are possible? In other words, out of a total of 5 people, how many ways can we select two people to be in the reward only group?]{style="color: blue;"}

d.  [Optional not on exam 2: Write out (by hand) all possible permutations resamples and calculate the difference in sample proportions for each resample.]{style="color: blue;"}

e.  [Optional not on exam 2: Based on your previous work, what is the $P$-value of the observed test statistic?]{style="color: blue;"}

## Solution to Question 6

------------------------------------------------------------------------

a.  

b.  

c.  [Optional not on exam 2]{style="color: blue;"}

d.  [Optional not on exam 2]{style="color: blue;"}

e.  [Optional not on exam 2]{style="color: blue;"}
