---
title: '09: Sampling Distributions'
author: "Adam Spiegler, University of Colorado Denver"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

## Importing Packages

------------------------------------------------------------------------

```{r, message = FALSE}
library(dplyr)
```

# Shiny Apps

------------------------------------------------------------------------

[`Shiny`](https://shiny.rstudio.com) is an R package that creates interactive apps/widgets.
If you have not install the `Shiny` package before, in the console first run the command `install.packages("shiny")`.

-   See <https://shiny.rstudio.com/tutorial/>.

```{r, message = FALSE}
library(shiny) # loads shiny package
```

# Set the Working Directory

------------------------------------------------------------------------

## Identifying the Working Directory

------------------------------------------------------------------------

-   In `R`, the command `getwd()` will identify your current working directory.
-   By default, the working directory should be the same as the directory where you have saved this file.

```{r, eval = FALSE}
getwd()
```

## Changing the Working Directory.

------------------------------------------------------------------------

-   If you would like to change your working directory to some other location, you can use the `setwd()` command.

```{r, eval = FALSE}
# Sample of a typical file path on Mac
#setwd("/Users/adamspiegler/Documents/my/working/directory")

# Sample of a typical file path on Windows
#setwd("c:/Documents/my/working/directory")
```

## Downloading Shiny App Files

------------------------------------------------------------------------

We will be using 3 different **Shiny Apps** in this worksheet that you will need to save locally on your computer.

-   Create three folders named **clt_bmi**, **clt_wait**, and **clt_quake**.

1.  In the folder you named **clt_bmi**, save the file **app.R** linked here: [link to BMI app](https://www.dropbox.com/s/gyg67fnqn7s12qf/app.R?dl=0)

2.  In the folder you named **clt_wait**, Save the file **app.R** linked here: [link to Wait Time app](https://www.dropbox.com/s/ns71bugx2lzp1nu/app.R?dl=0)

3.  In the folder you named **clt_quake**, Save the file **app.R** linked here: [link to Quake app](https://www.dropbox.com/s/9ipgek818m8s6jc/app.R?dl=0)

-   [Do not change the folder or file names. By default, Shiny App scripts are named **app.R**]{style="color: blue;"}.

# Sampling Distributions

------------------------------------------------------------------------

-   A [**sampling distribution**]{style="color: blue;"} is the distribution of sample statistics (such as a mean, proportion, median, maximum, etc.) computed for **different samples of the same size from the same population**.
    A sampling distribution shows us how the sample statistic varies from sample to sample.

-   The problems below compare the sampling distributions for means from three different distributions.

    -   Let $X$ denote the distribution of Body Mass Index (BMI) of all adult men.
    -   Let $Y$ denote all times (in minutes) that people wait before their train arrives at a certain stop.
    -   Let $Z$ denote the depth (in km) of all earthquakes that have occurred near Fiji since 1964.

## Question 1: Plotting a Normal Population

------------------------------------------------------------------------

Let $X$ denote the distribution of BMI of all adult men.
We can approximate this distribution by $X \sim N(26, 4)$.
Interpret the code below.
Add comments to explain what each command will do.
Then run the code.

```{r}
# Create a vector name bmi of 100 bmi values 
# chosen between x=10 and x=42
bmi <- seq(26-4*4, 26+4*4, length=100)

# Add your comment here
# get the height of each BMI value
pdf.bmi <- dnorm(bmi, 26, 4)

# Add your comment here
#plot a group of points and draw a line through those 100 points
plot(bmi, pdf.bmi, 
     type="l", lty=1, # type="l" draws line lty=1 is solid line
     xlab="Body Mass Index (BMI)",
  ylab="Density", main="Distribution of Population")
```

### Solution to Question 1:

------------------------------------------------------------------------

Enter comments in code cell above.

\
\

## Question 2: Picking One Random Sample

-   We can pick a random sample of size $n$ from a normal distribution $N(\mu, \sigma)$ using `rnorm(n, mean, sd)`.

Replace each question mark in the code below to randomly select 4 individual BMI's from the population $X \sim N(26,4)$.

```{r, eval=FALSE}
my.sample <- rnorm(4, 26, 4) #Randomly picks 4 values from N(26,4)
my.sample
```

### Solution to Question 2:

------------------------------------------------------------------------

Replace each ??
in the code cell above with an appropriate value.

\
\

## Question 3: Comparing Statistics and Parameters

------------------------------------------------------------------------

Calculate the mean and standard deviation of your sample by running the code below.
Then:

-   Discuss how the statistics of your sample will compare to the population parameters $\mu = 26$ and $\sigma =4$?
-   Discuss how the statistics of your sample will compare to the statistics that others in class obtain with their own samples?

```{r}
mean(my.sample)
sd(my.sample)
```

### Solution to Question 3:

------------------------------------------------------------------------

\
Each student has a different sample space (4 different people in each sample)

sample mean = x bar

sample standard deviation = s

sample mean and standard deviation can vary for each sample

sampling distribution: distribution of all sample means selected from groups of random samples all size n\

## Plotting a Sampling Distribution with $n=4$

------------------------------------------------------------------------

A sample of $n=4$ adult men are randomly selected.
The mean BMI of the sample is calculated: $$ \bar{x} = \frac{x_1 + x_2 + x_3+x_4}{4}.$$

Then another random sample of $n=4$ adult men are randomly selected, and again the mean BMI of this sample is computed.
This is repeated 1000 times (each sample size $n=4$), and the sampling distribution for the mean BMI can be constructed with the code below:

```{r}
# creates an empty vector to store results
n4.bmi.bar <- numeric(1000) 

# A for loop that generates 1000 random samples 
# Each size n=4, and calculates the sample mean.
for (i in 1:1000)
{
  n4.bmi.sample <- rnorm(4, 26, 4) #Randomly picks 4 values from N(26,4)
  #to make it an exponential distribution, change rnorm to rexp
  n4.bmi.bar[i] <- mean(n4.bmi.sample)
}

# Plot the sampling distribution
hist(n4.bmi.bar, xlim = c(14, 38), 
     xlab = "Mean BMI of Sample",
     main = "Sampling Distribution of Mean BMI for n=4",
     xaxt='n')
axis(1, at=seq(14, 38, 4), pos=0)
abline(v = 26, col = "red", lwd = 2, lty = 2)
```

## Question 4: Center and Spread of the Sampling Distribution

------------------------------------------------------------------------

In the R code block below, enter commands to compute the center (as measured by the mean) and spread (as measured by the standard deviation) of the sampling distribution when $n=4$.

-   Then comment on how these values compare to the population parameters $\mu=26$ and $\sigma =4$.

```{r}
mean(n4.bmi.bar) #this mean is pretty close to the population mean
sd(n4.bmi.bar) #this std is much lower than the population std
```

### Solution to Question 4:

------------------------------------------------------------------------

\
notation for statistics of all random samples generated:

$\mu_\bar{x} = mean \ of \ all \ random \ samples \ generated$

$\sigma_\bar{x} = standard \ deviation \ of \ all \ random \ samples \ generated$\
\

## Shape of the Sampling Distribution

------------------------------------------------------------------------

We can use a **Quantile-Quantile** plot (called a **qqplot**) to compare the shape of our sampling distribution to the standard normal distribution $N(0,1)$.

-   The closer the points are to the line, the more normal the distribution.
-   The plot below seems mostly normal in the middle, but the tails are slightly deviating from the tails of a normal distribution.

```{r}
qqnorm(n4.bmi.bar)
qqline(n4.bmi.bar)
```

## Question 5: Center, Shape, and Spread BMI Sampling Distribution

------------------------------------------------------------------------

Run the app below to investigate properties of the sampling distribution for the mean BMI, $\mu_X$ using the distribution $X \sim N(26,4)$.
Based your observations, complete the table below.

### Solution to Question 5 for BMI:

------------------------------------------------------------------------

```{r, eval = FALSE, message = FALSE}
runApp("clt_bmi") # be sure you've loaded shiny package and setwd
```

| Property           | Population | $n=4$  | $n=9$  | $n=16$ | $n=81$ |
|--------------------|------------|--------|--------|--------|--------|
| Shape              | Normal     | Normal | Normal | Normal | Normal |
| Mean               | 26         | 25.98  | 26.04  | 25.97  | 26.01  |
| Standard Deviation | 4          | 2.0387 | 1.3497 | 0.9996 | 0.4513 |

-spread in sample distribution stats is getting smaller: the standard deviation of the sampling distribution is getting smaller =\> standard error is getting smaller

$\sigma_\bar{x} = SE$

The larger the sample size gets, the more similar the means and std would be because you are choosing the people multiple times for each sample.

## Question 6: Center, Shape, and Spread Wait Time Sampling Distribution

------------------------------------------------------------------------

Run the app below to investigate properties of the sampling distribution for the mean wait time $\mu_Y$ between successive trains at a certain train stop using the distribution $Y \sim \mbox{Exp} \left( \frac{1}{40} \right)$.
Based your observations, complete the table below.

### Solution to Question 6 for Wait Times:

------------------------------------------------------------------------

```{r, eval = FALSE, message = FALSE}
runApp("clt_wait") # be sure you've loaded shiny package and setwd
```

| Property           | Population   | $n=4$        | $n=9$        | $n=16$       | $n=81$        |
|--------------------|--------------|--------------|--------------|--------------|---------------|
| Shape              | Skewed Right | Skewed Right | Skewed Right | Skewed Right | Almost Normal |
| Mean               | 40           | 39.68        | 39.58        | 39.78        | 40.14         |
| Standard Deviation | $\sqrt{40}$  | 19.4088      | 12.9812      | 9.8321       | 4.4347        |

As the sample size gets larger, the sampling distribution becomes more normal even thought the population is skewed.

## Working with Empirical Data: Earthquake Depth

------------------------------------------------------------------------

The dataset `quakes` in the `dplyr` package provides the following summary:

> give the locations of 1000 seismic events of MB \> 4.0.
> The events occurred in a cube near Fiji since 1964.

-   `lat`: Latitude of event
-   `long`: Longitude
-   `depth`: Depth (km)
-   `mag`: Richter Magnitude
-   `stations`: Number of stations reporting

### Numerical Summary of Quakes Data

------------------------------------------------------------------------

```{r}
#library(dplyr)  # requires dplyr package
summary(quakes)
```

### Graphical Summary of Depth of Quakes

------------------------------------------------------------------------

```{r}
plot(density(quakes$depth), 
         xlab = "Depth (in km)",
         main = "Depths of All Earthquakes in Fiji Since 1964",
         xaxt='n')
    axis(1, at=seq(-100, 800, 100), pos=0)
    abline(v = mean(quakes$depth), col = "red", lwd = 2, lty = 2)
```

## Question 7: Center, Shape, and Spread Wait Time Sampling Distribution

------------------------------------------------------------------------

Run the app below to investigate properties of the sampling distribution for the mean depth of earthquakes near Fiji since 1984, $\mu_Z$, using the observed data in `quakes`.
Based your observations, complete the table below.

### Solution to Question 7 for Earthquake Depth:

------------------------------------------------------------------------

```{r, eval = FALSE}
runApp("clt_quake") # be sure you've loaded shiny package and setwd
```

```{r}
mean(quakes$depth)
sd(quakes$depth)
```

| Property           | Population | $n=4$    | $n=9$         | $n=16$        | $n=81$        |
|--------------------|------------|----------|---------------|---------------|---------------|
| Shape              | Bimodal    | Bimodal  | Almost Normal | Almost Normal | Almost Normal |
| Mean               | 331        | 306.01   | 313.56        | 312.5         | 311.7         |
| Standard Deviation | 215.5      | 111.4113 | 72.6518       | 54.0547       | 22.6241       |

The sampling distribution becomes almost normal for small sample sizes because the population distribution is split about half of the data on the left and right of the mean.
It is like two normal distributions in one distribution.

# Notation for Population, Mean, and Distribution of Sample Means

------------------------------------------------------------------------

When describing the **mean** of a distribution we use the notation:

-   Population mean: $\mu_X$
-   Sample mean: $\bar{x}$
-   Center of the Sampling distribution for a mean: $\mu_{\overline{X}}$

When describing the **standard deviation** of a distribution we use the notation:

-   Population standard deviation: $\sigma_X$

-   Sample standard deviation: $s_X$

-   Spread of the sampling distribution is called the **Standard Error**.

    -   The standard error measures the variability in sample statistics due to randomness.
    -   We use the notation $\mbox{SE}(\overline{X}) = \sigma_{\overline{X}}$.

## Question 8: Shape of Sampling Distributions for a Mean

------------------------------------------------------------------------

In each of the three sampling distributions we examined, lets summarize what seems to be happening as the size of the samples, $n$, is increased.

Does the shape of the sampling distribution stay the same as the population or does it change as $n$ increases?

### Solution to Question 8:

------------------------------------------------------------------------

The shape of the sampling distribution does not stay the same as the population.
As the sample size becomes larger, the shape becomes more normal regardless of the shape of the population.

When n is small:

-   If the population is skewed, the shape is skewed.
    As n gets larger the shape is more normal.

-   If the population is symmetric, then the sampling distribution will be normal for small values of n and gets more and more normal as n gets larger

As long as n is large enough, the sampling distribution will become more normal.\

\
\

## Question 9: Center of Sampling Distributions for a Mean

------------------------------------------------------------------------

Does the center of the sampling distribution, $\mu_{\overline{X}}$, stay the same or change as $n$ increases?
How does the value of $\mu_{\overline{X}}$ compare to the population mean $\mu_X$?

### Solution to Question 9:

------------------------------------------------------------------------

Center stays the same (more or less) regardless of n.
The center of the sampling distribution is approximately equal to the population mean.\

\
\

## Question 10: Spread of Sampling Distributions for a Mean

------------------------------------------------------------------------

Does the standard error of the sampling distribution, $\mbox{SE}(\overline{X})$, stay the same or change as $n$ increases?

### Solution to Question 10:

------------------------------------------------------------------------

\
The standard error of the sampling distribution gets smaller and smaller as n gets larger and larger.\

\

# Formal Statement of the Central Limit Theorem (for Sample Means)

------------------------------------------------------------------------

Let $X_1, X_2, \ldots , X_n$ be independent, identically distributed (iid) random variables from a population with mean and standard deviation $\mu$ and $\sigma$, then as long as $n$ is large enough (informally $\mathbf{n \geq 30}$), the sampling distribution for the mean, $\bar{X}$ will:

-   Be (approximately) normally distribution.
-   Have mean equal to the mean of the population, $\mu$.
-   Have standard error $\mbox{SE}(\bar{X}) = \frac{\sigma}{\sqrt{n}}$.

We summarize the results more concisely below:

$$ \overline{X} \sim N \left( \mu_{\overline{X}} , \sigma_{\overline{X}} \right) = N \left( \mu  , \frac{\sigma}{\sqrt{n}} \right)$$

## Question 11:

------------------------------------------------------------------------

Using properties of expected value and variance of linear combinations of independent events, prove each of the following:

a.  $\displaystyle E \left( \bar{X} \right) = E \left( \frac{X_1 + X_2 + \ldots + X_n}{n} + \right) = \mu_X$

b.  $\displaystyle \mbox{Var} \left( \bar{X} \right) = \mbox{Var} \left( \frac{X_1 + X_2 + \ldots + X_n}{n} \right) = \frac{\sigma^2_X}{n}$

### Solution to Question 11a:

------------------------------------------------------------------------

Properties of expected value: $\bar{X} = \frac{\sum{X}}{n}$

$E \left( \bar{X} \right) = E \left( \frac{X_1 + X_2 + \ldots + X_n}{n} \right)$

$=\frac{1}{n} \  E \left( {X_1 + X_2 + \ldots + X_n}\right)$

$=\frac{1}{n} \  \left( {E(X_1) + E(X_2) + \ldots + E(X_n)}\right)$

$=\frac{1}{n} \  \left( {\mu_x + \mu_x + \ldots + \mu_x}\right)$

$=\frac{1}{n} \  n \cdot {\mu_x} => \mu_x$

\
\
\

### Solution to Question 11b:

------------------------------------------------------------------------

Properties of variance: $Var(aX + bY) = a^2Var(X) + b^2Var(Y)$

$\mbox{Var} \left( \frac{X_1 + X_2 + \ldots + X_n}{n} \right) = \frac{\sigma^2_X}{n}$

$\frac{1}{n^2} \ Var({X_1 + X_2 + \ldots + X_n})$

$=\frac{1}{n^2} \ {Var(X_1) + Var(X_2) + \ldots + Var(X_n)}$

$=\frac{1}{n^2} [n \cdot \sigma^2]$\
\
\

# Appendix: Some Sample Shiny Apps to Explore

------------------------------------------------------------------------

```{r, eval = FALSE}
runExample("01_hello")  # a histogram
#runExample("02_text")  # tables and data frames
#runExample("03_reactivity")  # a reactive expression
#runExample("04_mpg")  # global variables
#runExample("05_sliders")  # slider bars
#runExample("06_tabsets")  # tabbed panels
#runExample("07_widgets")  # help text and submit buttons
#runExample("08_html")  # Shiny app built from HTML
#runExample("09_upload")  # file upload wizard
#runExample("10_download")  # file download wizard
#runExample("11_timer")  # an automated timer
```
