---
title: '01: Introduction to Statistical Inference'
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

# Loading Libraries for This Session

------------------------------------------------------------------------

Have you installed `DataExplorer` and `dplyr`?
You will need them to run some of the code cells in this document.

-   If so great!
    -   But we will still need to load the package before we are able to access the libraries of functions and data in the package.
-   If you have not installed the packages, you will need to before running the code cell below.
    -   There should be a message on top of these window asking if you would like to install. Click that link!
-   Run the code cell below to load `DataExplorer` and `dplyr` packages.

```{r, message=FALSE}
# message=FALSE means output to console not displayed as output

library(DataExplorer)
library(dplyr)
```

-   Nothing appears on screen, but code is running in the console.
-   The `message=FALSE` option means the message received as output will not be displayed after running.
-   See <https://rmarkdown.rstudio.com/lesson-3.html> for a deeper explanation of code chunk options.

# Introduction to Statistical Inference

------------------------------------------------------------------------

[**Statistics**]{style="color: blue;"} is the study of collection, organization, analysis, interpretation, and presentation of data.

## Getting to Know Your Data

------------------------------------------------------------------------

The package `dplyr` contains a dataset called `storms`.
Let's check out the documentation for the data.

```{r}
?storms  # must load dplyr
```

-   Whenever you have a question about a dataset or function, type `?data_name` or `?function_name` to view help documentation.
-   For example, run the code cell below to learn about the `glimpse` function.

```{r}
?glimpse  # must load dplyr
```

### Getting a Glimpse of the Data

------------------------------------------------------------------------

```{r}
glimpse(storms)  # get a glimpse of storms data
```

## Question 1:

------------------------------------------------------------------------

What do you notice about the dataset `storms`?
How would you summarize what information is contained in this dataset?

### Solution to Question 1:

------------------------------------------------------------------------

\
\
\
\
\
\

## Question 2:

------------------------------------------------------------------------

What additional information would be nice to know about the `storms` dataset?

### Solution to Question 2:

------------------------------------------------------------------------

\
\
\
\
\
\

```{r set-options}
options(width = 100)  # sets width of output so summary displays

#summary(storms)  # summary of each variable in storms
#head(storms)  # prints first 6 rows to screen
#View(storms)  # opens new tab to display full dataset
#tail(storms)  # opens last 6 rows to screen
#plot_intro(storms)  # summary of data, requires DataExplorer
#plot_missing(storms)  # where is missing data, requires DataExplorer
```

# The Structure of Data

------------------------------------------------------------------------

[**Data frames**]{style="color: blue;"} are two-dimensional data objects and are the **fundamental** data structure used by most of R's libraries of functions and datasets.

-   Tabular data is [**tidy**]{style="color: blue;"} if each row corresponds to a different observation and column corresponds to a different variable.

Each column of a data frame is a [**variable**]{style="color: blue;"} (stored as a **vector**) of possibly different data types.
If the variable:

-   Is measured or counted by a number, it is a [**quantitative**]{style="color: blue;"} or [**numerical**]{style="color: blue;"} variable.
    -   Quantitative variables may be discrete (integers) or continuous (decimals).
-   Groups observations into different categories or rankings, it is a [**qualitative**]{style="color: blue;"} or [**categorical**]{style="color: blue;"} variable.

## Working with Categorical Data

------------------------------------------------------------------------

-   Sometimes we think a variable is one data type, but it is actually being stored (and thus interpreted by R) as a different data type.
-   One common issue is categorical data is stored as characters. We would like observations with the same values to be group together.

Categorical data should be stored as a [**factor**]{style="color: blue;"} in R.

```{r}
storms$status <- factor(storms$status)
summary(storms$status)
```

# Data Visualization

------------------------------------------------------------------------

For an overview of different types of plots, see the document **Quick Reference: Plotting Data in R** file posted in the **R Help** Module on Canvas (source R Markdown file also posted on Canvas).

The type analysis we can do depends on whether:

-   We are investigating a single variable, or looking for correlation between multiple variables.
-   The variable(s) are numerical and/or categorical.
-   The data satisfies certain assumptions.

```{r fig2, fig.height = 4, fig.width = 6}
par(mfrow = c(2, 2))  # Create a 2 x 2 array of plots

# The next 4 plots created will be arranged in the array
boxplot(storms$wind)  # create boxplot of wind speed

# Code below creates a histogram of wind speed
hist(storms$wind,
     col = "steelblue")  # change color of bars

plot(storms$status, 
     col = "gold")  # plots status, which is categorical

plot(wind ~ pressure, data = storms)  # plots two numerical variables
# ~ means comparing between two variables (x ~ y)

par(mfrow = c(1, 1))   # change settings so one image displayed in a window

# Compare numerical wind speed for different categories of storms
plot(wind ~ status, data = storms, col = "springgreen4")

#go to overview-of-plots file for list of plots in one document
```

-   **For one numerical variable:** histograms, boxplots, and density plots.
-   **For one categorical variable:** barplots and pie charts.
-   **For two numerical variables:** scatter plots.
-   **For one numerical and one categorical variables:** side-by-side boxplots or density plots.
-   **For two categorical variables:** grouped barplots.
-   **For three or more variables:** add distinguishing colors, shape/line types, and/or interactivity to plots.

# Statistical Inference

------------------------------------------------------------------------

-   A [**population**]{style="color: blue;"} includes all individuals or objects of interest.
-   A [**sample**]{style="color: blue;"} is a subset of the population.
-   [**Statistical inference**]{style="color: blue;"} is the process of drawing conclusions about the entire population based on information in a sample.
-   This semester we will **focus on inference**, and we will need some **probability** to do so.

![Diagram of Statistical Inference (see footnote to display image)[^1]](../Images/01-Inference.png){width="400"}

[^1]: The image above will not display for you by default.
    To see the image, go to [Diagram of Statistical Inference](https://drive.google.com/file/d/1wLs6EwNjoswH1rQk82H9I93N7g_m20hd/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.
    Image credit: <https://bolt.mph.ufl.edu/6050-6052/unit-4/>

[^2]: The image above will not display for you by default.
    To see the image, go to [Diagram of Statistical Inference](https://drive.google.com/file/d/1wLs6EwNjoswH1rQk82H9I93N7g_m20hd/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.
    Image credit: <https://bolt.mph.ufl.edu/6050-6052/unit-4/>

## Question 3:

------------------------------------------------------------------------

In the `storms` data example, is the data from a sample or a population?

### Solution to Question 3:

------------------------------------------------------------------------

sample\
\
\
\
\
\

## Question 4:

------------------------------------------------------------------------

What statistical questions might be worth investigating among the variables in the `storms` dataset?

### Solution to Question 4:

------------------------------------------------------------------------

Can location determine the occurrence of a storm?

-   answered w/ probabilities

What is the correlation b/w max wind speed and month?

-   are storms stronger in certain months?

How does pressure, wind speed, and/or diameter affect the status of the storm?\

\
\*\*This is what the final project is like: taking a data set, asking questions about it, and analyzing it\
\
\

# Collecting Data: Sampling

------------------------------------------------------------------------

Since drawing a sample that resembles the population in every way (except smaller in number) is critical for drawing valid conclusions, how we pick samples is sometimes the most important step.

![Diagram of Sampling Methods (see footnote to display image)[^3]](../Images/01-SampleMethods.png){width="450"}

[^3]: The image above will not display for you by default.
    To see the image, go to [Diagram of Sampling Methods](https://drive.google.com/file/d/1EpW3jTyO8c3YqjVcJ7T0Rz_Olq8uXZhG/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.
    Image credit: <https://towardsdatascience.com/8-types-of-sampling-techniques-b21adcdd2124> by Prakhar Mishra

[^4]: The image above will not display for you by default.
    To see the image, go to [Diagram of Sampling Methods](https://drive.google.com/file/d/1EpW3jTyO8c3YqjVcJ7T0Rz_Olq8uXZhG/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.
    Image credit: <https://towardsdatascience.com/8-types-of-sampling-techniques-b21adcdd2124> by Prakhar Mishra

## Summary of Sampling Methods

------------------------------------------------------------------------

-   When selecting a [**simple random sample**]{style="color: blue;"}, all individuals are equally likely to be selected.
-   When selecting a [**stratified sample**]{style="color: blue;"}, the population is subdivided into groups based on some meaningful characteristic.
-   When selecting a [**systematic sample**]{style="color: blue;"}, the first individual is chosen at random. Then a rule is used so that every $\mbox{n}^{\mbox{th}}$ individual is selected after that.
-   When selecting a [**cluster sample**]{style="color: blue;"} groups rather than individual units of the target population are selected at random for the test. For example, only people with last digit of phone number equal to 8 are chosen.
-   A [**convenience sample**]{style="color: red;"} is when people or elements in a sample are selected on the basis of their accessibility and availability.
-   [**Voluntary sampling**]{style="color: red;"} is a type of a convenience sample.

[**Sampling bias**]{style="color: blue;"} occurs when the method of selecting a sample causes the sample to differ from the population in some relevant way.

[**Randomly selecting samples is the best way to avoid bias!**]{style="color: blue;"}

# Collecting Data: Designing Studies

------------------------------------------------------------------------

Often in statistics we would like to investigate whether one variable is associated to another.
Researchers carry out studies to understand the conditions and causes of certain outcomes.

-   Does smoking cause lung cancer?
-   Is paying people or punishing people a more effective incentive to get vaccinated?
-   Is a new vaccine effective at preventing disease?

If we are using one variable to help us understand or predict the values (or category) of another variable, we call the first variable the [**explanatory or predictor variable**]{style="color: blue;"} and the second the [**response variable**]{style="color: blue;"}.

## Question 5:

------------------------------------------------------------------------

Both studies below are designed to examine determine whether rewarding good behavior or punishing bad behavior is a more effective method to help people quit smoking.
Which study do you believe is better designed?
Why?

### Study a:

------------------------------------------------------------------------

Employees at a large company voluntarily enroll in a quit smoking study.
When they join, they are provided two options to select from:

-   Option 1 (Reward-based group): If after six months the participant has quit smoking, they get an \$800 reward.
-   Option 2: (Deposit-based group): Pay an initial \$150 refundable deposit.

If after six months the participant:

-   Has quit smoking, they receive their \$150 deposit back plus an additional \$800 reward.
-   Has not quit smoking, then they do not receive their \$150 deposit back.

After six months, the success rate is compared between the two groups.

### Study b:

------------------------------------------------------------------------

Employees at a large company voluntarily enroll in a quit smoking study.

-   When they join, they are randomly assigned to either be in the Reward-based or Deposit-based group (same as described above).

After six months, the success rate is compared between the two groups.

### Solution to Question 5:

------------------------------------------------------------------------

I believe study 2 is slightly better designed.
In option 1, participants are never penalized, so they are rewarded either with quitting smoking and money or still smoking and no money.
In option 2, it provides 'incentive' for participants to work hard to get their deposit back.

study 2 may be better because it is random.

in study 1, if participants get to choose their treatment, there is bias

-   people that choose reward group may not be as motivated as those who choose the deposit group

-   the two options may result in unintentional categorical/characteristic splits (one group has younger people and another has older people)

\
\
\
\
\
\

## Confounding Variables

------------------------------------------------------------------------

A variable that is associated with both the explanatory variable and the response variable is called a [**confounding variable**]{style="color: blue;"}.

![Diagram of Confounding Variable (see footnote to display image)[^5]](../Images/01-Confounding.png){width="250"}

[^5]: The image above will not display for you by default.
    To see the image, go to [Diagram of Confounding Variable](https://drive.google.com/file/d/1pTd1yic3sBi0BnnHAhz2DwwmVH7V6CHT/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.

[^6]: The image above will not display for you by default.
    To see the image, go to [Diagram of Confounding Variable](https://drive.google.com/file/d/1pTd1yic3sBi0BnnHAhz2DwwmVH7V6CHT/view?usp=sharing).
    To display the image in your file, download the image file linked above and save the file in a folder named `Images` that is located one directory level back from the directory where you have saved the current file.

## Experiments and Observational Studies

------------------------------------------------------------------------

-   An [**observational study**]{style="color: blue;"} is a study in which the researcher does not actively control the value of any variable.
-   An [**experiment**]{style="color: blue;"} is a study in which the researcher actively controls one or more of the explanatory variables.
-   The different categories of the explanatory variable are called [**treatments**]{style="color: blue;"}.
-   In a [**randomized experiment**]{style="color: blue;"} the explanatory variable for each unit is determined randomly, before the response variable is measured.
-   If treatment groups are randomly determined, they should be similar in every way except for the treatment.
-   [There are almost always confounding variables in observational studies. Thus observational studies can almost never be used to establish causation.]{style="color: red;"}
