---
title: '03: Introduction to Probability'
author: 'Adam Spiegler, University of Colorado Denver'
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
# include=FALSE means code will not display in html
# eval=FALSE is a global option for blank worksheet
knitr::opts_chunk$set(eval=FALSE)
```

# Introduction

------------------------------------------------------------------------

A **statistical experiment or observation** is any random activity that results in a definite outcome.

-   The [**sample space**]{style="color: blue;"} $\Omega$ is the set of all possible outcomes of an experiment.
-   An [**outcome**]{style="color: blue;"} $\omega$, is a result from an experiment or observation.
-   An [**event**]{style="color: blue;"}, $A$, is a collection of one or more outcomes from an experiment or observation.

## Question 1

------------------------------------------------------------------------

In the 1970s, research[^1] by Alfred Sommer on vitamin A and night blindness suggested that vitamin A also reduced childhood death rates.
Children were randomly assigned to two groups, one group took a vitamin A pill every day, and children in the other group were given a **placebo** (a pill that had no vitamin A).
Below is a [**two-way table**]{style="color: blue;"} displaying the results of Sommer's research.

[^1]: [www.laskerfoundation.org/rprimers/sommer/saving1.htm](www.laskerfoundation.org/rprimers/sommer/saving1.htm) and *Introduction to the Practice of Statistics*, 4th ed, by David Moore, George.
    McCabe

![Two-Way Table of Vitamin A Study](https://lh5.googleusercontent.com/LQpBI0fgmumM0KAeNabaBPpOe6uXDSWrfhfN9LygJLkCCs-H1kKp1GCbv6_KHWPjNqM=w2400){width="423"}

-   a two way table describes the relation between 2 categorical variables

-   the explanatory variable is the treatment: in this study, it is taking vitamin A or not

-   the response variable is the survival, in this study it is whether the child lived or died

### Question 1 Part a:

------------------------------------------------------------------------

What is the probability that a randomly selected child in the study:

i.  Died?
ii. Received the treatment, vitamin A?
iii. Received vitamin A and died?
iv. Received vitamin A or died?
v.  Died given that they received vitamin A?

#### Solution to Question 1a

------------------------------------------------------------------------

Complete the code cell below to compute each of the probabilities (i)-(v).\

\

**Replace each ?? with an appropriate fraction.**

```{r}
# probability that a randomly selected child in the study:
part.i <- 231/25200  # died 
part.ii <- 12991/25200  # Received the treatment, vitamin A
part.iii <- 101/25200  # Received vitamin A and died
part.iv <- (12991+130)/25200  # Received vitamin A or died
part.v <- 101/12991  # Died given that they received vitamin A

# print results to screen
cat("answer to part i is", part.i,
      "\nanswer to part ii is", part.ii,
      "\nanswer to part iii is", part.iii,
      "\nanswer to part iv is", part.iv,
      "\nanswer to part v is", part.v)
```

\
\

Summarize results below

i.  The probability that a randomly selected child in the study died is $P(D)=$

ii. The probability that a randomly selected child in the study received the treatment, vitamin A, is $P(A)=??$

iii. The probability that a randomly selected child in the study received vitamin A and died is $P(A \cap D)=??$

iv. The probability that a randomly selected child in the study received vitamin A or died is $P(A \cup D)=??$

v.  The probability that a randomly selected child in the study died given that they received vitamin A is $P(D \ | \ A)=??$

\

### Question 1 Part b:

------------------------------------------------------------------------

Based on the data from this experiment, do you believe researchers can claim vitamin A decreases child mortality rates?
Which probabilities from part a did you use to draw this conclusion?

#### Solution to Question 1b:

------------------------------------------------------------------------

```{r}
130/12209
```

P(died given no vitamin a) = 0.011

P(died given vitamin a) = 0.0077

# Simple and Compound Probabilities

------------------------------------------------------------------------

Let $A$ and $B$ denote two events in sample space $\Omega$, then

-   $\color{blue}{P(A)}$ is the probability that event $A$ occurs.
-   $\color{blue}{P(A^C) = P(\bar{A}) = P(A')}$ is the probability that **event** $A$ does NOT occur.
    -   The notation $\color{blue}{A^C}$, $\color{blue}{\bar{A}}$, or $\color{blue}{A'}$ are used to denote the [**complement**]{style="color: blue;"} of $A$.
-   $\color{blue}{P(A \cap B)}$ is the probability that events $A$ **and** $B$ both occur.
-   $\color{blue}{P(A \cup B)}$ is the probability that either event $A$ **or** event $B$ occurs.
-   $\color{blue}{P(B \ | \ A )}$ is the [**conditional probability**]{style="color: blue;"} that event $B$ occurs [**given that**]{style="color: blue;"} event $A$ occurs.
-   $\color{blue}{P(A - B)}$ is the probability that event A occurs and event B does not occur.

## Question 2:

------------------------------------------------------------------------

Match one of the Venn diagrams labelled (i)-(vi) in the table below to one of the set operations below.
Note that two of the Venn Diagrams do not match any of the set operations.

\

| diagram (i)                                                                                                                                   | diagram (ii)                                                                                                                                   | diagram (iii)                                                                                                                                   |
|-----------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------|
| ![Venn Diagram (i)](https://lh5.googleusercontent.com/TUbLZRe21eo_5FSUTrfVDuP26E8Aq4xPNWwqjGwQF3akpsVK8EhXeFoKprNW00FDR7I=w2400){width="200"} | ![Venn Diagram (ii)](https://lh3.googleusercontent.com/hkkqAI9gm52KyvRx_pHNcbipeZL1q4xi_6ii46yp5c1yfBdPMODf0jrdUlYJUX81nrk=w3600){width="200"} | ![Venn Diagram (iii)](https://lh4.googleusercontent.com/qEmIplM6M83g7rAC6Ynmy2d_hxVR7PGltdmKhQ7r0ZveLPKOUemEpCo-JM1_Mcr7Dss=w2400){width="200"} |

| diagram (iv)                                                                                                                                   | diagram (v)                                                                                                                                   | diagram (vi)                                                                                                                                   |
|------------------------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------|
| ![Venn Diagram (iv)](https://lh6.googleusercontent.com/uxlcEFjv7wndArEu4Hz-oDgRihd3Xv1Q5RssinEL2vP0rvvVfpOaQN6a84s3qmlzu_8=w2400){width="200"} | ![Venn Diagram (v)](https://lh6.googleusercontent.com/4PO_EM9925B6Te3ym35GO0qu4gi-4_97-YF2XjfKCY78RS4ZHwN8gZLGj3Td_-4Gc90=w2400){width="200"} | ![Venn Diagram (vi)](https://lh4.googleusercontent.com/_7vPNG8VSuDeTskHwRrwqJHtqdjDm6mtTktK4PXAZQcWKjCD24oGLyI-RB5_wSa3bUU=w2400){width="200"} |

### Solution to Question 2:

------------------------------------------------------------------------

| $A^C$      | $A \cup B$ | $A \cap B$ | $A-B$       |
|------------|------------|------------|-------------|
| Diagram vi | Diagram i  | Diagram ii | Diagram iii |

\
\
\

# Probability Rules

------------------------------------------------------------------------

We can generalize the calculations from the previous case study on vitamin A and childhood morbidity to obtain the following results:

Let $A$ and $B$ denote two events in sample space $\Omega$, then

-   [**Additive rule**]{style="color: blue;"}: $P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
-   [**Bayes' Theorem**]{style="color: blue;"}: $\displaystyle P(B | A) = \frac{P(A \cap B)}{P(A)}$
-   [**Multiplicative rule**]{style="color: blue;"}: $P(A \cap B) = P(A) \cdot P(B | A)$
-   [**Complement rule**]{style="color: blue;"}: $P(A^C) = 1 - P(A)$

## Question 3:

------------------------------------------------------------------------

When a customer purchases a new car, they are presented with a menu of options such as heated steering wheel, parking assistant, satellite radio, etc.
The two most popular options on a certain type of new car are a sunroof (denoted $S$) and heated seats (denoted by $H$).
Answer the following questions if we know that

$$ P(S) = 0.6, \quad P(H) = 0.45,\mbox{ and} \quad P(H | S ) = 0.65. $$

### Question 3 Part a:

------------------------------------------------------------------------

Interpret the practical meaning of $P(H | S ) = 0.65$.

#### Solution to Question 3a:

------------------------------------------------------------------------

This describes the chance of getting a car with a heated seat given that the car has a sunroof.
The chance of getting a car like this is 65%.\

\
\
If a person chose a car with a sunroof, there is a 65% chance that they also got heated seats.\

\

### Question 3 Part b:

------------------------------------------------------------------------

Compute $P(H^C)$ and interpret the meaning.

#### Solution to Question 3b:

------------------------------------------------------------------------

1 - P(H) = 1 - 0.45 = 0.55\

This describes the chance that a chosen car does not have heated seats.
The chance of not having a heated seat car is 55%.\

\
55% of the cars do not have heated seats.\

### Question 3 Part c:

------------------------------------------------------------------------

Compute $P(S \cap H)$ and interpret the meaning.

#### Solution to Question 3c:

------------------------------------------------------------------------

= P(S) \* P(H \| S)

= 0.6 \* 0.65

= 0.39

This describes the chance that the chosen car will have a sunroof and heated seats.
The chance of getting both features is 39%.\

```{r}
0.6*0.65
```

### Question 3 Part d:

------------------------------------------------------------------------

Compute $P(S | H)$ and interpret the meaning.

#### Solution to Question 3d:

------------------------------------------------------------------------

Using Bayes theorem\
P(S \| H) = P(S and H) / P(H)

= 0.39 / 0.45

= 0.867

```{r}
0.39 / 0.45
```

# Independent Events

------------------------------------------------------------------------

Often in statistics we want to investigate questions such as:

-   Is a newly developed vaccine effective?
-   Do certain sentencing laws have an effect on crime rates?
-   Did increasing the minimum wage for fast food workers effect fast food prices?
-   **Does the occurrence of one event** $A$ (getting a sunroof) **effect the likelihood that another event** $B$ (heated seats) **occurs**?

## Question 4:

------------------------------------------------------------------------

In the car option example in question , we know that $P(S) = 0.6$, $P(H) = 0.45$, and $P(H | S ) = 0.65$.
Based on this information, **if a customer has purchased the sunroof option, are they more, less, or equally likely to get the heated seats option?** Explain how you determined your answer.

### Solution to Question 4:

------------------------------------------------------------------------

It seems that customers are more likely to get the heated seats if they purchased a car with a sunroof.
The probability of getting a heated seat given a sunroof is 65%, and the probability of not getting a heated seat given a sunroof is 35%\
\
\
\

## Definition of Independent Events

Two events $A$ and $B$ are [**independent**]{style="color: blue;"} if the occurrence of one has no effect on the occurrence of the other:

$$ P(B) = P(B \ | \ A) \quad \mbox{or} \quad P(A) = P(A \ | \ B).$$

**Special case:** If events $A$ and $B$ are independent events then we have $P(A \cap B) = P(A)P(B)$.

## Question 5:

------------------------------------------------------------------------

A person flips a fair coin and stops once they get at least one head and one tail.
What is the probability that it takes exactly four flips to get at least one head and one tail.

### Solution to Question 5:

------------------------------------------------------------------------

Outcomes in at most 4 flips:

HHHT

HHT(H/T)

HT(H/T)(H/T)

TTTH

TTH(H/T)

TH(H/T)(H/T)

Outcomes in exactly 4 flips:

HHHT

TTTH

P(TTTH) + P(HHHT) = 1/16 + 1/16 = 1/8\
when things are independent, can break up and probabilities into :

P(T)P(T)P(T)P(T)\
\
This is a disjoint event, events that cannot occur at the same time: P(A) + P(B)

Those that occur at the same time: P(A) + P(B) - P(intersection of A and B)\

# Disjoint Events

------------------------------------------------------------------------

Two events $A$ and $B$ are [**disjoint**]{style="color: blue;"} (or [**mutually exclusive**]{style="color: blue;"}) if they cannot occur at the same time, and therefore $P(A \cap B) = 0$.

**Special case:** If events $A$ and $B$ are disjoint then we have $P(A \cup B) = P(A)+P(B)$.

# Probability Distributions

------------------------------------------------------------------------

A function $P$ that assigns a real number $P(A)$ to each event $A$ is a [**probability distribution**]{style="color: blue;"} or a [**probability measure**]{style="color: blue;"} if it satisfies the following three axioms:

-   $P(A) \geq$ 0 for all $A$.

-   $P(\mbox{full sample space})=P(\Omega) = 1$.

-   If $A$ and $B$ are disjoint, then $$ P \left( A \cup B \right) = P(A) + P(B).$$

## Question 6:

------------------------------------------------------------------------

Complete the axioms in the definition above be replacing each ??
with an appropriate value or expression.

### Solution to Question 6:

------------------------------------------------------------------------

\
\
Replace each ??
in the axioms above.\

\

# OPTIONAL: Counting

------------------------------------------------------------------------

## Question 7:

------------------------------------------------------------------------

5 people have volunteered to work on a committee.
The committee will consist of a total of three people.
How many different committees of 3 people can be formed from the 5 volunteers?

### Solution to Question 7:

------------------------------------------------------------------------

\
\
\
\
\

## Different Ways of Choosing $k$ Items Out of $n$ Items

------------------------------------------------------------------------

We often need to count the number of ways of choosing $k$ items out of $n$ possible items.
You may recall this is often called [**n choose k**]{style="color: blue;"} and is denoted as

$$ \left( \begin{array}{c} n \\ k \end{array}\right) = \frac{n!}{k!(n-k)!}.$$

-   In R we can use the function `nchoosek(n, k)` which is in the `pracma` library.

```{r, message=FALSE}
library(pracma)
```

```{r}
# Calculate 6 choose 4
nchoosek(6, 4)
```

## Bonus Question 8:

------------------------------------------------------------------------

Suppose $n$ people are in a room.
What is the probability that there is at least one pair of people that have the same birthday?

*Hint: Let* $A$ be the event there is no match.
Calculate $P(A)$, and then find $P(A^C)$.

# OPTIONAL: Bayes' Theorem

------------------------------------------------------------------------

## Partitions of Sample Spaces

------------------------------------------------------------------------

A [**partition**]{style="color: blue;"} of a space $\Omega$ is a collection of disjoint sets such that $\displaystyle \bigcup_{i=1}^{\infty} A_i = \Omega$.

## Bonus Question 9:

------------------------------------------------------------------------

Fill in the blank to complete theorem below and explain (in words, pictures, or equations) how you determined your answer.

![Total Probability Venn Diagram](https://lh5.googleusercontent.com/ayOHh6EOjqG84K8VVfx_DHI6OsSRBxQKzf0ACciT__CgQPJh3FWq-vwc0oVB_RP_e8A=w2400){width="400"}

Let $A_1$, $A_2$, $\ldots , A_k$ be a partition of $\Omega$.
Then for any event $B$, $$ P(B) = \sum_{i=1}^k ?? $$.

## Statement of Bayes' Theorem

------------------------------------------------------------------------

Let $A_1$, $A_2$, $\ldots , A_k$ be a partition of $\Omega$ such that $P(A_i)>0$ for each $i$.
If $B$ is any event with $P(B)>0$, then for each $i=1, \ldots k$, we have Then for any event $B$,

$$ P(A_i \mid B) = \frac{P(A_i \cap B) }{P(B)} = \frac{P(B \mid A_i) P(A_i)}{\sum_{j=1}^k P(B \mid A_j)P(A_j)}.$$

## Bonus Question 10:

------------------------------------------------------------------------

Suppose that 30% of computers run Mac, 50% use PC, and 20% use Linux.
A computer virus is created by hackers, and suppose that 65% of Mac, 82% of PC, and 50% of Linux computers get the virus.

### Bonus Question 10 Part a:

------------------------------------------------------------------------

What is the probability that a randomly selected computer has the virus?

#### Solution to Question 10a:

------------------------------------------------------------------------

\
\
\
\
\

### Bonus Question 10 Part b:

------------------------------------------------------------------------

What is the probability that a randomly selected computer is PC given that the computer is infected by the virus?

#### Solution to Question 10b:

------------------------------------------------------------------------

\
\
\
\
\
