---
title: '04: Introduction to Random Variables'
author: 'Adam Spiegler, University of Colorado Denver'
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
# include=FALSE means code will not display in html
# eval=FALSE is a global option for blank worksheet
knitr::opts_chunk$set(eval=FALSE)
```

# Random Variables

------------------------------------------------------------------------

Suppose a company knows that 1% of the lithium batteries they manufacture are defective.
The manufacturer sends a shipment of 10 batteries to one of their clients.
How likely is it that none of the batteries are defective?
How can we connect the data to the concepts of sample spaces and events to answer this question?
This link can be made using [**random variables**]{style="color: blue;"}:

A [**random variable**]{style="color: blue;"} is a mapping

$$ X: \Omega \to \mathbb{R} $$

that assigns a real number $X(\omega)$ to each outcome $\omega \in \Omega$.

## Example:

------------------------------------------------------------------------

For example, the random variable $X$ could map an outcome in which 10 batteries are selected at random to a number corresponding to how many batteries in the sample are not defective, denoted $G$ for good.
For instance if $\omega = DDGGGGGGGG$, then $X(DDGGGGGGGG)=8$.
We can calculate the probability that $X=10$ (none are defective) using independence of events: $$ P(X=10) = P(G)P(G) \cdots P(G) = (P(G))^{10} = (0.99)^{10} \approx 0.9044.$$

## Question 1:

------------------------------------------------------------------------

A student takes a 3 question quiz.
Each of the 3 questions is multiple choice with 4 possible answer choices.
Let the random variable $X$ be the number of correct guesses out of the 3 questions.

### Question 1 Part a:

------------------------------------------------------------------------

What is the sample space $\Omega$?

#### Solution to Question 1a:

------------------------------------------------------------------------

Sample Space (assuming each choice is independent):

CCC

CCW

CWC

WCC\
WWC

WCW

CWW

WWW\

8 total outcomes, 2 choices for each question, 3 questions total\
\
\
\

### Question 1 Part b:

------------------------------------------------------------------------

Compute $P(X=0)$, how likely is it that the student gets none of the questions correct?

#### Solution to Question 1b:

------------------------------------------------------------------------

\
Assuming for each question, there is only 1 right answer.
So there is a 25% of getting a correct answer and 75% of getting a wrong answer for one question.

The possible outcomes for getting no correct answers is 1, which is WWW.

The probability of getting a wrong answer for each question is:

(0.75) \* (0.75) \* (0.75) \* 1

```{r}
0.75^3
```

P(X = 0) = 1/8 assumes that the events are equally likely to happen (but in this case it is not)

### Question 1 Part c:

------------------------------------------------------------------------

Fill in the empty entries in table below

#### Solution to Question 1c:

------------------------------------------------------------------------

\

| $x$      | 0        | 1        | 2        | 3        |
|----------|----------|----------|----------|----------|
| $P(X=x)$ | 0.421875 | 0.421875 | 0.140625 | 0.015625 |

```{r}
0.25^3 #all correct answers
3 * (0.25 * 0.75**2) #one correct answer
3 * (0.25**2 * 0.75) #two correct answers
0.75^3 #no correct answers
```

### Question 1 Part d:

------------------------------------------------------------------------

Compute $P(X \leq 1)$ and interpret the practical meaning of this value.

#### Solution to Question 1d:

------------------------------------------------------------------------

\
P(X = 0) + P(X = 1) = 0.84

```{r}
0.421875 + 0.421875
```

# Distributions for Discrete Random Variables

------------------------------------------------------------------------

-   If $X$ is a [**discrete**]{style="color: blue;"} random variable, we define the probability function or [**probability distribution**]{style="color: blue;"} or [**probability mass function (pmf)**]{style="color: blue;"} for $X$ by

$$\color{blue}{p(x) = P(X=x)}.$$

-   If $X$ is a random variable, we define the [**cumulative distribution function (cdf)**]{style="color: blue;"} as the function

$$\color{blue}{F(x)=P(X \leq x) = \sum_{k=\rm{min \ value}}^x p(k) }.$$

## Question 2:

------------------------------------------------------------------------

**On a separate piece of paper,** sketch the **pmf** and **cdf** for the random variable $X$ (number of correct guesses on the 3 question quiz) in Question 1.
Be sure to label the tickmarks on the vertical axes with an appropriate scale.

![Blank Axes for pmf](https://lh5.googleusercontent.com/vbxPA1pFqeuCN8dEYJU9a-DiiJ3C5vS15Cfs5Yv3LM9IRVp0TWo-haiVUKBtciYp6co=w2400){width="300"}             ![Blank Axes for cdf](https://lh6.googleusercontent.com/XlipSscMkNVAR6RSWiwtste1VfcklAtQPAxJgdWfYESkkaPHui2brC9aiaZRJHfYqGk=w2400){width="300"}

### Solution to Question 2:

------------------------------------------------------------------------

\
Sketch your graph on a separate piece of paper.\

```{r}
n <- 3  # set number of trials n
p <- 0.25  # prob of success in a trial
x <- 0:n  # vector from 0 to n=3

par(mfrow=c(1, 2))
plot(x, dbinom(x, size = n, prob = p), main = "Graph of pmf")
plot(x, pbinom(x, size = n, prob = p), type="s", main = "Graph of cdf")
par(mfrow=c(1, 1))
```

## Question 3:

------------------------------------------------------------------------

Let $X$ is a discrete random variable with pmf and cdf denoted $p(x)$ and $F(x)$, respectively.
Determine if each statement is True or False.

a.  $0 \leq p(x) \leq 1$ for all $x$.
b.  $0 \leq F(x) \leq 1$ for all $x$.
c.  $\displaystyle \sum_{\rm{all}\  x} p(x) = 1$.
d.  $\displaystyle \sum_{\rm{all}\  x} F(x) = 1$.
e.  $\displaystyle \lim_{x \to \infty} p(x) = 1$.
f.  $\displaystyle \lim_{x \to \infty} F(x) = 1$.
g.  The pmf must be a nondecreasing function.
h.  The cdf must be a nondecreasing function.

### Solution to Question 3:

------------------------------------------------------------------------

List the properties (a)-(h) that are indeed TRUE.

a\.
True\
b.
True

c\.
True: sum of probabilities all possible outcomes add up to 1 

d\.
False (see above example) 

e\.
False (see above example) 

f\.
True: see part b as supplement reasoning 

g\.
False (see example above) 

h\.
True\

# Expected Value and Variance

------------------------------------------------------------------------

-   The average or [**expected value**]{style="color: blue;"} for a discrete random variable $X$ is denote $E(X)$ or $\mu$ and computed using the formula

$$ E(X) = \mu =  \sum_x \left( x \cdot p(x) \right) = \sum_x \left( x \cdot P(X=x) \right). $$

example from problem about batteries:

E(X) = 0(0.42) + 1(0.42) + 2(0.14) + 3(0.015)

-   The [**variance**]{style="color: blue;"} for a discrete random variable $X$ is one common way to measure how spread out (in relation to the expected value) are the values of $X$. The variance is denoted $\mbox{Var}(X)$ or $\sigma^2$ and computed using the formula

$$ \mbox{Var}(X) = \sigma^2 =  \sum_x \left( (x-\mu)^2 \cdot p(x) \right) = \sum_x \left( (x-\mu)^2 \cdot P(X=x) \right). $$

-   The [**standard deviation**]{style="color: blue;"} for a discrete random variable $X$ is the square root of the variance and is denoted by $\sigma$. It more or less measures the average of the distances for each value of $X$ from the mean $\mu$.

$$ \mbox{SD}(X) = \sigma =  \sqrt{\mbox{Var}(X)}. $$

## Question 4:

------------------------------------------------------------------------

Using properties of the pmf, $p(x)$, show that

$$\mbox{Var}(X) = E(X^2)  - \mu^2.$$ \### Solution to Question 4:

------------------------------------------------------------------------

**Finish the proof below!**

$$\begin{array}{rcll}
\mbox{Var}(X) &=&  \sum_x \left( (x-\mu)^2 \cdot p(x) \right) & \mbox{(by definition)}\\
 &=& \sum_x \left( (x^2 -2 \mu x + \mu^2) \cdot p(x) \right) & \mbox{(distributing)}\\
 &=& \sum_x \left( x^2 \cdot p(x) -2 \mu x  \cdot p(x)  + \mu^2  \cdot p(x)  \right) & \mbox{(distributing again)}\\
 &=& \sum_x \left( x^2 \cdot p(x) \right) - \sum_x \left(2 \mu x  \cdot p(x) \right)  + \sum_x \left( \mu^2  \cdot p(x)  \right) & \mbox{(properties of summation)}\\
 &=& \sum_x \left( x^2 \cdot p(x) \right) - 2\mu\sum_x \left(x  \cdot p(x) \right)  + \mu^2\sum_x \left(p(x)  \right)  & \mbox{(factor)} \\
 &=& \sum_x \left( x^2 \cdot p(x) \right) - 2\mu\sum_x \left(x  \cdot p(x) \right)  + \mu^2(1) & \mbox{(property of pmf)} \\
 &=& \sum_x \left( x^2 \cdot p(x) \right) - 2\mu \cdot \mu  + \mu^2(1) & \mbox{(definition of E(X))} \\
 &=& E(X^2) - 2\mu^2  + \mu^2 & \mbox{(definition of E(X^2))} \\
 &=& E(X^2) - \mu^2 & \mbox{(combine like terms)} \\
\end{array}$$

\
\

## Question 5:

------------------------------------------------------------------------

A charity is running a raffle as a fundraiser.
They offer one grand prize of \$500, two second prizes of \$$100$ each, and ten third prizes of \$20 each.
They plan to sell 1,000 tickets each at a price of \$2.
Let $X$ denote the amount of money won by a lottery ticket

### Question 5 Part a:

------------------------------------------------------------------------

Fill in the values of $x$ and $p(x)$ in the table below.

#### Solution to Question 5a:

------------------------------------------------------------------------

**Fill in the blanks to complete the table.**

| $x$    | 0        | 20      | 100    | 500    |
|--------|----------|---------|--------|--------|
| $p(x)$ | 987/1000 | 10/1000 | 2/1000 | 1/1000 |

\
1) find pmf table

2)  calculate E(X)\

### Question 5 Part b:

------------------------------------------------------------------------

Calculate $E(X)$ and $\mbox{Var}(X)$.

#### Solution to Question 5b:

------------------------------------------------------------------------

\
$E(X) = 0(0.987) + 20(0.01) + 100(0.002) + 500(0.001) = 0.9$

-this says, if you bought 1000 tickets, you would get 90 cents on average.\

$$
Var(X) = E(X^2) - (E(X))^2 = ((0)^2)(0.987) + (20)^2(0.01) + (100)^2(0.002) + (500)^2(0.001) - (0.90)^2 = 273.19
$$\
\
\
