---
title: '05: Discrete Random Variables'
author: "Adam Spiegler, University of Colorado Denver"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: sentence
---

# Introduction

------------------------------------------------------------------------

Suppose a company knows that 1% of the lithium batteries they manufacture are defective.
The manufacturer sends a shipment of 10 batteries to one of their clients.
Let the random variable $X$ denote the number of good (not defective) batteries in the shipment.
How likely is it that exactly two of the batteries are defective?

## Question 1:

------------------------------------------------------------------------

What is the probability of getting the outcome $GGGGGGGGDD$?
Note $G$ and $D$ denote good and defective batteries, respectively.

### Solution to 1:

------------------------------------------------------------------------

0.01 are defective

(0.99)\^8 \* (0.01)\*2

```{r}
#probability of one combination of 8 good batteries and 2 bad batteries are defective
(0.99)**8 * (0.01)**2
```

## Question 2:

------------------------------------------------------------------------

How many outcomes are in the event exactly 2 out of 10 batteries are defective?

### Solution to 2:

------------------------------------------------------------------------

All possible outcomes of 8 good batteries and 2 defective ones:

n choose k: combinatorics, check counting section in worksheet 3

getting all the permutations:

$$
10! \ / \ (8! \ * \ 2!)
$$

```{r}
library(pracma)
nchoosek(10, 2)
```

## Question 3:

------------------------------------------------------------------------

Calculate the probability that exactly 2 are defective.

### Solution to 3:

------------------------------------------------------------------------

P(X = 8) = 45(9.227447e-05)

```{r}
45 * 9.227447e-05
```

## Question 4:

------------------------------------------------------------------------

Generalize the result in the previous example.
If you repeat $n$ trials which are independent from one another, and each has the same probability of success, $p$, what is the probability of getting exactly $s$ successes out of $n$ trials?

### Solution to 4:

------------------------------------------------------------------------

$$f(s) = P(X=s) = \left( \begin{array}{c} 
n \\ s
\end{array} \right) p^{s} (1-p)^{n-s}$$

\
\
If we have n batteries shipped and probability that a battery is good is p:

P(X = s) = probability that exactly s batteries out of n are good.

P(X = s) = number of outcomes with s successes \* probability that we have one outcome with s successes\

## Question 5:

------------------------------------------------------------------------

How many good batteries would you expect to receive if you had a shipment of 100 batteries (where it is known that $1$% of all batteries are defective)?
If you received a shipment of 200 batteries?
A shipment of 10 batteries?

### Solution to 5:

------------------------------------------------------------------------

defective batteries = 0.01

n = 100, expected value = 99

n = 200, expected value = 198

n = 10, expected value = 9.9\

$E(X) = np = sum(p(x)*x)$\
\
The formula E(X) = np only works when it is a binomial distribution

## Question 6:

------------------------------------------------------------------------

Write a general formula for the expected number of successes when $n$ trials are repeated with probability of success $p$ in each trial.

### Solution to 6:

------------------------------------------------------------------------

\
E(X) = np\
\
\

# Trials with Two Possible Outcomes

------------------------------------------------------------------------

## Bernoulli Trials

------------------------------------------------------------------------

A [**Bernoulli trial**]{style="color: blue;"} is an experiment that has [**exactly two possible outcomes**]{style="color: blue;"}:

-   The probability that the outcome of a trial is a [success]{style="color: blue;"} ($\color{blue}{X=1}$) is denoted $\color{blue}{p}$.
-   Otherwise, the probability of a [failure]{style="color: red;"} ($\color{red}{X=0}$) is $\color{red}{q=1-p}$.
-   $X$ has a [**Bernoulli Distribution**]{style="color: blue;"} with probability mass function

$$f(x) = \left\{ \begin{array}{ll}
p^x(1-p)^{1-x} & \mbox{for } x \in \left\{ 0, 1 \right \} \\
0 , & \mbox{otherwise}
\end{array} \right.$$

## The Binomial Distribution

------------------------------------------------------------------------

Let random variable $X$ be the number of successes out of $n$ trials, where [**each trial is identical and independent**]{style="color: blue;"}.

-   $X$ has a [**Binomial Distribution**]{style="color: blue;"}, written $X \sim \mbox{Binomial}(n,p)$.
-   The probability mass function is

$$f(x) = \left\{ \begin{array}{ll}
\left( \begin{array}{c} n\\ x \end{array} \right) p^x(1-p)^{n-x} & \mbox{for } x =0,1,2, \ldots , n \\
0 & \mbox{otherwise} 
\end{array} \right. $$

**LaTeX Notes:**

-   The `array` environment is started with `\begin{array}`
-   Then we indicate how many columns by how each each is aligned.
    -   `{ll}` means two columns aligned to the left.
    -   `{lrc}` would be three columns the first aligned left, then right, then the last is centered,
    -   We use the `&` symbol to indicate a column break.
    -   We use `\\` to indicate a row break.
-   If we want a big curly brace on the left of the array, enter `\left\{` before beginning the array.
-   We do not want a brace on the other side of the array, so we right `\right.` to close off the left brace without using any symbol on the right.
-   We use `\mbox{otherwise}` to type the text *otherwise* inside the equation.

### Expected Value and Variance of Binomial Distributions

------------------------------------------------------------------------

-   The **expected value** can be calculated with the shortcut $\color{blue}{E(X) = np}$.
-   The **variance** can be calculated with the shortcut $\color{blue}{\mbox{Var}(X) = npq}$.

### Working with Binomial Distributions in R

------------------------------------------------------------------------

In R, the we can use the functions:

-   `dbinom(x, n, p)` calculates the probability of [exactly $x$ success]{style="color: blue;"} out $n$ trials, $\color{blue}{P(X=x)}$.
-   `pbinom(x, n, p)` calculates the probability of [at most $x$ success]{style="color: red;"} out $n$ trials, $\color{red}{P(X \leq x)}$.
-   `rbinom(m, n, p)` randomly sample $m$ values from $X \sim \mbox{Binom}(\mbox{size}, p)$ (with replacement).
-   `qbinom(q, n, p)` compute the qth quantile from $X \sim \mbox{Binom}(n, p)$.

```{r}
par(mfrow=c(1,2))  # plot in a 1 by 2 array
x <- 0:40  # vector of values 1, 2, ..., 20

# plot pmf n=40 and p = 0.25
plot(x, dbinom(x, size=40, prob=0.25), 
     type='p')  # point plot

# plot cdf n=40 and p = 0.25
plot(x ,pbinom(x, size=40, prob=0.25),
     type='s',  # stairs plot
     main='X ~ Binom(40, 0.25)',
     ylab='Probability',
     xlab ='At most x Successes',
     lwd=3)

par(mfrow=c(1,1)) # return to single plot
```

## Question 7:

------------------------------------------------------------------------

Consider the lithium battery example from [Question 1:] .
Write a command in R that computes the probability of each event.

### Question 7a:

------------------------------------------------------------------------

Exactly 7 batteries are good.

#### Solution to Question 7a:

------------------------------------------------------------------------

Feel free to insert a code cell or not!

```{r}
dbinom(7,10,0.99)
```

### Question 7b:

------------------------------------------------------------------------

At most 8 batteries are good.

#### Solution to Question 7b:

------------------------------------------------------------------------

Feel free to insert a code cell or not!

```{r}
pbinom(8, 10, 0.99)
```

\
\
\

### Question 7c:

------------------------------------------------------------------------

At most 4 batteries are defective.

At least 6 batteries are good.
$P(X \geq 6) = P(X \leq 5)$

#### Solution to Question 7c:

------------------------------------------------------------------------

Feel free to insert a code cell or not!

```{r}
pbinom(4, 10, 0.01) #at most 4 batteries are defective
1-pbinom(5, 10, 0.99) #1 - at most 5 batteries are good = at least 6 batteries are good
```

# $n$ Equally Likely Outcomes

------------------------------------------------------------------------

## Question 8:

------------------------------------------------------------------------

Let $X$ be the value of the result of rolling a fair six-sided die.

### Question 8a:

------------------------------------------------------------------------

Write out the probability mass function $f(x)$.

#### Solution to Question 8a:

------------------------------------------------------------------------

\
\

$$f(x) = \left\{ \begin{array}{ll} 
\frac{1}{6} & x = 1, 2, 3, 4, 5, 6 \\
0 & \mbox{otherwise} 
\end{array} \right.$$

\
6 outcomes each equally likely.\

\

### Question 8b:

------------------------------------------------------------------------

What is the expected value of rolling a fair-six sided die?

#### Solution to Question 8a:

------------------------------------------------------------------------

$$
E(X) = \frac{1}{6}(1+2+3+4+5+6)
= 7/2
$$\
\
\
\
\

## Uniform Distributions

------------------------------------------------------------------------

Let $X$ be a discrete random variable with $k$ different outcomes that each have the same likelihood of occurring.

-   $X$ has a [**uniform distribution**]{style="color: blue;"} on $\left\{ 1, 2, 3, \ldots , k \right\}$
-   The probability mass function is $$ f(x) = \left\{ \begin{array}{ll} \frac{1}{k} \ \ & \mbox{for } x = 1, 2, \ldots,  k\\
    0 \ \ & \mbox{otherwise} \end{array} \right. .$$

```{r}
par(mfrow=c(1,2))  # plot in a 1 by 2 array

x.min <- 1
x.max <- 6
x <- x.min:x.max  # vector of integers  x.min,  ..., x.max
p.x <- rep(1/(x.max - x.min+1), length(x) ) 
c.x <- p.x * (x)

plot(x, p.x, 
     type='p',  # point plot
     main='PMF')  

plot(x, c.x, 
     type='s',  # stairs plot
     main='CDF') 

par(mfrow=c(1,1)) # return to single plot
```

### Expected Value and Variance of Uniform Distributions

------------------------------------------------------------------------

-   The expected value is $E(X) = \frac{k+1}{2}$.
-   The variance is $\mbox{Var}(X) = \frac{k^2-1}{12}$.

## Question 9:

------------------------------------------------------------------------

A sports marketer for the Denver Nuggets randomly calls people in the Denver area until she encounters someone who attended a Nuggets' game.
What is the probability the market encounters $7$ people who did not attend a game before the first success when it is known that 10% of the population attended a game last season?

### Solution to Question 9:

------------------------------------------------------------------------

\
outcome: NNNNNNNY

$P(N)^7 * P(Y)^1$\
\
(0.9)^7^ \* (0.1)^1^\

```{r}
0.9**7 * 0.1
```

# Geometric Distributions: Number of Failures Before First Success

------------------------------------------------------------------------

If we repeat a Bernoulli trial that has probability of success $p$ for each trial, we can [count the number of failures, $X$, that occur before the first success.]{style="color: blue;"}

## Formula and Plots of Geometric Distributions

------------------------------------------------------------------------

-   $X$ has a [**Geometric Distribution**]{style="color: blue;"}, written $X \sim \mbox{Geom}(p)$.
-   The probability mass function is

$$f(x) = q^xp \quad \mbox{for } x \in \left\{ 0, 1, \ldots \right\}$$.

```{r}
par(mfrow=c(1,2))  # plot in a 1 by 2 array

x <- 0:10 

plot(x, dgeom(x, prob=0.75), 
     type='p',  # point plot
     main='PMF')  

plot(x, pgeom(x, prob=0.75),
     type='s',  # stairs plot
     main='CDF') 

par(mfrow=c(1,1)) # return to single plot
```

### Expected Value and Variance of Geometric Distributions

------------------------------------------------------------------------

-   The expected value is $\mu=E(X) = \frac{q}{p}$.
-   The variance is $\sigma^2 = \mbox{Var}(X) = \frac{q}{p^2}$.

### R Functions for Geometric Functions

------------------------------------------------------------------------

In R, the we can use the functions:

-   `dgeom(x, p)` calculates the probability of [exactly $x$ failures]{style="color: blue;"} before first success.
-   `pgeom(x, p)` calculates the probability of [at most $x$ failures]{style="color: red;"} before first success.
-   `rgeom(n, p)` randomly pick $n$ values from $X \sim \mbox{Geom}(p)$ (with replacement).
-   `qgeom(q, p)` compute the qth quantile from $X \sim \mbox{Geom}(p)$.
-   There is no fixed number of trials $n$.

# Number of Times an Event Occurs in a Fixed Time Period

------------------------------------------------------------------------

## Poisson Distribution

------------------------------------------------------------------------

The [**Poisson distribution**]{style="color: blue;"} applies when the average frequency of occurrences in a given time period is known, and each occurrence is independent the others.
Let $X$ denote [the number of occurrences in the given time period]{style="color: blue;"}.

-   $X \sim \mbox{Poisson}(\lambda)$, where $\lambda$ denotes the mean number of occurrences in the given time.
-   The probability mass function is

$$f(x) = e^{-\lambda} \frac{\lambda^x}{x!} \quad \mbox{for } x \in \lbrace 0, 1, 2, \ldots \rbrace .$$

```{r}
par(mfrow=c(1,2))  # plot in a 1 by 2 array

x <- 0:15  # creates vector of possible occurrences
lam <- 5  # on average 5 occurrence per time period

plot(x, dpois(x, lam), 
     type='p',  # point plot
     main='PMF')  

plot(x, ppois(x, lam),
     type='s',  # stairs plot
     main='CDF') 

par(mfrow=c(1,1)) # return to single plot
```

### Expected Value and Variance of Poisson Distributions

------------------------------------------------------------------------

-   The expected value is $\mu = E(X) = \lambda$.
-   The variance is $\sigma^2 = \mbox{Var}(X) = \lambda$ (same as the expected value).

### R Functions for Geometric Functions

------------------------------------------------------------------------

In R, the we can use the functions:

In R, the we can use the functions:

-   `dpois(x, lambda)` calculates the probability of [exactly $x$ occurrences]{style="color: blue;"} in the given time.
-   `ppois(x, lambda)` calculates the probability of [at most $x$ occurrences]{style="color: red;"} in the given time.
-   `rpois(n, lambda)` randomly pick $n$ values from $X \sim \mbox{Poisson}(\lambda)$ (with replacement).
-   `qpois(q, lambda)` compute the qth quantile from $X \sim \mbox{Poisson}(\lambda)$.

## Question 10:

------------------------------------------------------------------------

If there are twelve cars crossing a bridge per minute on average, find the probability of having seventeen or more cars crossing the bridge in a particular minute.

### Solution to Question 10:

------------------------------------------------------------------------

Feel free to insert a code cell or not!

```{r}
ppois(17, 12) #x >= 17, lamda = 12 cars per minute avg
```

# Practice Problems

------------------------------------------------------------------------

For each situation, identify which distribution best describes the distribution of the random variable.
Then use a probability distribution function to calculate the probability.

## Practice Problem 1:

------------------------------------------------------------------------

A online retailer sells an average of 5 big screen TV's on a given day.
What is the probability they sell 9 TV's in a day?

### Solution to Practice Problem 1:

------------------------------------------------------------------------

\
\

Feel free to insert a code cell or not!

\
\

## Practice Problem 2:

------------------------------------------------------------------------

It is known that 3% of airbags manufactured by a certain car company are defective.
What is the probability that the first defective air bag occurs when the fifth item is inspected?

### Solution to Practice Problem 2:

------------------------------------------------------------------------

\
\

Feel free to insert a code cell or not!

\
\

## Practice Problem 3:

------------------------------------------------------------------------

Recently, a nurse commented that when a patient calls the medical advice line claiming to have the flu, the chance that he or she truly has the flu (and not just a nasty cold) is only about 4%.
Of the next 25 patients calling in claiming to have the flu, what is the probability that exactly 4 patients will have the flu?

### Solution to Practice Problem 3:

------------------------------------------------------------------------

\
\

Feel free to insert a code cell or not!

\
\
