---
title: '06: Introduction to Continuous Random Variables'
author: "Adam Spiegler, University of Colorado Denver"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Revisiting Our Definition of a Random Variable

A manufacturer of lithium batteries measures the weight of each box they ship out to customers.
Let $X$ denote the weight (in pounds) of a randomly selected shipment.
It is possible that $X=8$ or $X=9$, but the weight could potentially be any value (greater than 0) such as $8.3671$ pounds if they want to be really precise.
Recall the previous definition of a random variable below.

A [**random variable**]{style="color: blue;"} is a mapping

$$X: \Omega \to \mathbb{R}$$

that assigns a real number $X(\omega)$ to each outcome $\omega \in \Omega$.

-   With a [**discrete random variable**]{sytle="color: blue;"}, the sample space is mapped to the **integers** (or a subset of the integers).
-   With a [**continuous random variable**]{sytle="color: blue;"}, the sample space is mapped to **an interval of values** in $\mathbb{R}$.

## Question 1:

------------------------------------------------------------------------

Imagine the GPA distribution for all students at a university follows a [**uniform distribution**]{style="color: blue;"} with GPA's of 0 and 4 being the smallest and largest GPA's possible.

### Question 1a:

------------------------------------------------------------------------

Write a formula for probability density function $f_X(x)$ for the uniform distribution.

#### Solution to Question 1a:

------------------------------------------------------------------------

\
f(x) = l x h (uniform distribution)\
\
\

### Question 1a:

------------------------------------------------------------------------

What proportion of students earned a GPA between $3$ and $3.5$?

#### Solution to Question 1a:

------------------------------------------------------------------------

\
l = 4 (because the range is from 0-4)

h = 1/4 because the probability/area under the curve should equal 1 (100%)

0.25 \* 0.5

```{r}
.25 * .5
```

```{r, echo=FALSE}
par(mfrow=c(1,2))

x <- 0:4
uni <- c(0.25, 0.25, 0.25, 0.25, 0.25)

plot(x, uni, 
     type="l", 
     col="red", 
     lwd=3, 
     xlim=c(0,4),
     ylim=c(0,0.3),
     yaxp  = c(0, 0.375, 3),
     xlab="GPA", 
     ylab="f(x)", 
     main="Uniform GPA Distribution")
abline(h =c(0, 0.125, 0.25), 
       v = c(0,1,2,3,4), 
       col=c("black", "grey", "grey","black",  "grey", "grey",  "grey", "grey"), 
       lwd=1, 
       lty=c(1,2,2,1,2,2,2,2)) 

x1 <- 0:2
x2 <- 2:4
y1 <- 0.25*x1
y2 <- -0.25*x2 + 1

plot(c(x1, x2), c(y1, y2),
     type="l", 
     col="red", 
     lwd=3, 
     xlim=c(0,4),
     ylim=c(0,0.5),
     yaxp  = c(0, 0.5, 4),
     xlab="GPA", 
     ylab="f(x)", 
     main="Other GPA Distribution")
abline(h =c(0, 0.125, 0.25, 0.375, 0.5), 
       v = c(0,1,2,3,4), 
       col=c("black", "grey", "grey", "grey", "grey"), 
       lwd=1, 
       lty=c(1,2,2,2,2)) 

par(mfrow=c(1,1))
```

## Question 2:

------------------------------------------------------------------------

The graph of $f_X(x)$ on the right above shows the probability density function for the GPA's of all students at a university.

### Question 2a:

------------------------------------------------------------------------

What proportion of students at the university have a GPA between 3 and 4?

#### Solution to Question 2a:

------------------------------------------------------------------------

area between 3 and 4 is a triangle

1/2 \* 0.25 \* 1

```{r}
.5 * .25 * 1
```

### Question 2b:

------------------------------------------------------------------------

What proportion of students at the university have a GPA between 1 and 3?

#### Solution to Question 2b:

------------------------------------------------------------------------

steps to calculate:

1)  sketch the area

2)  compute the are using geometry or calculus

area between 0 and 1 = area between 3 and 4 (graph is a normal distribution)

total area = 1

.125 + area between 1 and 3 + .125 = 1

```{r}
1-.125-.125
```

# Continuous Probability Distributions

------------------------------------------------------------------------

## Probability Density Function (pdf)

------------------------------------------------------------------------

If $X$ is a **continuous** random variable, the[**probability density function (pdf)**]{style="color: blue;"}, denoted $f(x)$ satisfies the following properties:

-   $f(x) \geq 0$ for all $x$,
-   $\displaystyle \int_{-\infty}^{\infty} f(x) = 1$, and
-   $\displaystyle P(a < x < b) = \int_a^b f(x) \, dx$

```{r, echo=FALSE}
# Data
set.seed(3)
y <- rnorm(150)

# Density estimation
den <- density(y)

# Plot
plot(den, 
     xlab="x", 
     ylab="f(x)", 
     main="P(a < x < b)",
     xaxt ="n")
axis(1, at = c(-1.5, 0.75),
     labels = c("a", "b"))

# Area between -1.5 and 2
value1 <- -1.5
value2 <- 0.75

# Lower and higher indices on the X-axis
l <- min(which(den$x >= value1))
h <- max(which(den$x < value2))

polygon(c(den$x[c(l, l:h, h)]),
        c(0, den$y[l:h], 0),
        col = "slateblue1")

```

## Question 3:

------------------------------------------------------------------------

The probability of a transistor failing between time $x=a$ and $x=b$ months is given by the probability density function

### $$ f(x) = c \int_a^b e^{-cx} \, dx.$$ Question 3a:

------------------------------------------------------------------------

If the probability of failure within the first six months, $0 < x < 6$, is 10%, set up (but do not solve) an equation to find the value of $c$?

#### Solution to Question 3a:

------------------------------------------------------------------------

$$
f(x) = c \int_0^6 e^{-cx} \, dx = 0.1
$$\
\

### Question 3b:

------------------------------------------------------------------------

Set up (but do not evaluate) an integral to represent the probability the transistor fails within the second six months ($6<x<12$)?

#### Solution to Question 3b:

------------------------------------------------------------------------

$$
f(x) = c \int_6^{12} e^{-cx} \, dx = 0.1
$$

\
\
\

### Question 3c:

------------------------------------------------------------------------

Interpret the meaning of $P(X \leq 9)$ and $P(X < 9)$.

#### Solution to Question 3c:

------------------------------------------------------------------------

P(X \<= 9): the probability that the transistor fails in less than or equal to 9 months.

P(X\<9): the probability that the transistor fails in less than 9 months.\

\
\
P(X\<=9) and P(X\<9) are the same\

## Cumulative Distribution Function (cdf)

------------------------------------------------------------------------

If $X$ is a **continuous** random variable, the [**cumulative distribution function (cdf)**]{style="color: blue;"}, denoted $F(x)$ is

$$P(X < x) = F(x) = \int_{-\infty}^x f(t) \, dt.$$

[**In other words,** $F(x)$ is an antiderivative of $f$,]{style="color: blue;"} and [$f(x)$ is the derivative of $F(x)$]{style="color: red;"}.

PDF is the slope of CDF.

## Question 4:

------------------------------------------------------------------------

Match the graphs of the density functions (a), (b), and (c) with the graphs of the cumulative distribution functions I, II, and III.

<img src="https://drive.google.com/uc?id=1toF3ssRugZB5_xu_7MDeNa4YgIaIsZFj" alt="PDf to CDF Matching Figure"/>

### Solution to Question 4:

------------------------------------------------------------------------

\

| Graph (a) | Graph (b) | Graph (c) |
|-----------|-----------|-----------|
| ii        | i         | iii       |

\
CDFs are non-decreasing and flatten out at 1.\

# Mean, Median, and Variance of Continuous Random Variables

------------------------------------------------------------------------

-   The [**mean**]{style="color: blue;"} of a continuous random variable is

$$E(X) = \mu = \int_{-\infty}^{\infty} x \cdot f(x) \, dx.$$

-   The [**variance**]{style="color: blue;"} of a continuous random variable is

$$\mbox{Var}(X) = E(X-\mu)^2 = E(X^2) - \big( E(X) \big)^2  \ \ \ \mbox{(this can be proven similar as with discrete case)}.$$

-   The [**median**]{style="color: blue;"} is the value $x$ such that $P(X < x) = 0.5$. Thus to find the median we solve the following for $x$:

$$F(x) = \int_{-\infty}^x f(t) \, dt = 0.5.$$

# Practice:

------------------------------------------------------------------------

## Question 5:

------------------------------------------------------------------------

Consider the random variable with pdf

$$ f(x) = \left\{ \begin{array}{ll} 
\frac{x}{8}, &  0 \leq x \leq 4 \\ 
0, &  \mbox{otherwise} 
\end{array} \right. .$$

### Question 5a:

------------------------------------------------------------------------

On a separate piece of paper, sketch a graph of the pdf, $f$.

#### Solution to Question 5a:

------------------------------------------------------------------------

Sketch a graph on a separate piece of paper.

\
\

### Question 5b:

------------------------------------------------------------------------

Give a formula for the cdf, $F$, and sketch its graph (on a separate piece of paper).

#### Solution to Question 5b:

------------------------------------------------------------------------

Type the formula for $F$ below, and then sketch the graph on another piece of paper.

\

$$F_X(x) = \left\{ \begin{array}{ll}
0 & x < 0\\
\frac{x^2}{16} & 0 \leq x \leq 4 \\
1 & x > 4 
\end{array} \right.$$

\
\

### Question 5c:

------------------------------------------------------------------------

Calculate $P(X < 1)$ and illustrate this value on both of your graphs.

#### Solution to Question 5c:

------------------------------------------------------------------------

\
\
\
\

### Question 5d:

------------------------------------------------------------------------

Calculate $E(X)$.

#### Solution to Question 5d:

------------------------------------------------------------------------

\
\
\
\

### Question 5e:

------------------------------------------------------------------------

Give the median value and illustrate this value on both of your graphs.

#### Solution to Question 5e:

------------------------------------------------------------------------

\
\
\
\
