---
title: '10: Sampling Distributions for Proportions'
author: "Adam Spiegler, University of Colorado Denver"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Sampling from a Binomial Distribution

------------------------------------------------------------------------

Very often we encounter statistical questions that ask us to approximate or compare [**proportions**]{style="color: blue;"}.
For example

-   What proportion of voters support a new law?
-   What proportion of the population follow public health recommendations?
-   For a certain model smartphone, what proportion of all smartphones produced are defective?

## Computing the Sample Proportion ${\large \widehat{P}}$

------------------------------------------------------------------------

In these cases, we have a certain population in mind.
A sample size $n$ is randomly selected.

-   Each selection is considered a trial. We have $n$ trials.
-   Since we pick from the same population, the probability of a success in each trial is $p$.
-   We count $X$, the number of "successes" out of $n$ independent and identical trials.
-   Note that we have $X \sim \mbox{Binom}(n,p)$.
-   Then we can calculate the sample proportion:

$$\hat{p} = \frac{\mbox{Number of successes}}{\mbox{Size of sample}} = \frac{X}{n}.$$ If we repeat this process over and over again (say 1000 times), then we can look at the [**Distribution of Sample Proportions**]{style="color: blue;"} that we denote $\color{blue}{\widehat{P}}$, and we often call this statistic "p hat".

p = population proportion, p hat is sample proportion

proof:

$E(\frac{X}{n}) = \frac{1}{n}E(X) = \frac{1}{n}(np) = p = E(\widehat{P})$

$Var(\frac{X}{n}) = \frac{1}{n^2}Var(X) = \frac{1}{n^2}(npq) = \frac{p(1-p)}{n} = Var(\widehat{P})$

$\sigma_{\widehat{P}} = SE(\widehat{P}) = \sqrt{\frac{p(1-p)}{n}}$

## Question 1:

------------------------------------------------------------------------

Let $p$ denote the proportion of all packages mailed by the United States Postal Service (USPS) that contain illegal contents.
We randomly select a sample of $n$ packages, count the number of illegal packages $x$, and then compute the proportion of packages in the sample that have illegal contents $\hat{p} =\frac{x}{n}$.
We repeat this over and over again and construct the distribution of sample proportions, $\widehat{P}$.

### Question 1a:

------------------------------------------------------------------------

Let $X$ denote the number of successes (illegal packages) out of a random sample of $n$ USPS packages.
What are the mean and standard deviation of $X$.
Your answers will depend on $n$ and $p$.

#### Solution to Question 1a:

------------------------------------------------------------------------

\
\
\

### Question 1b:

------------------------------------------------------------------------

Let $\widehat{P} = \frac{X}{n}$ denote the distribution of sample proportions.
Using formulas from part (a) and properties of expected value and variance, give formulas for $E( \widehat{P} )$ and $\mbox{Var}( \widehat{P} )$.

#### Solution to Question 1b:

------------------------------------------------------------------------

\
\
\

# Central Limit Theorem for Proportions

------------------------------------------------------------------------

Let $X \sim \mbox{Binom}(n,p)$ be a binomial random variable, and let $\widehat{P} = \frac{X}{n}$ denote the distribution of sample proportions.
Then if the sample is large enough [**both** $\mathbf{np \geq 10}$ and $\mathbf{n(1-p) \geq 10}$]{style="color: blue;"} , the sampling distribution for $\widehat{P}$ will:

-   Be (approximately) normally distribution.
-   Have mean equal to the population proportion, $p$.
-   Have standard error $\mbox{SE}(\widehat{P}) = \sqrt{\frac{p(1-p)}{n}}$.

We summarize the results more concisely below:

$$\color{blue}{\widehat{P} \sim N \left( \mu_{\widehat{P}} , \sigma_{\widehat{P}} \right) = N \left( p  , \sqrt{\frac{p(1-p)}{n}} \right)}$$

## Question 2:

------------------------------------------------------------------------

Census Bureau data for 2017 shows nearly half (48 percent) of residents in United States five largest cities now speak a language other than English at home[^1].
If a sample of 150 people are selected at random from the five largest cities in the US, what is the probability that at most 40% speak a language other than English at home?

[^1]: [https://cis.org/Report/Almost-Half-Speak-Foreign-Language-Americas-Largest-Cities](https://cis.org/Report/Almost-Half-Speak-Foreign-Language-Americas-Largest-Cities%7D)

### Question 2a:

------------------------------------------------------------------------

Is $n$ large enough to use the CLT?
Explain why or why not?

#### Solution to Question 2a:

------------------------------------------------------------------------

$np = 150(0.48) \geq 10 \ (yes)$ $nq = 150(0.52) \geq 10 \ (yes)$

\
\
\

### Question 2b:

------------------------------------------------------------------------

Using the CLT for a proportion, find the z-score of the proportions $0.44$ and $0.48$.

#### Solution to Question 2b:

------------------------------------------------------------------------

$\sigma_{\widehat{P}} = \sqrt{\frac{(0.48)(0.52)}{150}}$

$z = \frac{(0.44 - 0.48)}{\sqrt{\frac{(0.48)(0.52)}{150}}}$

about 0.04079 for SE

```{r}
sqrt((0.48*0.52)/150)

#zscore for 0.44
(0.44-0.48) / sqrt((0.48*0.52)/150)

#zscore for 0.48
(0.48-0.48) / sqrt((0.48*0.52)/150)
```

### Question 2c:

------------------------------------------------------------------------

What is the probability that between 44% and 48% (out of the random sample of 150 people) speak a language other than English at home?
**(what a question might look like on an exam)**

#### Solution to Question 2c:

------------------------------------------------------------------------

$P(0.44 \leq \widehat{P} \leq 0.48)$

\*fill in the blanks

```{r}
pnorm(0.48, 0.48, 0.04079) - pnorm(0.44, 0.48, 0.04079)
```

```{r, echo=FALSE, eval=TRUE}
library(ggplot2)
mux <- 0.48
sdx <- (0.48*(1-0.48)/150)**(1/2)
xmax <- mux + 3*sdx
xmin <- mux - 3*sdx

ggplot(NULL, aes(c(xmin, xmax))) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(xmin, 0.44)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = "red", 
            xlim = c(0.44, 0.48)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(0.48, xmax)) +
  labs(x = "p-hat", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous("Sampling Distribution for Proportion", breaks=c(0.44,0.48)) + 
  geom_hline(yintercept=0) + 
  theme_bw()
```

# Continuity Correction for Discrete Random Variables

------------------------------------------------------------------------

A binomial random variable is a **discrete random variable**, but a normal distribution approximation is **continuous density function**.
When using a normal distribution to approximate the sampling distribution for a discrete random variable $X$, we can improve the estimate by using a [**continuity correction**]{style="color: blue;"} as follows.
If we want to calculate $P( a \leq X \leq b)$ where $a<b$ are integers, then we use the following correction:

$$\color{blue}{P( a \leq X \leq b) \approx P(a-0.5 < X < b+0.5) =  P \left( \frac{(a-0.5)}{n} < \widehat{P} < \frac{(b+0.5)}{n} \right)}.$$

```{r, echo=FALSE, eval=TRUE}
x <- 1:150
y <- dbinom(x, size=150, prob=.48)
breaks <- c(1, 65, 73, 150)
cols <- c("white", "grey", "white")[findInterval(x, vec=breaks)]

barplot(y, names.arg=c(1:150), space=0, xlim=c(50,94), ylim=c(0,0.08), col = cols, yaxt='n', xpd=FALSE)
lines((c(64,73)-.5), dnorm(c(64,73), 72, sqrt(150*.48*.52)), xlim=c(50,94), type="h", lwd=2, col="red")
curve(dnorm(c(x), mean=72, sd=sqrt(150*.48*.52)), from=50, to=94, xlim = c(50,94), add=T, col="blue")
```

## Question 3:

------------------------------------------------------------------------

In [Question 2c:] we calculated $P\left( 0.44 \leq \widehat{P} \leq 0.48 \right)$ using a normal distribution (by way of the CLT).
We could equivalently rewrite this probability in terms of the discrete random variable $X \sim \mbox{Binom}(150,0.48)$ as $P( 66 \leq X \leq 72$).

### Question 3a:

------------------------------------------------------------------------

Using a binomial distribution, calculate the exact value of $P( 66 \leq X \leq 72$).

#### Solution to Question 3a:

------------------------------------------------------------------------

```{r}
pbinom(72, 150, 0.48) - pbinom(65, 150, 0.48)
```

if $\widehat{P} = 0.44 => X_1 = (0.44)(150) = 66, X_2 = (0.48)(150) = 72$\

```{r, echo=FALSE, eval = TRUE}
library(dplyr)
#library(ggplot2)

data.frame(success = 48:96, 
           pmf = dbinom(x = 48:96, size = 150, prob = 0.48),
           cdf = pbinom(q = 48:96, size = 150, prob = 0.48, 
                        lower.tail = TRUE)) %>%
  mutate(Success = ifelse(success >= 65 & success <= 72, "Between 65 and 72", "Not")) %>%
ggplot(aes(x = factor(success), y = pmf, fill = Success)) +
  geom_col() +
  labs(title = "Binomial Dist n=150 and p=0.48", x = "Number of Households out of 150", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_discrete(breaks=c(48, 65, 72, 96)) + 
  geom_hline(yintercept=0) + 
  theme_bw()
```

### Question 3b:

------------------------------------------------------------------------

Calculate the $z$-score using a corrected lower limit, $\hat{p}_1^* = \frac{66-0.5}{150} = 0.4367$.

#### Solution to Question 3b:

------------------------------------------------------------------------

\
\
\

### Question 3c:

------------------------------------------------------------------------

Calculate the $z$-score using a corrected upper limit, $\hat{p}_2^* = \frac{72+0.5}{150} = 0.4833$.

#### Solution to Question 3c:

------------------------------------------------------------------------

\
\
\

### Question 3d:

------------------------------------------------------------------------

Using the z-scores from [Question 3b:] and [Question 3c:] obtained by applying a continuity correction, approximate the probability that between 44% and 48% (out of the random sample of 150 people) speak a language other than English at home?

#### Solution to Question 3d:

------------------------------------------------------------------------

```{r}
pnorm(0.4833, 0.48, 0.04079) - pnorm(0.4367, 0.48, 0.04079)
```

```{r, echo=FALSE, eval=TRUE}
library(ggplot2)
mux <- 0.48
sdx <- (0.48*(1-0.48)/150)**(1/2)
xmax <- mux + 3*sdx
xmin <- mux - 3*sdx

ggplot(NULL, aes(c(xmin, xmax))) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(xmin, 0.4367)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = "blue", 
            xlim = c(0.4367, 0.44)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = "red", 
            xlim = c(0.44, 0.48)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = "blue", 
            xlim = c(0.48, 0.4833)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(0.4833, xmax)) +
  labs(x = "p-hat", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous("Sampling Distribution for Proportion", breaks=c(0.4367, 0.4833)) + 
  geom_hline(yintercept=0) + 
  theme_bw()
```

### Question 3e:

------------------------------------------------------------------------

Compare approximations from [Question 2c:] and [Question 3d:] with the exact calculation in [Question 3a:].
Comment on whether or not the continuity correction improved the approximation or not.

#### Solution to Question 3e:

------------------------------------------------------------------------

The continuity correction improved the approximation.\

\
\

# Sampling Distributions for Other Statistics

------------------------------------------------------------------------

Frequently we are interested in the minimum value in a random sample, denoted $X_{\rm{min}}$, or maximum value, denoted $X_{\rm{max}}$ of random sample.

## Question 4:

------------------------------------------------------------------------

Derive the cdf for the maximum of a random sample, $X_1,X_2, \ldots, X_n$ each independently picked from a distribution with corresponding cdf $F(x)$.

$$\begin{aligned}
F_{X_{\rm{max}}} (a) &= P(X_1 \leq a, X_2 \leq a, \ldots, X_n \leq a) & \mbox{Explanation 1} \\
&= P\big( (X_1 \leq a) \cap (X_2 \leq a) \cap \ldots \cap (X_n \leq a) \big) & \mbox{Explanation 2}\\
&= P(X_1 \leq a) \cdot P(X_2 \leq a) \cdot  \ldots  \cdot P(X_n \leq a) & \mbox{Explanation 3}\\
&= F(a) \cdot F(a) \cdot \ldots \cdot F(a) & \mbox{Explanation 4}\\
&= \big( F(a) \big)^n
\end{aligned}$$

### Solution to Question 4:

------------------------------------------------------------------------

**Explanation 1:**

**Explanation 2:**

**Explanation 3:**

**Explanation 4:**

## Question 5:

------------------------------------------------------------------------

Using the result from the previous problem, find $\displaystyle f_{X_{\rm{max}}} (a)$, the pdf for the maximum of a random sample.

*Hint: Recall the relation between cdf's and pdf's of continuous random variables.*

### Solution to Question 5:

------------------------------------------------------------------------

\
\
\

## Question 6:

------------------------------------------------------------------------

Let $X \sim \mbox{Unif}(0,1)$.

### Question 6a:

------------------------------------------------------------------------

What are the cdf and pdf, $F(x)$ and $f(x)$ respectively, of $X$?

#### Solution to Question 6a:

------------------------------------------------------------------------

\
\
\

### Question 6a:

------------------------------------------------------------------------

If we pick a random sample of size $n=10$, how likely is it that $X_{\rm{max}}$ is greater than or equal to $0.9$?

#### Solution to Question 6a:

------------------------------------------------------------------------

\
\
\
