---
title: '16: Confidence Intervals for Means'
author: "Adam Spiegler, University of Colorado Denver"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Estimating a Population Mean: Average Sea Surface Temperature

------------------------------------------------------------------------

What is the average sea surface temperature (SST) on Earth?
There are a variety of methods that have been used to estimate this parameter.
One possible method would be to randomly select locations and times around the world from which to sample, and then compute the average temperature of the sample.
The graph[^1] below plots interval estimates for the deviation of the average sea surface temperature compared to the 1971-2000 average.

[^1]: Image Credit: United States Environmental Protection Agency, <https://www.epa.gov/climate-indicators/climate-change-indicators-sea-surface-temperature>, Accessed Nov. 1, 2019.

![Average Sea Surface Temperature](https://drive.google.com/uc?export=view&id=11nZqO-suXmiOZqtj0wQ5nNOkhzX_QXYa)

-   We have already seen that one unbiased estimator for the parameter $\mu$ is the sample mean $\bar{x}$.
-   However, based on one sample we have no idea how far off $\bar{x}$ is from the actual value of $\mu$.
-   One way build some uncertainty into the estimate is to use a bootstrap distribution to get a range of plausible values.
-   In this problem set, we will learn some other methods for constructing confidence intervals.

## Question 1

------------------------------------------------------------------------

Let's imagine instead of the sea surface, we would like to estimate the average length of all words in the song "Raspberry Beret".
Let $X$ denote the length (measured by the number of letters) of a randomly selected word in "Raspberry Beret".
Like the sea surface, imagine the population $X$ consists of so many words that it is impractical to calculate the mean length of all words in "Raspberry Beret".

**Determine whether each of the statements below are TRUE or FALSE. Support your answer with an explanation.**

### Question 1a

------------------------------------------------------------------------

If we pick random samples of size $n$ words and compute the mean length of the sample of words, $\bar{X}$, the values of the $\bar{X}$'s will be different for different random samples.

#### Solution to Question 1a

------------------------------------------------------------------------

They should be different.\

\
\

### Question 1b

------------------------------------------------------------------------

Let $\mu$ and $\sigma$ denote the population mean and population standard deviation (for the length of all words).
The values of $\mu$ and $\sigma$ will be different for different random samples.

#### Solution to Question 1b

------------------------------------------------------------------------

\
They population parameters should be the same regardless of the sample.\

\

## Question 2

------------------------------------------------------------------------

Let $X$ denote the length (measured by the number of letters) of a randomly selected word in "Raspberry Beret" from [Question 1].

### Question 2a

------------------------------------------------------------------------

Enter R code to compute $P \left( -1.96 < Z < 1.96 \right)$ for $Z \sim N(0,1)$

#### Solution to Question 2a

------------------------------------------------------------------------

```{r}

pnorm(1.96,0,1)-pnorm(-1.96,0,1)  # Compute P(-1.96 < Z < 1.96)
```

$$\large{P \left( -1.96 < Z < 1.96 \right) = 0.95}$$

### Question 2b

------------------------------------------------------------------------

Let $\mu$ and $\sigma$ denote the mean and standard deviation, respectively, of all words in "Raspberry Beret".
Let $\bar{X}$ denote the distribution of sample means for samples size $n$.
Recall we denote the mean and standard error of the sampling distribution as $\mu_{\bar{X}}$ and $\sigma_{\bar{X}}$, respectively.

Justify each of the steps below to compute $P \left( \bar{X} - 1.96 \cdot \frac{\sigma}{\sqrt{n}} < \mu < \bar{X}+1.96 \cdot \frac{\sigma}{\sqrt{n}} \right)$

$$\begin{aligned}
?? &= P \left( -1.96 < Z < 1.96 \right) & \mbox{Substitute value from previous answer.}\\
   &= P \left( -1.96 < \frac{\bar{X} - \mu_{\bar{X}}}{\sigma_{\bar{X}}} < 1.96 \right) & \mbox{Explanation 1}\\
   &=  P \left( -1.96 < \frac{\bar{X} - \mu}{\frac{\sigma}{\sqrt{n}}} < 1.96 \right) & \mbox{Explanation 2}\\
   &= P \left( -1.96 \cdot \frac{\sigma}{\sqrt{n}} < \bar{X} - \mu <  1.96 \cdot \frac{\sigma}{\sqrt{n}} \right) & \mbox{Explanation 3} \\
   &= P \left( - \bar{X} -1.96 \cdot \frac{\sigma}{\sqrt{n}} < - \mu <  - \bar{X} + 1.96 \cdot \frac{\sigma}{\sqrt{n}} \right) & \mbox{Explanation 4} \\
   &= P \left( \bar{X} - 1.96 \cdot \frac{\sigma}{\sqrt{n}} < \mu < \bar{X}+1.96 \cdot \frac{\sigma}{\sqrt{n}} \right) & \mbox{Explanation 5}\\
\end{aligned}$$

#### Solution to Question 2b

------------------------------------------------------------------------

Explanation 1:

Using the definition of calculating z scores\
\

Explanation 2:

Replacing the definitions of mu and standard deviation with definitions of sample means using central limit theorem\
\

Explanation 3:

Multiply denominator on both sides of inequality\
\

Explanation 4:

Subtract sample mean of X on both sides of inequality

\

Explanation 5:

Multiply inequality by -1\
\
\

### Question 2c

------------------------------------------------------------------------

Interpret the practical meaning of the value of $P \left( \bar{X} - 1.96\cdot \sigma_{\bar{X}} < \mu < \bar{X}+1.96 \cdot \sigma_{\bar{X}} \right)=0.95$.

#### Solution to Question 2c

------------------------------------------------------------------------

This interval has a 95% chance of containing the actual population mean.\

A way to get a confidence interval without bootstrapping.\

We can do this because there is a central limit theorem for means.\

## Question 3

------------------------------------------------------------------------

Let $X$ denote the length (measured by the number of letters) of a randomly selected word in "Raspberry Beret".
Let's assume for now that although we do not know the value of the parameter $\mu$ (average number of letters of all words in the song), we do know that $\mbox{Var}(X) = \sigma^2 = 3.45$ letters.

The vector `lyrics` is a vector of all words from the song "Raspberry Beret".

```{r echo=FALSE}
# Create a vector of all words

lyrics <- c("Yeah", "I", "was", "working", "part", "time", "in", "a", "five", "and", "dime", "My", "boss", "was", "Mr", "McGee", "He", "told", "me", "several", "times", "that", "he", "didnt", "like", "my", "kind", "Cause", "I", "was", "a", "bit", "too", "leisurely", "Seems", "that", "I", "was", "busy", "doing", "something", "close", "to", "nothing", "But", "different", "than", "the", "day", "before", "Thats", "when", "I", "saw", "her", "ooh", "I", "saw", "her", "She", "walked", "in", "through", "the", "out", "door", "out", "door", "She", "wore", "a", "Raspberry", "beret", "The", "kind", "you", "find", "in", "a", "second", "hand", "store", "Raspberry", "beret", "And", "if", "it", "was", "warm", "she", "wouldnt", "wear", "much", "more", "Raspberry", "beret", "I", "think", "I", "love", "her", "Built", "like", "she", "was", "She", "had", "the", "nerve", "to", "ask", "me", "If", "I", "planned", "to", "do", "her", "any", "harm", "So", "look", "here", "I", "put", "her", "on", "the", "back", "of", "my", "bike", "And", "we", "went", "riding", "Down", "by", "old", "man", "Johnsons", "farm", "I", "said", "now", "overcast", "days", "never", "turned", "me", "on", "But", "something", "about", "the", "clouds", "and", "her", "mixed", "She", "wasnt", "to", "bright", "But", "I", "could", "tell", "when", "she", "kissed", "me", "She", "knew", "how", "to", "get", "her", "kicks", "She", "wore", "a", "Raspberry", "beret", "The", "kind", "you", "find", "in", "a", "second", "hand", "store", "Raspberry", "beret", "And", "if", "it", "was", "warm", "she", "wouldnt", "wear", "much", "more", "Raspberry", "beret", "I", "think", "I", "love", "her", "The", "rain", "sounds", "so", "cool", "when", "it", "hits", "the", "barn", "roof", "And", "the", "horses", "wonder", "who", "you", "are", "Thunder", "drowns", "out", "what", "the", "lightning", "sees", "You", "feel", "like", "a", "movie", "star", "Listen", "They", "say", "the", "first", "time", "aint", "the", "greatest", "But", "I", "tell", "ya", "If", "I", "had", "the", "chance", "to", "do", "it", "all", "again", "I", "wouldnt", "change", "a", "stroke", "Cause", "baby", "Im", "the", "most", "With", "a", "girl", "as", "fine", "as", "she", "was", "then", "The", "kind", "you", "find", "The", "kind", "you", "find", "The", "kind", "you", "find", "In", "a", "second", "hand", "store", "Oh", "no", "no", "Raspberry", "beret", "And", "if", "it", "was", "warm", "Where", "have", "all", "the", "raspberry", "women", "gone", "Yeah", "Raspberry", "beret", "I", "think", "I", "I", "think", "I", "I", "think", "I", "love", "her", "No", "no", "no", "No", "no", "no", "The", "kind", "you", "find", "In", "a", "second", "hand", "store", "Raspberry", "beret", "Tell", "me", "Where", "have", "all", "the", "raspberry", "women", "gone", "And", "if", "it", "was", "warm", "she", "Wouldnt", "wear", "much", "more", "Raspberry", "beret")
```

The code cell below converts the vector `lyrics` to a numeric vector named `all.words` that gives the number of letters of each word in `lyrics`.

```{r}
all.words <- nchar(lyrics)  # nchar calculates the number of characters in a string
head(lyrics)
head(all.words)

#x = how many letters are in a randomly selected word
#nchar(word) = 4 (length of the string)
#all.words = length of each word in song
```

### Question 3a

------------------------------------------------------------------------

Using the sample function, pick a random sample (without replacement) of $n=30$ words from all the lyrics.
Print your 30 selected words to the screen.

#### Solution to Question 3a

------------------------------------------------------------------------

```{r}
# Pick a random sample of 30 words, without replacement
sample.lyrics <- sample(lyrics, 30, replace = FALSE)  

# convert the vector sample.lyrics to a numeric vector
# each word in sample lyrics is replaced by the 
# corresponding number of letters of each word
sample.length <- nchar(sample.lyrics)  

# print your vectors to the screen
sample.lyrics
sample.length
```

### Question 3b

------------------------------------------------------------------------

Create a histogram to display the lengths of the words in your sample.

#### Solution to Question 3b

------------------------------------------------------------------------

```{r}
# Enter code to create a histogram to display your sample
hist(sample.length)
```

### Question 3c

------------------------------------------------------------------------

Using the CLT what is the value of $\sigma_{\bar{X}}$, the standard error of the sampling distribution for $\bar{X}$?
Show your work below.

#### Solution to Question 3c

------------------------------------------------------------------------

```{r}
variance <- var(all.words)
se <- sqrt(variance)/sqrt(30) #sigma / sqrt(n)

se
```

standard error of sample, use sqrt(variance) / sqrt(n)

standard deviation of a population, use r command sd

### Question 3d

------------------------------------------------------------------------

Give an interval of values $3.40126 < \mu < 4.732074$ that has a 95% chance of containing the actual value of $\mu$.
Show or explain how you determined your answer.

#### Solution to Question 3d

------------------------------------------------------------------------

```{r}
x.bar <- mean(sample.length)
(lower <- x.bar - 1.96 * se)
(upper <- x.bar + 1.96 * se)
```

# Confidence Interval for a Mean

------------------------------------------------------------------------

For a sample size $n$ drawn from a normal distribution with unknown $\mu$ and known $\mbox{Var}(X)=\sigma^2$, a [**95% confidence interval for the mean**]{syle="color: blue;"} is

$$\color{blue}{\large \bar{X} - 1.96 \frac{\sigma}{\sqrt{n}} < \mu <  \bar{X} + 1.96 \frac{\sigma}{\sqrt{n}}}.$$

If we draw 1000's of random samples each size $n$ from a normal distribution with parameters $\mu$ and $\sigma$ and compute a 95% confidence interval from each sample, then [**about 95% of the intervals would contain** $\mu$]{style="color: blue;"}.

# Interpreting Confidence Intervals

------------------------------------------------------------------------

## Question 4

------------------------------------------------------------------------

A researcher calculates collects a random sample of words and gets a 95% confidence interval for the mean length of a word that is from $2.57$ letters to $3.90$ letters.
**For each statement, determine whether the interpretation is correct or not. If not, explain why not.**

### Question 4a

------------------------------------------------------------------------

There is a 95% chance that randomly selected word has between $2.57$ and $3.90$ letters.

#### Solution to Question 4a

------------------------------------------------------------------------

\
Not correct.
Confidence interval tells us if there is a chance that the confidence interval contains the population mean, it does not give a probability.
It is used to estimate a population parameter of interest.

Not used to estimate length of one individual in population.\

\

### Question 4b

------------------------------------------------------------------------

There is a 95% chance that the mean length of all words in "Raspberry Beret" is between $2.57$ and $3.90$ letters.

#### Solution to Question 4b

------------------------------------------------------------------------

\
Not correct.
Since confidence intervals can vary, confidence intervals does not ensure that the mean is in the given interval.
Confidence intervals talk about sureness of an interval, not sureness of a population parameter.

Uncertainty is attached to the interval estimate due to the randomness of sampling.\

\

### Question 4c

------------------------------------------------------------------------

The interval from $2.57$ and $3.90$ letters will contain the mean length of all words in "Raspberry Beret" 95% of the time.

#### Solution to Question 4c

------------------------------------------------------------------------

Correct.\

\
\

### Question 4d

------------------------------------------------------------------------

95% of all random samples of size $n=30$ words have a sample mean length between $2.57$ and $3.90$ letters.

#### Solution to Question 4d

------------------------------------------------------------------------

Confidence interval should estimate a population parameter.
This estimates a sample parameter, which is not correct interpretation of confidence interval.\

\
\

### Question 4e

------------------------------------------------------------------------

We are 95% confident that the mean length of all words in the song Raspberry Beret is between $2.57$ and $3.90$ letters.

#### Solution to Question 4e

------------------------------------------------------------------------

Correct.\

\
\

```{r echo=FALSE}
counter <- 0
plot(x = c(0, 8), y = c(1, 100), type = "n", xlab = "", ylab = "")

for (i in 1:100)
{
  x <- sample(all.words, size = 30, replace = FALSE)
  L <- mean(x) - 1.96*sqrt(3.45/30)
  U <- mean(x) + 1.96*sqrt(3.45/30)
  if ( L < mean(all.words) && mean(all.words) < U)
  {
    counter <- counter + 1
    plot.color  <- "green"
  }
  else
    plot.color  <- "red"
  segments(L, i, U, i, col = plot.color)    
}

abline(v = mean(all.words), col = "blue")
```

The proportion of confidence intervals from the code cell above that contain the actual value of the population mean is given below.

```{r}
# Out of 100 confidence intervals:
# The proportion that are successful is
counter/100
```

## Question 5

------------------------------------------------------------------------

What are some warnings to keep in mind when interpreting confidence intervals?

### Solution to Question 5

------------------------------------------------------------------------

Warning 1: Measure confidence interval of a population parameter

\
\

Warning 2: Randomness is due to sampling process

\
\

## Question 6

------------------------------------------------------------------------

Researchers want to estimate the average length of all female Kamodo dragons.
The pick a random sample of female Kamodo dragons with the following weights (in pounds):

```{r}
# Sample weights 
dragon.wt <- c(145, 178, 142, 139, 160, 190, 168, 122)
dragon.wt
```

and suppose they know that **the population standard deviation** of weights of all Kamodo dragons is $\sigma = 25$ pounds.

Process:

-   pick 1 sample size n

-   compute an initial point estimate = sample statistic

    -   either a sample mean or sample proportion

-   build some uncertainty to the estimate

    -   MoE (margin of error) = 1.96 x SE

        -   1.96 is the value for 95% confidence interval

    -   SE($\bar{X}$) = $\frac{\sigma}{\sqrt{n}}$

        -   25 / sqrt(8) = 8.84

    -   MoE of example problem above: 1.96 \* 8.84

-   Construct confidence interval

    -   $\bar{X} \pm MoE$

### Question 6a

------------------------------------------------------------------------

Give a 95% confidence interval, and interpret the meaning of your answer.

#### Solution to Question 6a

------------------------------------------------------------------------

```{r}
dragon.mean <- mean(dragon.wt)  # Enter a command to calculate the sample mean
dragon.se <- 25 / sqrt(8)
dragon.lower95 <- dragon.mean - 1.96 * dragon.se   # Enter a command to calculate the lower limit
dragon.upper95 <- dragon.mean + 1.96 * dragon.se   # Enter a command to calculate the upper limit

# Print your answers
dragon.lower95
dragon.upper95
```

\

**Interpret your answer in practical terms.**

\
\
\

### Question 6b

------------------------------------------------------------------------

Give a 90% confidence interval, and interpret the meaning of your answer.

#### Solution to Question 6b

------------------------------------------------------------------------

```{r}
new.z <- qnorm(0.95,0,1)  # Enter a command to calculate z value for 90% Conf Level
  # confidence level (area we want) + include left tail (leftover area/2): 0.9 + 0.05
dragon.lower90 <- dragon.mean - new.z * dragon.se   # Enter a command to calculate the lower limit
dragon.upper90 <- dragon.mean + new.z * dragon.se   # Enter a command to calculate the upper limit

# Print your answers
dragon.lower90
dragon.upper90
```

**Interpret your answer in practical terms.**

There is a 90% confidence that this mean is between these values.\

\
\

### Question 6c

------------------------------------------------------------------------

As we **decrease the confidence level of the interval**, what happens to the width of the interval estimate?

#### Solution to Question 6c

------------------------------------------------------------------------

If the confidence interval is decreased, the interval is less likely to be successful.
The width of the interval is going to shrink.
If the interval is more narrow, the interval is more specific which is helpful for making inferences.\

\
point estimate: estimating population parameter with just one value\

# Changing the Confidence Level

------------------------------------------------------------------------

If $X_i \sim N(\mu, \sigma^2)$, $i=1,2,\ldots , n$ with known $\sigma$, and confidence level equal to CL, then a corresponding confidence interval is given by

$$\bar{X} - z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}} < \mu < \bar{X} + z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}},$$

where the area under $N(0,1)$ between $\pm z_{\alpha/2}$ is equal to the confidence level.

-   Let $\color{blue}{CL}$ denote a selected [**confidence level**]{style="color: blue;"}.
-   We have $\color{red}{\alpha = 1 - CL}$.
-   The [**area in each tail**]{style="color: green;"} is therefore $\color{green}{\dfrac{\alpha}{2}}$.

```{r message=FALSE}
# You will need load packages before creating plot below
library(ggplot2)
library(latex2exp)
```

```{r echo=FALSE}
mux <- 0
sdx <- 1

ggplot(NULL, aes(c(-3, 3))) + 
  geom_area(stat = "function",   fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(-3, -1.65)) +
  geom_area(stat = "function",   fun = dnorm, args = list(mean = mux, sd = sdx),
            color = "black", fill = "red", 
            xlim = c(-1.65, 1.65)) +
  geom_area(stat = "function",   fun = dnorm, args = list(mean = mux, sd = sdx), 
            color = "black", fill = NA, 
            xlim = c(1.65, 3)) +
  labs(x = "", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks=c(-1.65, 0, 1.65), labels=c(TeX("$-z_{\\alpha/2}$"), "0",  TeX("$z_{\\alpha/2}$"))) + 
  geom_hline(yintercept=0) + 
  theme_bw()
```

-   The distance $\color{blue}{\mbox{MoE} = z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}}$ is called the [**Margin of Error (MoE)**]{style="color: blue;"} of the confidence interval.

# Confidence Intervals for Means, Variance $\sigma^2$ Unknown

------------------------------------------------------------------------

In most real-life settings, in addition to not knowing the population mean $\mu$, we **do not know the value of the variance of the population**, $\mbox{Var}(X)=\sigma^2$.
In deriving a confidence interval, we have used $\bar{x}$ as an estimate of $\mu$, so it seems natural to use the [**sample standard deviation**]{style="color: red;"} $\color{red}{s}$ as an estimate for $\sigma$.
The plot on the left gives a histogram for the standardized sampling distribution

$$W = \frac{\bar{X}-\mu}{\color{red}{s}/\sqrt{n}}$$

and the plot on the right is a qqplot comparing the distribution of $W$ to $N(0,1)$.

```{r echo=FALSE}
mu <- 156  # population mean 156 pounds
sigma <- 25  # population sd is 25 pounds
N <- 10^4
w <- numeric(N)
n <- 8  # sample size 8 Kamodo dragons

for (i in 1:N)
{
  x <- rnorm(n, mu, sigma)  # pick a random sample of 8 Kamodo dragons
  x.mean <- mean(x)  # sample mean
  s <- sd(x)  # sample standard deviation
  w[i] <- (x.mean - 156)/(s/sqrt(n))  # standardized distance
}

par(mfrow=c(1,2))
hist(w)

qqnorm(w)
qqline(w)
par(mfrow=c(1,1))
```

# T Confidence Interval for a Mean: Unknown Population Variance

------------------------------------------------------------------------

If $X_i \sim N(\mu, \sigma^2)$, $i=1,2,\ldots , n$ with unknown $\mbox{Var}(X)=\sigma^2$, and confidence level equal to CL, then a corresponding confidence interval is given by

$$\bar{X} - \color{red}{t_{\alpha/2}} \cdot \frac{\color{red}{s}}{\sqrt{n}} < \mu < \bar{X} + \color{red}{t_{\alpha/2}} \cdot \frac{\color{red}{s}}{\sqrt{n}},$$

where the area under the $t$-distribution with $n-1$ degrees of freedom between $\pm t_{\alpha/2}$ is equal to the confidence level.

-   The command `qt(0.975, 7)` gives the $0.975$ quantile from a $t$-distribution with 7 degrees of freedom.
    -   The command `pt(2.5, 7)` gives the $P(T < 2.5)$ for random variable $T$ from a $t$-distribution with 7 degrees of freedom.

## Question 7

------------------------------------------------------------------------

Complete the R code cell below to construct a 95% confidence interval for the sample of Kamodo dragons from [Question 6] assuming the variance of the population is unknown.

### Solution to Question 7

------------------------------------------------------------------------

```{r}
se.samp <- sd(dragon.wt)/sqrt(8) #sd(dragon.wt) gets s, the sample standard deviation
t.star <- qt(0.975, 7)  # t_{\alpha/2} for 95%, 7 is the degrees of freedom (8[num samp]-1)
(lower.tconf <- dragon.mean - t.star * se.samp) # compute lower cutoff of t conf interval
(upper.tconf <- dragon.mean + t.star * se.samp)  # compute upper cutoff of t conf interval
```

Z confidence interval (when we know the standard deviation) is more precise than the T confidence interval (when we don't know the standard deviation of population)

## Methods for Finding $T$ Confidence Intervals

------------------------------------------------------------------------

-   Use the formula $\displaystyle \bar{X} - \color{red}{t_{\alpha/2}} \cdot \frac{\color{red}{s}}{\sqrt{n}} < \mu < \bar{X} + \color{red}{t_{\alpha/2}} \cdot \frac{\color{red}{s}}{\sqrt{n}}$.

-   You can also use a $t$-distribution table to estimate values for $t_{\alpha/2}$ and areas under $t$-distributions with specified degrees of freedom.

-   The command `t.test(x, conf.level = 0.95)$conf` will give a 95% confidence level.

```{r}
t.test(dragon.wt, conf.level = 0.95)$conf.int
```

```{r echo=FALSE}
# Generate a vector of 100 values between -6 and 6
x <- seq(-6, 6, length = 100)

# Degrees of freedom
df = c(1,3,8,30)
color = c("red", "blue", "green", "gold", "black")

# Plot a normal distribution
plot(x, dnorm(x), type = "l", lty = 2, xlab = "t-value", ylab = "Density",
	main = "Comparison of t-distributions", col = "black")

# Add the t-distributions to the plot
for (i in 1:4)
{
  lines(x, dt(x, df[i]), col = color[i])
}

# Add a legend
legend("topright", c("df = 1", "df = 3", "df = 8", "df = 30", "normal"),
	col = color, title = "t-distributions", lty = c(1,1,1,1,2))
```

## Impact of the Shape of the Population

------------------------------------------------------------------------

-   When the population is known to be normally distributed, the distribution $W$ is a $t$-distribution with $n-1$ degrees of freedom.
-   If the underlying population is known to be symmetric (but not necessarily normal), a $t$-distribution with $n-1$ degrees of freedom is still a good estimate.
-   [**The more skewed the population, the less accurate using a** $t$-distribution becomes]{style="color: blue;"}.

## Question 8

------------------------------------------------------------------------

Using the dataset `NCBirths2004` from the `resampledata` package, give a 99% confidence interval for the mean weight (in grams) of all babies born in North Carolina in 2004.

```{r}
# Load required package
library(resampledata)
```

### Question 8a

------------------------------------------------------------------------

What is a reasonable point estimate?

#### Solution to Question 8a

------------------------------------------------------------------------

```{r}
point.est.nc <- ??  # Find a point estimate
```

### Question 8b

------------------------------------------------------------------------

What value of $t_{\alpha/2}$ should you use?

#### Solution to Question 8b

------------------------------------------------------------------------

```{r}
t.nc <- ??  # Find the value of t_alpha/2
```

### Question 8c

------------------------------------------------------------------------

What is the margin of error of your 99% confidence interval?

#### Solution to Question 8c

------------------------------------------------------------------------

```{r}
moe.nc <- ??  # Compute the margin of error
```

### Question 8d

------------------------------------------------------------------------

Give a 99% confidence interval based on your answers above.

#### Solution to Question 8d

------------------------------------------------------------------------

```{r}
lower.nc <- ??  # lower cutoff for mean weight of newborns
upper.nc <- ??  # lower cutoff for mean weight of newborns
```

### Question 8e

------------------------------------------------------------------------

Construct a 99% confidence interval using the `t.test` function.

#### Solution to Question 8e

------------------------------------------------------------------------

```{r}
# Compute 99% conf int for mean birth weight
t.test(??, conf.level = ??)$cont.int  
```

### Question 8f

------------------------------------------------------------------------

Interpret the practical meaning of your interval estimate.

#### Solution to Question 8f

------------------------------------------------------------------------

\
\
\

## Question 9

------------------------------------------------------------------------

Let $X$ and $Y$ be independent random variables with $X \sim N(\mu_1, \sigma_1$ and $Y \sim N(\mu_2, \sigma_2)$.
Using properties of expected value and variance, show for sample sizes $n_1$ and $n_2$, respectively, that

$$E(\bar{X}-\bar{Y}) = \mu_1-\mu_2 \qquad \mbox{and} \qquad \mbox{Var}(\bar{X}-\bar{Y}) = \frac{\sigma_1^2}{n_1} +  \frac{\sigma_2^2}{n_2}.$$

### Solution to Question 9

------------------------------------------------------------------------

(sample statistic) $\pm$ z val \* SE

For difference of means:

($\bar{X} - \bar{Y}) \pm Z_\frac{\alpha}{2} * (\sqrt{\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}})$ if both standard deviations are known

$\bar{X} - \bar{Y}) \pm t_\frac{\alpha}{2} * (\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}})$ if both standard deviations are unknown\
\
df (degrees of freedom) = $min(n_1 - 1, n_2 - 1)$\

Proof of problem, need to do on your own.

# T Confidence Intervals for a Difference in Means

------------------------------------------------------------------------

If $X_i \sim N(\mu_1, \sigma_1^2)$, $i=1,2,\ldots , n_1$ and $Y_j \sim N(\mu_2, \sigma_2^2)$, $j=1,2,\ldots , n_2$, then an approximate confidence interval for $\mu_1 - \mu_2$ is given by

$$(\bar{X} - \bar{Y}) \pm  t_{\alpha/2} \cdot  \sqrt{\frac{s_1^2}{n_1} +  \frac{s_2^2}{n_2}}$$

where the area under the $t$-distribution with $df$ degrees of freedom between $\pm t_{\alpha/2}$ is equal to the confidence level.

-   Informally, we can use the smaller of $n_1-1$ and $n_2-1$ as the degrees of freedom.

-   A more accurate rule is Welch's approximation: $$ df = \frac{\left( s_1^2/n_1+ s_2^2/n_2 \right)^2}{ \frac{(s_1^2/n_1)^2}{n_1-1} + \frac{(s_2^2/n_2)^2}{n_2-1}}.$$

-   The command `t.test(x, y, conf.level = 0.95)$conf.int` will give a 95% confidence level for the difference in means of samples `x` and `y` using Welch's approximation.

-   The command `t.test(numerical ~ categorical, data = data.name, conf.level = 0.95)$conf.int` will give a 95% confidence level for the difference in means of a specified numerical variable between two different classes of a categorical variable.

## Question 10

------------------------------------------------------------------------

Interpret the practical meaning of the output of the code cell below.

```{r}
t.test(Weight ~ Smoker, data = NCBirths2004, conf.level = 0.90)$conf.int
```

### Solution to Question 10

------------------------------------------------------------------------

The entire interval is positive.
Since 0 is not a plausible value in the interval, this shows that there is a difference between smoker and non smoker mean weights.
One is definitely bigger than the other.

Assuming it is non smoker - smoker, the non smoker mean weights is greater than smoker mean weights.\

\
\

## Question 11

------------------------------------------------------------------------

A study randomly assigned students to take notes either writing them by hand or using a laptop.
The resulting scores of the students on a test of the material are summarized in the table below:

| Group   | n   | $\bar{x}$ | s    |
|---------|-----|-----------|------|
| By hand | 38  | 25.6      | 10.8 |
| Laptop  | 40  | 18.3      | 9.0  |

### Question 11a

------------------------------------------------------------------------

Give a 95% confidence interval using an approximation with a $t$-distribution with degrees of freedom equal to the $n_{\rm min} -1$.

#### Solution to Question 11a

------------------------------------------------------------------------

```{r}
# Construct a 95% confidence interval
# Approx df = the smaller sample size - 1
t.notes <- qt(0.975,37)
pe.notes <- 25.6 - 18.3

moe.notes <- t.notes * sqrt(10.8^2/38 + 9^2/40)

(lower.notes <- pe.notes - moe.notes)
(upper.notes <- pe.notes + moe.notes)
```

point estimate: $\bar{X}_{hand} - \bar{X}_{laptop}$

### Question 11b

------------------------------------------------------------------------

Interpret the meaning of your answer.
Do you believe there is a difference in the exam scores of the two groups?
Explain why or why not.

#### Solution to Question 11b

------------------------------------------------------------------------

\
\
\

### Question 11c

------------------------------------------------------------------------

If you use Welch's approximation, you get $v = 72.1368 \approx 72$.
Give a 95% confidence interval using an approximation with a $t$-distribution with degrees of freedom equal to the $v=72$.

#### Solution to Question 11c

------------------------------------------------------------------------

```{r}
# Construct a 95% confidence interval
# Use Welch's approx df = 72


```
