---
title: '17: Confidence Intervals for Proportions'
author: "Adam Spiegler, University of Colorado Denver"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Confidence Intervals for Proportions

------------------------------------------------------------------------

A recent PBS NewsHour/NPR/Marist poll[^1] surveyed $1,\!082$ randomly selected registered voters in the US to gauge their opinions on how the US is handling the COVID pandemic.

[^1]: "Politics still drives how Americans fell about COVID response, one year in'', PBS, March 11, 2021

![Vaccination Poll Barchart](https://drive.google.com/uc?export=view&id=15UUGHUAPn5flVxg8mD38Ss_VbTW3s3eE)

[**Based on this survey, approximately what proportion of adults in the US do NOT plan to get vaccinated?**]{style="color: blue;"}

![Vaccination Poll Table](https://drive.google.com/uc?export=view&id=15UK92CrkvGjOq7_Y0mK_-AEoNgnqjlnM)

# Central Limit Theorem for Proportions

------------------------------------------------------------------------

Recall if $p$ is the proportion of a population that have a certain characteristic, then the distribution of the sample proportion (when samples size $n$ are randomly selected) will be

$$ \widehat{P} \sim N \left( \color{blue}{\mu_{\widehat{P}}}, \color{red}{\sigma_{\widehat{P}}} \right) = N \left( \color{blue}{p}, \color{red}{\sqrt{ \frac{p(1-p)}{n}}} \ \right) $$

provided both $np \geq 10$ and $n(1-p) \geq 10$.

## Question 1

------------------------------------------------------------------------

Justify each of the steps below to find an interval which will contain the value of $p$ 95% of the time.

$$\begin{aligned}
P \left( -1.96 < Z < 1.96 \right) &= P \left( -1.96 < \frac{\hat{p} - \color{blue}{\mu_{\widehat{P}}}}{\color{red}{\sigma_{\widehat{P}}}} < 1.96 \right) & \mbox{Definition of z-score} \\
&= P \left( -1.96 < \frac{\hat{p} -\color{blue}{p} }{\color{red}{\sqrt{ \frac{p(1-p)}{n}}}} < 1.96 \right) & \mbox{Explanation 1}\\
&=P \left(-1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} < \hat{p} -\color{blue}{p} <  1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} \right) & \mbox{Explanation 2}\\
&= P \left(- \hat{p} -1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} < -\color{blue}{p} <  - \hat{p} + 1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} \right) & \mbox{Explanation 3}\\
&= P \left( \hat{p} -1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} < \color{blue}{p} < \hat{p} + 1.96 \cdot \color{red}{\sqrt{ \frac{p(1-p)}{n}}} \right) & \mbox{Explanation 4}
\end{aligned}$$

Thus we have

$${\large P \left( \hat{p} -1.96 \cdot \sqrt{ \frac{p(1-p)}{n}} < \color{blue}{p} < \hat{p} + 1.96 \cdot \sqrt{ \frac{p(1-p)}{n}} \right) = 0.95}$$

There is a 95% chance the interval $\hat{p} -1.96 \cdot \sqrt{ \frac{p(1-p)}{n}}$ to $\hat{p} + 1.96 \cdot \sqrt{ \frac{p(1-p)}{n}}$ contains the actual value of the population proportion $p$.

### Solution to Question 1

------------------------------------------------------------------------

Explanation 1:

\
\

Explanation 2:

\
\

Explanation 3:

\
\

Explanation 4:

\
\

## Question 2

------------------------------------------------------------------------

In [Question 1] we determined there is a 95% chance the interval $\hat{p} -1.96 \cdot \sqrt{ \frac{p(1-p)}{n}}$ to $\hat{p} + 1.96 \cdot \sqrt{ \frac{p(1-p)}{n}}$ contains the actual value of the population proportion $p$.

-   **Explain why using this interval is problematic to apply in order to construct a confidence interval estimate for the population proportion** $p$.
-   **Give a possible suggestion to help fix the problem you identified**.

### Solution to Question 3

------------------------------------------------------------------------

\
\
\

# Generalizing Results of Confidence Intervals

------------------------------------------------------------------------

Summary of Confidence Intervals

| Parameter(s) of Interest                          | Point Estimate          | Margin of Error                                                                                           |
|---------------------------------------------------|-------------------------|-----------------------------------------------------------------------------------------------------------|
| A single mean ($\sigma^2$ known)                  | $\bar{x}$               | $z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$                                                              |
| A single mean ($\sigma^2$ unknown)                | $\bar{x}$               | $\color{red}{t_{\alpha/2}} \cdot \frac{\color{red}{s}}{\sqrt{n}}$                                         |
| A difference in two means (with unknown variance) | $\bar{x}_1 - \bar{x}_2$ | $\color{red}{t_{\alpha/2}} \cdot \sqrt{ \frac{\color{red}{s_1}^2}{n_1} + \frac{\color{red}{s_2}^2}{n_2}}$ |
| A single proportion                               | $\hat{p}$               | $z_{\alpha/2} \cdot \sqrt{ \frac{\color{red}{\widehat{p}}(1-\color{red}{\widehat{p}})}{n}}$               |

Before adding even more cases to the table above, let's comment on what all of these confidence intervals have in common.

1.  Find a point estimate using a sample statistic.
2.  Compute (estimate) the standard error, SE, of the underlying sampling distribution.
3.  Based on the chosen confidence level, find either $z_{\alpha/2}$ or $t_{\alpha/2}$.
4.  The margin of error, MoE, is $z_{\alpha/2} \cdot \mbox{SE}$ or $t_{\alpha/2} \cdot \mbox{SE}$

$$\mbox{sample statistic} - \mbox{MoE} < \mbox{population parameter} < \mbox{sample statistic} + \mbox{MoE}$$

## Question 3

------------------------------------------------------------------------

A random sample of 1,082 registered voters in the US, 29% of the people polled said they do not intend to get vaccinated for COVID-19.
Construct a 95% confidence interval to estimate the proportion of all adults in the US that do not plan to get vaccinated for COVID-19.

### Solution to Question 3

------------------------------------------------------------------------

```{r}
# Use this code cell to compute confidence interval
phat <- 0.29
n <- 1082
z <- qnorm((0.95 + (0.05/2)),0,1)
se.vaccine <- sqrt((phat * (1-phat))/n)
moe.vaccine <- z * se.vaccine
(lower.vaccine <- phat - moe.vaccine)
(upper.vaccine <- phat + moe.vaccine)
```

\
\
\

# Wald Confidence Interval for a Proportion

------------------------------------------------------------------------

The [**Wald confidence interval for a proportion**]{style="color: blue;"} is given by

$${\large \color{blue}{\hat{p}} - z_{\alpha/2} \cdot \sqrt{ \frac{\color{blue}{\hat{p}}(1-\color{blue}{\hat{p}})}{n}}  < p <  \color{blue}{\hat{p}} + z_{\alpha/2} \cdot \sqrt{ \frac{\color{blue}{\hat{p}}(1-\color{blue}{\hat{p}})}{n}}}$$

We use the [**plug in principle**]{style="color: blue;"} and use $\color{blue}{\hat{p}}$ for the unknown value of $p$ when calculating the standard error.

-   The advantage of this estimate is we can do it by hand.
-   The downside is that when we use $\hat{p}$ in place of $p$ we lose quite a bit of accuracy.

# Agresti-Coull Confidence Interval for a Proportion

------------------------------------------------------------------------

If $X$ denotes the number of successes in a sample of size $n$, let $\color{red}{\tilde{X} = X+2}$, $\color{red}{\tilde{n}=n+4}$, and $\color{red}{\tilde{p} = \dfrac{\tilde{X}}{\tilde{n}}}$.
The [**Agresti-Coull confidence interval for a proportion**]{style="color: blue;"} is

$$\color{red}{\tilde{p}} - z_{\alpha/2} \left( \sqrt{ \frac{\color{red}{\tilde{p}}(1-\color{red}{\tilde{p}})}{\color{red}{\tilde{n}}}} \right) < p <  \color{red}{\tilde{p}} + z_{\alpha/2} \left( \sqrt{ \frac{\color{red}{\tilde{p}}(1-\color{red}{\tilde{p}})}{\color{red}{\tilde{n}}}} \right)$$

## Why Adjust the Sample Proportion?

------------------------------------------------------------------------

### Comparing Bias of the Two Estimators

------------------------------------------------------------------------

-   $\color{blue}{\hat{p} = \dfrac{X}{n}}$ is an **unbiased estimator** of $p$:

$$E( \color{blue}{\hat{p}} ) = E \left( \frac{X}{n} \right) = \frac{1}{n} E(X) = \frac{1}{n} (np) = p$$

-   $\color{red}{\tilde{p} = \dfrac{X+2}{n+4}}$ is an **biased estimator** of $p$:

$$E( \color{red}{\tilde{p}} ) = E \left( \frac{X+2}{n+4} \right) = \frac{1}{n+4} E(X+2) = \frac{1}{n+4} ( np + 2) = \frac{np+2}{n+4} \ne p$$

-   However, as $n \to \infty$, we see that $\color{red}{\tilde{p}}$ gets less and less biased, which is great!

$$\lim_{n \to \infty} E(\color{red}{\tilde{p}}) = \lim_{n \to \infty} \frac{np+2}{n+4} = p$$

### Comparing Efficiency of the Two Estimators

------------------------------------------------------------------------

-   For $\color{blue}{\hat{p}}$, we have

$$\mbox{Var}(\color{blue}{\hat{p}}) =\mbox{Var} \left( \frac{X}{n} \right) = \frac{1}{n^2} \mbox{Var}(X) = \frac{1}{n^2} \big( np(1-p) \big) = \frac{p(1-p)}{n}.$$

-   For $\color{red}{\tilde{p}}$, we have

$$\mbox{Var}(\color{red}{\tilde{p}}) =\mbox{Var} \left( \frac{X+2}{n+4} \right) = \frac{1}{(n+4)^2} \mbox{Var}(X+2) = \frac{1}{(n+4)^2} \bigg( \mbox{Var}(X) + \mbox{Var}(2) \bigg) =  \frac{np(1-p)}{(n+4)^2}.$$

-   Since $\mbox{Var}(\color{red}{\tilde{p}}) < \mbox{Var}(\color{blue}{\hat{p}})$, we see that $\color{red}{\tilde{p}}$ is a more efficient estimator than $\color{blue}{\hat{p}}$.

-   [**This means our interval estimates will have smaller margins of error with Agresti-Coull!**]{style="color: blue;"}

## Question 4

------------------------------------------------------------------------

Using the same poll from [Question 3], find 95% confidence interval for the proportion of all adults in the US that do not plan to get vaccinated using the Agresti-Coull Confidence Interval for a Proportion.

### Solution to Question 4

------------------------------------------------------------------------

```{r}
# Use code to help with the calculations!
ptilde <- (314+2)/(1082+4)
ntilde <- 1082+4
se.tilde <- sqrt((ptilde*(1-ptilde))/ntilde)
moe.with_tilde <- z*se.tilde
(lower.tilde <- ptilde - moe.with_tilde)
(upper.tilde <- ptilde + moe.with_tilde)
```

\
\

# Score Confidence Interval For a Proportion

------------------------------------------------------------------------

Initially in [Question 1] we began constructing a 95% confidence interval for a proportion using the fact that

$$P \left( -1.96 < \frac{\hat{p} -\color{red}{p}}{\sqrt{\frac{\color{red}{p}(1-\color{red}{p})}{n}}} < 1.96 \right) =0.95$$

-   When we substitute an estimate such as $\hat{p}$ or $\tilde{p}$ in place of $p$ in the formula for the standard error, we introduce additional uncertainty and error.
-   Another approach is to algebraically solve the resulting equations for $p$.

$$\dfrac{\hat{p} -\color{red}{p}}{\sqrt{\dfrac{\color{red}{p}(1-\color{red}{p})}{n}}} = 1.96 \qquad \mbox{and} \qquad \dfrac{\hat{p} -\color{red}{p}}{\sqrt{\dfrac{\color{red}{p}(1-\color{red}{p})}{n}}} = -1.96$$

## Finding the Upper Cutoff

------------------------------------------------------------------------

To find the upper cutoff for $p$, we solve

$$\begin{aligned}
 \frac{\hat{p} -\color{red}{p}}{\sqrt{\frac{\color{red}{p}(1-\color{red}{p})}{n}}} &= 1.96\\
 \hat{p} -\color{red}{p} &= 1.96 \left( \sqrt{\frac{\color{red}{p}(1-\color{red}{p})}{n}} \right) \\
(1.96)^2 \left( \frac{\color{red}{p}(1-\color{red}{p})}{n} \right) &= \big( \hat{p} -\color{red}{p} \big)^2\\
\color{red}{p} \left( \frac{1.96^2}{n} \right) - \color{red}{p}^2 \left( \frac{1.96^2}{n} \right) &= \hat{p}^2 - 2 \color{red}{p}\hat{p} + \color{red}{p}^2\\
\left( 1+ \frac{1.96^2}{n} \right) \color{red}{p}^2 + \left(-2\hat{p}-\frac{1.96^2}{n} \right) \color{red}{p} + \hat{p}^2 &= 0
\end{aligned}$$

Solving the resulting the quadratic equation

$$a\color{red}{p}^2 + b \color{red}{p} + c = \left( 1+ \frac{1.96^2}{n} \right) \color{red}{p}^2 + \left(-2\hat{p}-\frac{1.96^2}{n} \right) \color{red}{p} + \hat{p}^2 = 0.$$

Solving $p$ gives an upper limit for a 95% score confidence interval

$${\large \mbox{upper cutoff for p} = \frac{ \hat{p} + \dfrac{1.96^2}{2n} + 1.96 \cdot \sqrt{ \dfrac{\hat{p}(1-\hat{p})}{n} + \dfrac{1.96^2}{4n^2}}}{1+\dfrac{1.96^2}{n}}}$$

## Score Confidence Interval Formulas

------------------------------------------------------------------------

-   If we adjust the confidence interval, then we use $z_{\alpha/2}$ in place of $1.96$ in the formula above.
-   We can repeat the algebraic process above to similarly find the value of the lower cutoff.

The [**score confidence interval for a proportion**]{style="color: blue;"} is given by

$$\begin{aligned}
&L= \dfrac{\hat{p} + \dfrac{z_{\alpha/2}^2}{2n} - z_{\alpha/2} \cdot \sqrt{ \dfrac{\hat{p}(1-\hat{p})}{n} + \dfrac{z_{\alpha/2}^2}{4n^2}}}{1+ \dfrac{z_{\alpha/2}^2}{n}} \\
\\
&U= \dfrac{\hat{p} + \dfrac{z_{\alpha/2}^2}{2n} + z_{\alpha/2} \cdot \sqrt{ \dfrac{\hat{p}(1-\hat{p})}{n} + \dfrac{z_{\alpha/2}^2}{4n^2}}}{1+ \dfrac{z_{\alpha/2}^2}{n}}
\end{aligned}$$

-   In R, use the command `prop.test(X, n, conf.level = CL, correct = FALSE)$conf.int`
    -   $X$ denotes the number of "successes" observed in the sample.
    -   $n$ denotes the total number of observations in the sample.
    -   `CL` is a chosen confidence level (as a proportion).
    -   The option `correct = FALSE` means no continuity correction is applied.

Note: most accurate method to least accurate method:

1)  score confidence interval: finds p upper and p lower algebraically

2)  p tilde: some bias but smaller MoE

3)  wald's method with phat: unbiased but larger MoE

## Question 5

------------------------------------------------------------------------

Find 95% confidence interval for the proportion of all adults in the US that do not plan to get vaccinated by completing the `prop.test()` command in the first code cell.

### Solution to Question 5

------------------------------------------------------------------------

```{r}
prop.test(0.29*1082, 1082, conf.level = 0.95, correct = FALSE)$conf.int
```

-   **CHECK YOUR WORK BY COMPLETTING AND RUNNING THE CODE CELL BELOW.**

```{r}
##################################################
# Replace the ?? in the two lines of code below
# with appropriate values
##################################################
X <- 0.29*1082  # number of successes in sample (do not plan to get vax)
n <- 1082  # sample size
  
phat <- X/n  # Compute sample proportion

###################################################
# Replace the ?? below with a command using qnorm
###################################################
z.star <- qnorm(0.975,0,1)  # find z_alpha/2

# Computes Cutoffs for Score Confidence Interval
lower <- (phat+z.star^2/(2*n) - 
            z.star*sqrt( (phat*(1-phat))/n + z.star^2/(4*n^2) ) )/(1+z.star^2/n)
upper <- (phat+z.star^2/(2*n) + 
            z.star*sqrt( (phat*(1-phat))/n + z.star^2/(4*n^2) ) )/(1+z.star^2/n)

# Print cutoffs to screen
lower
upper
```

## Applying the Continuity Correction

------------------------------------------------------------------------

Recall, when using a normal distribution to approximate the sampling distribution for a discrete random variable $X$, we can improve the estimate by using a continuity correction.
In the case of a confidence interval for a proportion, the continuity correction as applied as follows:

-   In the formula for the lower cutoff, we use the corrected sample proportion $\color{blue}{\hat{p}_L^*} = \dfrac{X-0.5}{n}$

-   In the formula for the upper cutoff, we use the corrected sample proportion $\color{red}{\hat{p}_U^*} = \dfrac{X+0.5}{n}$.

$$\begin{aligned}
&L^* = \dfrac{\color{blue}{\hat{p}_L^*} + \dfrac{z_{\alpha/2}^2}{2n} - z_{\alpha/2} \cdot \sqrt{ \dfrac{\color{blue}{\hat{p}_L^*}(1-\color{blue}{\hat{p}_L^*})}{n} + \dfrac{z_{\alpha/2}^2}{4n^2}}}{1+ \dfrac{z_{\alpha/2}^2}{n}} \\
\\
&U^* = \dfrac{\color{red}{\hat{p}_U^*} + \dfrac{z_{\alpha/2}^2}{2n} + z_{\alpha/2} \cdot \sqrt{ \dfrac{\color{red}{\hat{p}_U^*}(1-\color{red}{\hat{p}_U^*})}{n} + \dfrac{z_{\alpha/2}^2}{4n^2}}}{1+ \dfrac{z_{\alpha/2}^2}{n}}
\end{aligned}$$

Apply continuity correction because we are dealing with discrete variables using a normal distribution that is used to handle continuous values.
When applying normal distribution, the area under the curve will miss some of the area covered in the discrete graph.

### Applying the Continuity Correction in Code

------------------------------------------------------------------------

-   Below we perform the direct calculations using the formulas we derived.

```{r}
############################################
# Be sure you have run previous code cells
# And have already defined X, n, and z.star
############################################

# Continuity corrections applied to sample proportion
cc.phat.L <- (X - 0.5)/n
cc.phat.U <- (X + 0.5)/n

# Plugged into formulas for Score Conf Interval
cc.lower <- (cc.phat.L + z.star^2/(2*n) - 
            z.star*sqrt( (cc.phat.L*(1-cc.phat.L))/n + z.star^2/(4*n^2) ) )/(1+z.star^2/n)
cc.upper <- (cc.phat.U + z.star^2/(2*n) + 
            z.star*sqrt( (cc.phat.U*(1-cc.phat.U))/n + z.star^2/(4*n^2) ) )/(1+z.star^2/n)

# Print results to screen to check
cc.lower
cc.upper
```

-   In the code cell below, we apply the continuity correction using the `correct = TRUE` option in `prop.test()` to check the result.

```{r}
prop.test(314, 1082, conf.level = 0.95, correct = TRUE)$conf.int
```

\
\
\

# Confidence Intervals for a Difference in Two Proportions

------------------------------------------------------------------------

We can modify the [**Wald confidence interval**]{style="color: blue;"} to give an approximation for a confidence interval for a difference in two proportions

-   The [**point estimate**]{style="color: blue;"} where the interval will be centered is the difference in the two sample proportions, $\color{blue}{\hat{p}_1 - \hat{p}_2}$.
-   The [**standard error**]{style="color: blue;"} we approximate by plugging $\hat{p}_1$ and $\hat{p}_2$ in place of $p_1$ and $p_2$ in the formula for the standard error from the Central Limit Theorem

$$\mbox{SE} \left( \widehat{P}_1 - \widehat{P}_2 \right) = \sqrt{ \frac{p_1(1-p_1)}{n_1} + \frac{p_2(1-p_2)}{n_2}} \approx \sqrt{ \frac{\color{blue}{\hat{p}}_1(1-\color{blue}{\hat{p}}_1)}{n_1} + \frac{\color{blue}{\hat{p}}_2(1-\color{blue}{\hat{p}}_2)}{n_2}}$$

-   We use the standard normal distribution to identify $z_{\alpha/2}$ to find the [**margin of error**]{style="color: blue;"}.

$$(\hat{p}_1 - \hat{p}_2) - z_{\alpha/2} \cdot \sqrt{ \dfrac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \dfrac{\hat{p}_2 (1-\hat{p}_2) }{n_2}}  < p_1-p_2 < (\hat{p}_1 - \hat{p}_2) + z_{\alpha/2} \cdot \sqrt{ \dfrac{\hat{p}_1 (1-\hat{p}_1) }{n_1} + \frac{\hat{p}_2 (1-\hat{p}_2) }{n_2}}.$$

-   There are other variations pf confidence intervals for a difference in two proportions, similar to the Agresti-Coull and score confidence intervals we discussed for a single proportion.

-   [**For a difference in two proportions, we will use the Wald confidence interval.**]{style="color: blue;"}

-   The command `prop.test(c(x1, x2), c(n1, n2), conf.level = CL, correction = FALSE)$conf.int` will compute a **Wald confidence interval for a difference in two proportions**.

## Question 6

------------------------------------------------------------------------

Using the data below collected from a survey, construct a 90% interval for the difference in the proportion of all Democrats and proportion of all Republicans that do not plan to be vaccinated.

| Party      | Yes, will | Yes, already | No  | Unsure | Total |
|------------|-----------|--------------|-----|--------|-------|
| Democrat   | 213       | 108          | 40  | 7      | 368   |
| Republican | 93        | 70           | 120 | 9      | 292   |
| Total      | 306       | 178          | 160 | 16     | 660   |

### Solution to Question 6

------------------------------------------------------------------------

```{r}
#using wald confidence interval
#difference in proprotion use P(democrat) - P(republican)

phat.dem.no_vac <- 40/368
phat.rep.no_vac <- 120/292
z.vac <- qnorm((0.9 + (0.1/2)),0,1)
n1 <- 368
n2 <- 292

se.p_dem <- (phat.dem.no_vac * (1-phat.dem.no_vac)) / n1
se.p_rep <- (phat.rep.no_vac * (1-phat.rep.no_vac)) / n2
se.no_vac <- sqrt(se.p_dem + se.p_rep)

moe.no_vac <- z.vac * se.no_vac

diff.no_vac <- phat.dem.no_vac - phat.rep.no_vac

(lower.no_vac <- diff.no_vac - moe.no_vac)
(upper.no_vac <- diff.no_vac + moe.no_vac)
```

```{r}
# use code cell to help
#enter as a vector of successes and a vector of num trials using prop.test function
no_vac <- c(40, 120)
total <- c(368, 292)

prop.test(no_vac, total, conf.level=0.9, correct=FALSE)$conf.int
```

```{r}
#with continuity correction (try on your own)
phat.dem.no_vac <- 40/368
phat.rep.no_vac <- 120/292
z.vac <- qnorm((0.9 + (0.1/2)),0,1)
n1 <- 368
n2 <- 292

se.p_dem <- (phat.dem.no_vac * (1-phat.dem.no_vac)) / n1
se.p_rep <- (phat.rep.no_vac * (1-phat.rep.no_vac)) / n2
se.no_vac <- sqrt(se.p_dem + se.p_rep)

moe.no_vac <- z.vac * se.no_vac

diff.no_vac <- phat.dem.no_vac - phat.rep.no_vac

(lower.no_vac <- diff.no_vac - moe.no_vac)
(upper.no_vac <- diff.no_vac + moe.no_vac)
```

```{r}
?prop.test
```

Confidence interval for matched pair differences

mean(X1 - X2) +/- t \* sqrt((sdiff)\*\*2/n + (sdiff)\^2/n)
